# <em>R App Logging</em> {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

This lab covers how to use logging to monitor events in the application and the API. For a refresher, I've included an updated diagram from the previous lab of Shiny App and API architecture (i.e., the reactives (user inputs and the **Predict** button), the `httr2` API request/response, and the `plumber` API).:

```{=html}
<style>

.codeStyle span:not(.nodeLabel) {
  font-family: monospace;
  font-size: 1.5em;
  font-weight: bold;
  color: #d80808 !important;
  background-color: #ffffff;
  padding: 0.2em;
}
</style>
```

```{mermaid}
%%| fig-cap: 'Shiny R App and plumber API'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace'}}}%%

graph TD
    subgraph Inputs["<strong>User Inputs</strong>"]
    I1[/"input$bill_length"/]
    I2[/"input$species"/]
    I3[/"input$sex"/]
    I4[/"input$predict"/]
    end

    subgraph API["<strong>plumber API</strong>"]
      ModBoard("<code>pins::board_folder()</code>")
      ModPin("<code>vetiver::vetiver_pin_read()</code>")
      ModRout("<code>plumber::pr()</code>") 
      ModApi("<code>vetiver::vetiver_api()</code>") 
      ApiServ(["Launch<br>API Server"])
      ModBoard --"<em>connect to model board</em>"--> ModPin 
      ModPin --"<em>read pinned vetiver model</em>"--> ModRout 
      ModRout --"<em>create router</em>"--> ModApi 
      ModApi --"<em>start API server</em>"--> ApiServ
    end

    subgraph Pred["<strong>Predict</strong>"]
        Req("<code>httr2::request()</code>") --"<em>Creates HTTP<br>request object</em>"--> Meth("<code>httr2::req_method()</code>")
        Meth --"<em>Changes HTTP<br>method from<br>GET to POST</em>"--> BodyJson1("<code>httr2::req_body_json()</code>")
        BodyJson1 --"<em>Converts data.frame<br>to JSON & sets<br>request body</em>"--> Perform("<code>httr2::req_perform()</code>")
        BodyJson2("<code>httr2::resp_body_json()</code>")
    end

    Perform --"<em>Sends HTTP<br>request to<br>server &<br>waits for<br>response</em>"--> API 
    API --> BodyJson2


    subgraph PredReact["<strong>pred()</strong>"]
        Res(["<strong>response$.pred[1]</strong>"])
    end
    
    ValsReact(["<strong>vals()</strong><br>(in <code>data.frame</code>)"])
    BodyJson2 --"<em>Extracts JSON<br>response body<br>& converts to<br>R object</em>"--> PredReact
    
    I1 --> ValsReact
    I2 --> ValsReact
    I3 --> ValsReact
    I4 --"<em><code>reactive()</code> & <br><code>bindEvent()</code></em>"---> Pred
    
    ValsReact --"<em>as<br><code>request_data</code></em>"--> BodyJson1
    ValsReact --> RenPrnt[\"<strong>renderPrint()</strong><br><code>vals()</code>"\]
    PredReact -.-> RenTxt[\"<strong>renderText()</strong><br><code>pred()</code>"\]

    %% Style for functions %%
    style Req fill:#FFF,color:#000,rx:20,ry:20,font-size:18px
    style Meth fill:#FFF,color:#000,rx:20,ry:20,font-size:18px
    style BodyJson1 fill:#FFF,color:#000,rx:20,ry:20,font-size:18px
    style Perform fill:#FFF,color:#000,rx:20,ry:20,font-size:18px
    style BodyJson2 fill:#FFF,color:#000,rx:20,ry:20,font-size:18px
    
    %% Style for reactives %%
    style ValsReact fill:#e1f5fe,stroke:#000,color:#000,stroke-width:1px,rx:10,ry:10,font-size:16px
    style PredReact fill:#fff3e0,stroke:#000,color:#000,stroke-width:1px,rx:10,ry:10,font-size:16px
    
    %% Style for result %%
    style ApiServ fill:#e8f5e8,stroke:#000,color:#000
    style Res fill:#e8f5e8,stroke:#000,color:#000
```


## Configure 

Below we load the packages for app design (`bslib`, `bsicons`), making the API request (`httr2`), and logging and monitoring (`logger` and `jsonlite`):

```{r}
#| eval: false
#| code-fold: false
library(shiny)
library(bslib)
library(bsicons)
library(httr2)
library(logger)
library(jsonlite)

# configure logging 
logger::log_threshold(level = "DEBUG")
logger::log_appender(appender = appender_tee(file = "shiny_app.log"))
logger::log_formatter(logger::formatter_glue_or_sprintf)
```

The 


## UI 


```{r}
#| label: header
#| eval: false
#| code-fold: false
#| code-overflow: wrap
  div(
    style = "position: fixed; top: 10px; right: 10px; z-index: 1000;",
    textOutput("log_status", inline = TRUE)
  )
```

```{r}
#| label: main-content
#| eval: false
#| code-fold: false
#| code-overflow: wrap
  layout_columns(
    card(
      card_header("Penguin Parameters"),
      card_body(
        verbatimTextOutput(outputId = "vals")
      )
    ),
    card(
      card_header("Predicted Mass"),
      card_body(
        value_box(
          showcase_layout = "left center",
          title = "Grams",
          value = textOutput(outputId = "pred"),
          showcase = bs_icon("graph-up"),
          max_height = "200px",
          min_height = "200px",
        )
      )
    ),
    col_widths = c(6, 6),
    row_heights = c(3, 1),
  )
```

```{r}
#| label: sys-status
#| eval: false
#| code-fold: false
#| code-overflow: wrap
  layout_columns(
    card(
      card_header("System Status"),
      card_body(
        h5("API Status:"),
        textOutput("api_health"),
        h5("Recent Logs:"),
        div(
          style = "font-family: 'Courier New', monospace; font-size: 12px; background-color: #f8f9fa; padding: 10px; border-radius: 5px;",
          verbatimTextOutput("recent_logs", placeholder = TRUE)
        ),
        h6(
          "Last updated:", 
            textOutput("log_timestamp", inline = TRUE)
          ),
        style = "max-height: 350px; overflow-y: auto;"
      )
    ),
    col_widths = c(12), # full width 
    height = "300px" 
  )
```


```{r}
#| eval: false
#| code-fold: false
#| include: false 
# session tracking ----
api_url <- "http://127.0.0.1:8080/predict"

ui <- page_sidebar(
  title = "Penguin Mass Predictor",
  theme = bs_theme(bootswatch = "sketchy"),
  
  # logging status indicator ----
  div(
    style = "position: fixed; top: 10px; right: 10px; z-index: 1000;",
    textOutput("log_status", inline = TRUE)
  ),
  
  sidebar = sidebar(
    sliderInput(inputId = "bill_length", 
      label = "Bill Length (mm)",
      min = 30, 
      max = 60, 
      value = 45, 
      step = 1),
    selectInput(inputId = "sex", 
      label = "Sex", 
      choices = c("Male", "Female"), 
      selected = "Male"),
    selectInput(
      inputId = "species",
      label = "Species",
      choices = c("Adelie", "Chinstrap", "Gentoo"),
      selected = "Adelie"),
    actionButton(
      inputId = "predict", 
      label = "Predict", 
      class = "btn-primary")
  ),
  
  # main content -----
  layout_columns(
    card(
      card_header("Penguin Parameters"),
      card_body(
        verbatimTextOutput(outputId = "vals")
      )
    ),
    card(
      card_header("Predicted Mass"),
      card_body(
        value_box(
          showcase_layout = "left center",
          title = "Grams",
          value = textOutput(outputId = "pred"),
          showcase = bs_icon("graph-up"),
          max_height = "200px",
          min_height = "200px",
        )
      )
    ),
    col_widths = c(6, 6),
    row_heights = c(3, 1),
  ),
  
  # sys status ----
  layout_columns(
    card(
      card_header("System Status"),
      card_body(
        h5("API Status:"),
        textOutput("api_health"),
        h5("Recent Logs:"),
        div(
          style = "font-family: 'Courier New', monospace; font-size: 12px; background-color: #f8f9fa; padding: 10px; border-radius: 5px;",
          verbatimTextOutput("recent_logs", placeholder = TRUE)
        ),
        h6(
          "Last updated:", 
            textOutput("log_timestamp", inline = TRUE)
          ),
        style = "max-height: 350px; overflow-y: auto;"
      )
    ),
    col_widths = c(12), # full width 
    height = "300px" 
  )
)
```


## Server 

```{mermaid}
%%| fig-cap: 'Logging in Server'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"18px"}}}%%

graph TD
    A[/"Log File: <br><code>shiny_app.log</code>"/] -->|"<em>file changes</em>"| B["Read with: <br><code>reactiveFileReader()</code>"]
    B -->|"<em>polls every 1000ms</em>"| C["Check file modification time"]
    C -->|"<em>if changed</em>"| D["Read entire file"]
    D --> E["Extract last<br>5 lines"]
    E --> F["Update UI display"]
    
    G(["User Actions"]) -->|"<em>generate logs</em>"| H["Log using: <br><code>logger::log_info/warn/error()</code>"]
    H -->|"<em>writes to</em>"| A
    
    I(["API Calls"]) -->|"<em>generate logs</em>"| H
    J(["Input Validation"]) -->|"<em>generate logs</em>"| H
    K(["Session Events"]) -->|"<em>generate logs</em>"| H
    
    B --> L["Store metadata"]
    L --> M["File modification time"]
    L --> N["Total line count"]
    L --> O["Lines content"]
    
    F --> P(["<code>recent_logs</code> <strong>output</strong>"])
    M --> Q(["<code>log_timestamp</code> <strong>output</strong>"])
    
    style G fill:#FFF,color:#000,rx:20,ry:20,font-size:18px
    style I fill:#FFF,color:#000,rx:20,ry:20,font-size:18px
    style J fill:#FFF,color:#000,rx:20,ry:20,font-size:18px
    style K fill:#FFF,color:#000,rx:20,ry:20,font-size:18px
    
    style P fill:#e8f5e8,stroke:#000,color:#000,font-size:20px
    style Q fill:#e8f5e8,stroke:#000,color:#000,font-size:20px
    
```

### App startup

Log app startup within observer to access session data (high priority to run early)

```{r}
#| eval: false 
#| code-fold: false
#| code-overflow: wrap
  observe({
    logger::log_info("Shiny app started - Session: {session$token} - Host: {session$clientData$url_hostname}")
  }, priority = 1000)  
```

### Log file reading

Reactive file reader for log monitoring, polls the log file every 1000ms (1 second) for changes:    

*  Read all lines and get file modification time     
*  Return both content and timestamp   

```{r}
#| eval: false 
#| code-fold: false
log_file_content <- reactiveFileReader(
  intervalMillis = 1000,
  session = session,
  filePath = "shiny_app.log",
  readFunc = function(filePath) {
    
    if (file.exists(filePath)) {
      
      lines <- readLines(filePath, warn = FALSE)
      mod_time <- file.mtime(filePath)
      list(
        lines = lines,
        last_modified = mod_time,
        total_lines = length(lines)
      )
      
    } else {
      
      list(
        lines = character(0),
        last_modified = Sys.time(),
        total_lines = 0
      )
      
    }
  }
)
```

### User interactions

Track user interactions (`throttle` to avoid excessive logging).

```{r}
#| eval: false 
#| code-fold: false
#| code-overflow: wrap
observe({
  logger::log_debug("User input changed - Session: {session$token} - bill_length: {input$bill_length} - species: {input$species} - sex: {input$sex}")
}) |> 
  throttle(2000)  
```

### Reactive values 

Reactive values (with validation)

```{r}
#| eval: false 
#| code-fold: false
#| code-overflow: wrap
vals <- reactive({
  bill_length <- input$bill_length
  species <- input$species
  sex <- input$sex
  
  # input validation (with warnings)
  if (bill_length < 30 || bill_length > 60) {
    logger::log_warn("Bill length out of typical range - Session: {session$token} - bill_length: {bill_length}")
  }
  
  if (is.null(species) || is.null(sex)) {
    logger::log_error("Missing required inputs - Session: {session$token} - species_null: {is.null(species)} - sex_null: {is.null(sex)}")
    return(NULL)
  }
  
  data <- data.frame(
    bill_length_mm = bill_length,
    species_Chinstrap = as.numeric(species == "Chinstrap"),
    species_Gentoo = as.numeric(species == "Gentoo"),
    sex_male = as.numeric(sex == "Male")
  )
  
  logger::log_debug("Input data prepared - Session: {session$token} - data: {jsonlite::toJSON(data, auto_unbox = TRUE)}")
  
  return(data)
})
```


### API health check

```{r}
#| eval: false 
#| code-fold: false
#| code-overflow: wrap
api_health <- reactive({
  tryCatch({
    logger::log_debug("Checking API health - Session: {session$token}")
    
    response <- httr2::request("http://127.0.0.1:8080/ping") |>
      httr2::req_timeout(5) |>
      httr2::req_perform()
    
    if (httr2::resp_status(response) == 200) {
      logger::log_info("API health check successful - Session: {session$token}")
      return("✅ API Online")
    } else {
      logger::log_warn("API health check returned non-200 status - Session: {session$token} - status: {httr2::resp_status(response)}")
      return("⚠️ API Issues")
    }
  }, error = function(e) {
    logger::log_error("API health check failed - Session: {session$token} - error: {conditionMessage(e)}")
    return("❌ API Offline")
  })
})
```


### Predictions 

Enhanced prediction with logging

```{r}
#| eval: false 
#| code-fold: false
#| code-overflow: wrap
  pred <- reactive({
    request_start <- Sys.time()
    request_data <- vals()
    
    if (is.null(request_data)) {
      logger::log_error("Cannot make prediction with invalid inputs - Session: {session$token}")
      return("❌ Invalid inputs")
    }
    
    logger::log_info("Starting prediction request - Session: {session$token} - request_data: {jsonlite::toJSON(request_data, auto_unbox = TRUE)}")
    
    tryCatch({
      showNotification("Predicting penguin mass...", 
                       type = "default", duration = 3)
      
      response <- httr2::request(api_url) |>
        httr2::req_method("POST") |>
        httr2::req_body_json(request_data, auto_unbox = FALSE) |>
        httr2::req_timeout(30) |>
        httr2::req_perform()
      
      response_time <- as.numeric(difftime(Sys.time(), request_start, units = "secs"))
      response_data <- httr2::resp_body_json(response)
      
      logger::log_info("Prediction successful - Session: {session$token} - response_time_sec: {round(response_time, 3)} - prediction: {response_data$.pred[1]}")
      
      # performance monitoring
      if (response_time > 5) {
        logger::log_warn("Slow API response - Session: {session$token} - response_time_sec: {response_time}")
      }
      
      showNotification("✅ Prediction successful!", 
                       type = "message", duration = 3)
      
      return(response_data$.pred[1])
      
    }, error = function(e) {
      error_msg <- conditionMessage(e)
      response_time <- as.numeric(difftime(Sys.time(), request_start, units = "secs"))
      
      logger::log_error("Prediction request failed - Session: {session$token} - error: {error_msg} - response_time_sec: {round(response_time, 3)}")
      
      # classify error types for better monitoring
      if (grepl("Connection refused|couldn't connect", error_msg, ignore.case = TRUE)) {
        user_msg <- "API not available - is the server running on port 8080?"
        logger::log_error("API connection refused - Session: {session$token}")
      } else if (grepl("timeout|timed out", error_msg, ignore.case = TRUE)) {
        user_msg <- "Request timed out - API may be overloaded"
        logger::log_warn("API timeout occurred - Session: {session$token}")
      } else {
        user_msg <- paste("API Error:", substr(error_msg, 1, 50))
        logger::log_error("Unknown API error - Session: {session$token} - error: {error_msg}")
      }
      
      showNotification(paste("❌", user_msg), type = "error", duration = 5)
      return(paste("❌", user_msg))
    })
  }) |> 
    bindEvent(input$predict, ignoreInit = TRUE)
```




```{r}
#| eval: false 
#| code-fold: false
#| include: false
server <- function(input, output, session) {

  # outputs with logging
  output$pred <- renderText({
    prediction <- pred()
    
    if (is.numeric(prediction)) {
      result <- paste(round(prediction, 1), "grams")
      logger::log_info("Displaying prediction to user - Session: {session$token} - display_value: {result}")
      return(result)
    } else {
      logger::log_debug("Displaying error message to user - Session: {session$token} - message: {prediction}")
      return(prediction)
    }
  })
  
  output$vals <- renderPrint({
    data <- vals()
    if (!is.null(data)) {
      logger::log_debug("Displaying input values to user - Session: {session$token}")
      return(data)
    } else {
      return("Invalid inputs")
    }
  })
  
  output$api_health <- renderText({
    api_health()
  })
  
  output$log_status <- renderText({
    paste("Session:", substr(session$token, 1, 8))
  })
  
  # reactive log display that updates when file changes ----
  output$recent_logs <- renderText({
    log_data <- log_file_content()
    
    if (length(log_data$lines) > 0) {
      # get last 5 lines
      recent_lines <- if (log_data$total_lines > 5) {
        tail(log_data$lines, 5)
      } else {
        log_data$lines
      }
      
      # log that we're updating the display (at DEBUG level to avoid spam)
      logger::log_debug("Updating recent logs display - Session: {session$token} - showing {length(recent_lines)} lines")
      
      paste(recent_lines, collapse = "\n")
    } else {
      "No logs available"
    }
  })
  
  # timestamp of last log update ----
  output$log_timestamp <- renderText({
    log_data <- log_file_content()
    format(log_data$last_modified, "%H:%M:%S")
  })
  
  # log session end
  session$onSessionEnded(function() {
    logger::log_info("User session ended - Session: {session$token}")
  })
}
```

```{r}
#| eval: false 
#| code-fold: false

# log app startup at global level
logger::log_info("Shiny application initialized - timestamp: {Sys.time()} - r_version: {R.version.string}")

shinyApp(ui = ui, server = server)
```

