# More `httr2` functions

```{r}
#| eval: true 
#| echo: false 
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_dev
#| echo: false
#| results: asis
#| eval: false
co_box(
  color = "r", 
  look = "minimal",
  header = "Alert",
  contents = "The contents for section are being revised. Thank you for your patience."
)
```

Below are more `httr2` functions for communicating with APIs.  

### HTTP headers

`httr2::req_headers()` adds HTTP headers that provide metadata about the request:

-   **`Content-Type`**: Tells server what format we're sending

-   **`Authorization`**: Provides authentication credentials

-   **`Accept`**: Tells server what response format we prefer

-   **`User-Agent`**: Identifies our client application

```{mermaid}
%%| fig-width: 6.5
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace'}}}%%

graph TD
	A(["Request Object"]) --> B("<strong>req_headers()</strong>")
	B --> C("Adds metadata<br>to request")
	C --> D("Authentication,<br>Content-Type, etc.")

style A fill:#d8e4ff

```

```{r}
#| eval: false
#| code-fold: false
req |> req_headers(
    "Content-Type" = "application/json", # what format is the data?
    "Authorization" = "Bearer abc123", # auth token
    "User-Agent" = "R-httr2-client/1.0", # client identification
    "Accept" = "application/json" # what format to return?
)
```


### Query parameters

`httr2::req_url_query()` adds query parameters to the URL (the part after `?`):

-   **`GET` requests**: Primary way to send parameters

-   **Filtering data**: `?category=electronics&price_max=100`

-   **Pagination**: `?page=2&per_page=50`

-   **API options**: `?format=json&verbose=true`

```{mermaid}
%%| fig-width: 6.5
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace'}}}%%

graph TD
	A(["Base URL"]) --> B("<strong>req_url_query()</strong>")
	B --> C("Adds<br><strong>?param=value&key=data</strong>")
	C --> D("<strong>https://</strong><strong>api.com/data?species=Adelie&limit=10</strong>")

style A fill:#d8e4ff
```

```{r}
#| eval: false
#| code-fold: false
req |> req_url_query(
    species = "Adelie", # filter by species
    limit = 10, # limit results
    include_na = "false" # drop missing values
)
# result: https://api.com/penguins?species=Adelie&limit=10&include_na=false
```

### Status codes

`httr2::resp_status()` extracts the HTTP status code to determine if request succeeded.

- **200**: OK (success)

- **201**: Created (successful POST)

- **400**: Bad Request (client error)

- **401**: Unauthorized (authentication failed)

- **404**: Not Found (resource doesn't exist)

- **500**: Internal Server Error (server problem)

```{mermaid}
%%| fig-width: 6.5
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace'}}}%%

graph TD
	A(["Response Object"]) --> B("<strong>resp_status()</strong>")
	B --> C("Extracts status<br>code")
	C --> D("200, 404, 500, etc.")

  style A fill:#d8e4ff

```

```{r}
#| eval: false
#| code-fold: false
status <- resp_status(response)
if (status == 200) {
    print("Success!")
} else if (status == 404) {
    print("Not found")
} else if (status >= 500) {
    print("Server error")
}
```

### Response headers

`httr2::resp_headers()` extracts HTTP headers from the response for metadata. Useful response headers include:

- **Content-Type**: Format of response body

- **Content-Length**: Size of response

- **Cache-Control**: Caching instructions

- **Set-Cookie**: Cookies to store

```{mermaid}
%%| fig-width: 6.5
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace'}}}%%

graph TD
  A(["Response Object"]) --> B("<strong>resp_headers()</strong>")
  B --> C("Extracts response<br>metadata")
  C --> D("Content-Type,<br>Content-Length, etc.")

  style A fill:#d8e4ff

```

```{r}
#| eval: false
#| code-fold: false
headers <- resp_headers(response)
content_type <- headers$`content-type` # "application/json"
content_length <- headers$`content-length` # "1234"
server_info <- headers$server # "nginx/1.18.0"
```


