# <em>Install <code>splam</code></em> {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default",
  hsize = "1.25",
  size = "1.00",
  header = "Caution",
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

This isn't a chapter covered in DO4DS, but I decided to add it because it's a 'real world' example of deploying a Python application on AWS. The application is called [`splam`](https://mjfrigaard.github.io/splam/), or 'Shiny for Python Linux Admin Management.' You can view it's source code here: <https://github.com/mjfrigaard/splam>

This chapter will walk through how to deploy the `splam` application in an AWS EC2 instance. It will cover the topics we've touched on in previous chapters (and include other things to consider).

## What the app needs from the host

`splam` is a Shiny for Python app that shells out to `systemctl` and `journalctl`. Every button in the UI ends up as a subprocess call against the service manager on the machine the app is running on.

That constraint decides the whole deployment. The app can't run in a container, because the `systemd` it would see is the container's, not the host's. It can't run behind Lambda, because there's no host to manage. It has to run directly on the box it administers, which on AWS means an EC2 instance, daemonized with `systemd`, with something in front of it terminating TLS.

Here's the shape of what we're building:

```{mermaid}
%%| fig-cap: 'splam Deployment on EC2'
%%| fig-width: 8.0
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace', "fontSize":"12px"}}}}%%

graph TD
    Op(["Operator<br/>browser"])
    Admin(["You<br/>SSH client"])

    subgraph EC2["EC2 Instance"]
        Nginx["<strong>nginx</strong><br/>:443 TLS"]
        Splam["<strong>splam</strong><br/>127.0.0.1:8000"]
        Sysd[("<strong>systemd</strong> + <strong>journald</strong>")]
    end

    Op --"HTTPS :443"--> Nginx
    Admin --"SSH :22"--> EC2
    Nginx --"HTTP :8000"--> Splam
    Splam --"<em>systemctl / journalctl</em>"--> Sysd
    Sysd --"<em>supervises</em>"--> Splam

    style EC2 fill:#1B2A41,stroke:#fff,stroke-width:2px,color:#fff
    style Op fill:#d2562b,stroke:#fff,stroke-width:2px,color:#fff
    style Admin fill:#d2562b,stroke:#fff,stroke-width:2px,color:#fff
    style Nginx fill:#2a6f77,stroke:#fff,stroke-width:1px,color:#fff
    style Splam fill:#5b8c5a,stroke:#fff,stroke-width:2px,color:#fff
    style Sysd fill:#e8a33d,stroke:#000,stroke-width:2px,color:#000
```

Notice the loop at the bottom. `systemd` supervises `splam`, and `splam` controls `systemd`. That's fine, right up until you use the app to restart the `splam` unit itself and wonder why your browser session died.

```{r}
#| label: co_box_sg
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default",
  hsize = "1.15",
  size = "1.00",
  header = "Security group rules",
  fold = TRUE,
  contents = "
The instance security group we set up back in [AWS EC2](aws-ec2.qmd) needs exactly two inbound rules for this deployment:

- `22/tcp` from your own IP or VPN range, never `0.0.0.0/0`
- `443/tcp` from wherever operators connect

Do **not** open `8000/tcp`. The app binds to `127.0.0.1`, so it isn't reachable from outside the instance anyway, but leaving the port closed means a misconfigured bind address can't quietly expose an admin console to the internet.
  "
)
```

## Getting a modern Python onto the instance

`splam` needs Python 3.11 or newer. Amazon Linux 2 ships with 3.7:

```{bash}
#| eval: false
#| code-fold: false
python3 --version
```

```{verbatim}
Python 3.7.16
```

That's the copy `yum` depends on, so upgrading it in place is off the table for the reasons covered in [Install Apps](install-apps.qmd). The fix is the same `/opt` pattern from that chapter: build a second Python, install it into its own versioned directory, and leave the system one alone.

```{mermaid}
%%| fig-cap: 'Which Python Runs the App'
%%| fig-width: 8.0
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace', "fontSize":"12px"}}}}%%

graph TD
    Check{"<strong>python3 --version</strong><br/>3.11 or newer?"}
    Yes("Use it directly")
    Build("Build into<br/><strong>/opt/python/3.12.1</strong>")
    Venv("<strong>/opt/splam/.venv</strong>")
    Sys(["<strong>/usr/bin/python3</strong><br/><em>3.7, owned by yum</em>"])

    Check --"yes (AL2023, Ubuntu 22.04+)"--> Yes
    Check --"no (Amazon Linux 2)"--> Build
    Yes --> Venv
    Build --"<em>python3 -m venv</em>"--> Venv
    Sys --"<em>leave for</em>"--> Yum("yum, OS tooling")

    style Check fill:#e8a33d,stroke:#000,stroke-width:2px,color:#000
    style Yes fill:#5b8c5a,stroke:#fff,stroke-width:1px,color:#fff
    style Build fill:#2a6f77,stroke:#fff,stroke-width:1px,color:#fff
    style Venv fill:#5b8c5a,stroke:#fff,stroke-width:2px,color:#fff,font-family:monospace
    style Sys fill:#d2562b,stroke:#fff,stroke-width:2px,color:#fff
    style Yum fill:#666,stroke:#999,stroke-width:1px,color:#fff
```

Install the compiler and the development headers first. These are the same libraries from [Install Apps](install-apps.qmd), plus a few Python needs specifically for `ssl`, `bz2`, and `ctypes` support:

```{bash}
#| eval: false
#| code-fold: false
sudo yum groupinstall -y "Development Tools"
```


```{bash}
#| eval: false
#| code-fold: false
sudo yum install -y openssl-devel bzip2-devel libffi-devel zlib-devel git
```

Download and unpack the source:

```{bash}
#| eval: false
#| code-fold: false
cd /tmp
```

```{bash}
#| eval: false
#| code-fold: false
wget https://www.python.org/ftp/python/3.12.1/Python-3.12.1.tgz
```

```{bash}
#| eval: false
#| code-fold: false
tar xzf Python-3.12.1.tgz
```

```{bash}
#| eval: false
#| code-fold: false
cd Python-3.12.1
```

Then configure it to land in `/opt`, and build:

```{bash}
#| eval: false
#| code-fold: false
sudo ./configure --prefix=/opt/python/3.12.1 --enable-optimizations
```

```{bash}
#| eval: false
#| code-fold: false
sudo make altinstall
```

`--prefix` is what keeps the install self contained. `altinstall` is the flag that matters most: it installs `python3.12` without also writing a `python3` symlink that would shadow the system copy.

Confirm you got what you asked for:

```{bash}
#| eval: false
#| code-fold: false
/opt/python/3.12.1/bin/python3.12 --version
```

```{verbatim}
Python 3.12.1
```

```{r}
#| label: co_box_build_time
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "b",
  look = "default",
  hsize = "1.15",
  size = "1.00",
  header = "This takes a while",
  fold = TRUE,
  contents = "
`--enable-optimizations`, passed to `./configure` above, is what triples the compile time. It builds Python twice: once to produce an instrumented interpreter, then runs it against Python's own test suite to record which code paths get hit most, then rebuilds using that profile data to optimize those paths (profile guided optimization, or PGO). It also turns on link time optimization (LTO). On a `t3.micro` the whole thing takes 20 to 30 minutes with the CPU pegged.

'Runtime speed' here means how fast the finished interpreter executes Python code after it's installed, not how long the build itself takes. PGO typically buys 10 to 20 percent on CPU bound benchmarks. splam isn't CPU bound: it spends its time waiting on subprocess calls (`systemctl`, `journalctl`) and network I/O, so that speedup is not something you'd notice in practice. Drop the flag (`sudo ./configure --prefix=/opt/python/3.12.1`) and the build finishes in a few minutes with a standard, unoptimized interpreter that runs splam exactly as well.

If `make` dies partway through with an out of memory error, that's the 1 GB of RAM on a `t3.micro`, not a broken source tree. 

Either size up temporarily (see [Instance Size](instance-size.qmd)) or add swap. 

More on the flag itself: <https://docs.python.org/3/using/configure.html#cmdoption-enable-optimizations>
  "
)
```

## Installing splam into `/opt`

The app itself follows the same convention as the interpreter. It's third party software with its own directory tree, so it goes in `/opt/splam`.

Clone the repository:

```{bash}
#| eval: false
#| code-fold: false
sudo git clone https://github.com/mjfrigaard/splam.git /opt/splam
```

```{bash}
#| eval: false
#| code-fold: false
cd /opt/splam
```

Create the virtual environment from the `/opt` Python, not the system one:

```{bash}
#| eval: false
#| code-fold: false
sudo /opt/python/3.12.1/bin/python3.12 -m venv /opt/splam/.venv
```

Then install the package and its dependencies into that environment:

```{bash}
#| eval: false
#| code-fold: false
sudo /opt/splam/.venv/bin/pip install -e .
```

When it's complete, you should see something like the following: 

```{verbatim}
Successfully built splam
Installing collected packages: websockets, wcwidth, typing-extensions, setuptools, python-multipart, platformdirs, packaging, orjson, narwhals, mdurl, linkify-it-py, libsass, idna, h11, click, asgiref, annotated-types, uvicorn, typing-inspection, pydantic-core, prompt-toolkit, opentelemetry-api, markdown-it-py, htmltools, anyio, watchfiles, starlette, questionary, pydantic, mdit-py-plugins, shinychat, shiny, splam
Successfully installed annotated-types-0.8.0 anyio-4.15.1 asgiref-3.12.1 click-8.5.0 h11-0.16.0 htmltools-0.7.0 idna-3.19 libsass-0.23.0 linkify-it-py-2.2.0 markdown-it-py-4.2.0 mdit-py-plugins-0.6.1 mdurl-0.1.2 narwhals-2.26.0 opentelemetry-api-1.44.0 orjson-3.12.0 packaging-26.3 platformdirs-4.11.9 prompt-toolkit-3.0.53 pydantic-2.13.5 pydantic-core-2.46.5 python-multipart-0.0.32 questionary-2.1.1 setuptools-84.0.0 shiny-1.8.0 shinychat-0.7.1 splam-0.1.0 starlette-1.6.0 typing-extensions-4.16.0 typing-inspection-0.4.4 uvicorn-0.53.0 watchfiles-1.2.0 wcwidth-0.8.3 websockets-17.1
```

This pulls Shiny for Python and its transitive dependencies from PyPI, so the instance needs outbound internet access. If you put the box in a private subnet, verify the NAT gateway path works before you run this rather than after it hangs.

```{r}
#| label: co_box_no_ssl
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "r",
  look = "default",
  hsize = "1.15",
  size = "1.00",
  header = "pip fails with 'ssl module is not available'",
  fold = TRUE,
  contents = "
The interpreter you just built has no SSL support, so pip can't reach PyPI over HTTPS at all. The cause traces back to the `./configure` step: Amazon Linux 2's default `openssl-devel` package is OpenSSL 1.0.2k, and Python 3.10 and newer require 1.1.1 or later to build the `_ssl` extension. Configure looked for a new enough OpenSSL, didn't find one, and silently skipped the module rather than failing the build. `make altinstall` finished without complaint because none of the steps you ran so far touch SSL, so the problem doesn't show up until pip tries to talk to the internet.

AL2 ships that newer OpenSSL under a separate package, `openssl11`, but its `-devel` package can't be installed alongside the old one: both own the same header path, `/usr/include/openssl/*.h`, so `yum` refuses with a conflict rather than letting them collide. The fix is to swap the devel package out, build, then swap it back, since Python only needs the headers at build time, not at runtime:

> `sudo yum remove -y openssl-devel`
>
> `sudo yum install -y openssl11-devel`
>
> `cd /tmp/Python-3.12.1`
>
> `sudo ./configure --prefix=/opt/python/3.12.1 --enable-optimizations`
>
> `sudo make altinstall`

Python's configure script checks for `openssl11` via `pkg-config` before falling back to the system default, which is what makes this work without a `--with-openssl` flag. You don't need to redownload the source or delete the existing `/opt/python/3.12.1` tree, `altinstall` overwrites in place.

Confirm the module actually built this time before retrying the venv:

`/opt/python/3.12.1/bin/python3.12 -c 'import ssl; print(ssl.OPENSSL_VERSION)'`

That should print an OpenSSL 1.1.1 version string, not a traceback. Once it does, put the original devel package back so other builds on the box aren't left pointing at `openssl11`:

> `sudo yum remove -y openssl11-devel`
>
> `sudo yum install -y openssl-devel`

Then delete the broken venv and redo the two commands from earlier in this section:

> `sudo rm -rf /opt/splam/.venv`
>
> `sudo /opt/python/3.12.1/bin/python3.12 -m venv /opt/splam/.venv`
>
> `sudo /opt/splam/.venv/bin/pip install -e .`
  "
)
```

When it finishes, `shiny` plus three `splam-*` entry point scripts exist inside the venv:

```{bash}
#| eval: false
#| code-fold: false
ls /opt/splam/.venv/bin | grep -E 'shiny|splam'
```

```{verbatim}
shiny
splam-chat
splam-incidents
splam-users
```

```{r}
#| label: co_box_venv_path
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "b",
  look = "default",
  hsize = "1.15",
  size = "1.00",
  header = "Call by path, not by activation",
  fold = TRUE,
  contents = "
Activating a virtual environment is just a `PATH` edit for the current shell. It doesn't survive logout, and `systemd` never runs a login shell at all, so a unit file that says `ExecStart=shiny run ...` will fail with `command not found` no matter how many times activation worked interactively.

The habit that avoids the whole class of problem is calling the venv binaries by their absolute path: `/opt/splam/.venv/bin/shiny`, `/opt/splam/.venv/bin/splam-users`. That's what goes in the unit file, and it's what I use in every command below.
  "
)
```

## Where splam keeps its files

The install spreads across four locations, and each one follows the convention from [Install Apps](install-apps.qmd) rather than piling everything into one folder.

| Location | What lives there |
|:---------|:-----------------|
| `/opt/splam` | Application source and the virtual environment |
| `/var/lib/splam` | Credentials file and the audit trail |
| `/etc/systemd/system/splam.service` | The unit file telling `systemd` how to run it |
| `journald` | Application stdout and stderr, read with `journalctl -u splam` |

: Where a system install of splam puts things {#tbl-splam-locations}

By default `splam` writes its credentials and audit trail to `~/.local/share/splam/`, which is the right answer for a laptop and the wrong answer for a service. A daemon's home directory is an implementation detail of whatever account it happens to run as, and audit records that move when the service account changes aren't much of an audit trail.

The `SPLAM_DATA_DIR` environment variable moves them somewhere durable:

```{bash}
#| eval: false
#| code-fold: false
sudo mkdir -p /var/lib/splam
```

`/var/lib` is the conventional location for state a service owns and expects to persist across restarts. The app creates the directory mode `0700` if it has to, but making it explicitly up front means the first run doesn't fail for a reason unrelated to the app.

```{mermaid}
%%| fig-cap: 'splam File Locations'
%%| fig-width: 7.5
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace', "fontSize":"12px"}}}}%%

graph TD
    Root(["/"])
    Opt("<strong>/opt/splam</strong>")
    Venv("<strong>.venv/bin/shiny</strong><br/><strong>.venv/bin/splam-users</strong>")
    Var("<strong>/var/lib/splam</strong>")
    Creds("<strong>credentials.json</strong><br/><strong>audit_trail.json</strong>")
    Etc("<strong>/etc/systemd/system</strong>")
    Unit("<strong>splam.service</strong>")
    Log("<strong>journald</strong>")
    Jctl("<em>journalctl -u splam</em>")

    Root --"<em>application</em>"--> Opt --> Venv
    Root --"<em>state</em>"--> Var --> Creds
    Root --"<em>config</em>"--> Etc --> Unit
    Root --"<em>logs</em>"--> Log --> Jctl

    style Root fill:#d2562b,stroke:#fff,stroke-width:2px,color:#fff
    style Opt fill:#e8a33d,stroke:#000,stroke-width:2px,color:#000,font-family:monospace
    style Venv fill:#e8a33d,stroke:#000,stroke-width:1px,color:#000,font-family:monospace
    style Var fill:#5b8c5a,stroke:#fff,stroke-width:2px,color:#fff,font-family:monospace
    style Creds fill:#5b8c5a,stroke:#fff,stroke-width:1px,color:#fff,font-family:monospace
    style Etc fill:#2a6f77,stroke:#fff,stroke-width:2px,color:#fff,font-family:monospace
    style Unit fill:#2a6f77,stroke:#fff,stroke-width:1px,color:#fff,font-family:monospace
    style Log fill:#666,stroke:#999,stroke-width:2px,color:#fff,font-family:monospace
    style Jctl fill:#666,stroke:#999,stroke-width:1px,color:#fff,font-family:monospace
    
```

## Creating the first account

`splam` has no default account, which means a fresh install has a login screen nobody can get past. Create an admin before you daemonize anything, and set `SPLAM_DATA_DIR` on the command so the credentials land where the service will look for them:

```{bash}
#| eval: false
#| code-fold: false
sudo SPLAM_DATA_DIR=/var/lib/splam /opt/splam/.venv/bin/splam-users add admin
```

```{verbatim}
Password:
Confirm password:
User 'admin' saved with role 'admin'
```

The password is salted and hashed, not stored. List what exists at any point:

```{bash}
#| eval: false
#| code-fold: false
sudo SPLAM_DATA_DIR=/var/lib/splam /opt/splam/.venv/bin/splam-users list
```

```{verbatim}
admin     admin
```

Set the `SPLAM_DATA_DIR` as `/var/lib/splam`:

```{bash}
#| eval: false
#| code-fold: false
sudo SPLAM_DATA_DIR=/var/lib/splam
```

If you skip the `SPLAM_DATA_DIR` prefix, the account gets written to `/root/.local/share/splam/` and the service will start with an empty credentials file. Nothing errors, and the symptom is a login screen that rejects a password you know is correct.

There's a second role worth knowing about. An `auditor` account can read Status, Logs, Tasks, and the Audit Trail, but can't start, stop, or restart anything:

```{bash}
#| eval: false
#| code-fold: false
/opt/splam/.venv/bin/splam-users add qa-reviewer --role auditor
```

## Privilege to control services

`systemctl start`, `stop`, and `restart` all require privilege. This is the part of the deployment worth slowing down on, because the app authenticates fine without it and then fails at the moment you actually need it.

Run it as `root` for a first deploy and every action works. That's also the widest possible blast radius: anyone who gets past the login screen can restart anything on the box.

```{r}
#| label: co_box_root
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "r",
  look = "default",
  hsize = "1.15",
  size = "1.00",
  header = "Root is a starting point, not a destination",
  fold = FALSE,
  contents = "
Running the unit as `root` is the fastest way to a working deployment, and it's what I'd do to confirm the plumbing works end to end. It is not what I'd leave in place.

The tighter option is a dedicated service account, scoped to exactly the `systemctl` commands operators need and nothing else. The next section walks through setting one up.
  "
)
```

Whatever you decide, `Status` and `Logs` work at any privilege level. It's only the action buttons that need rights.

## Creating a dedicated service account

A system account exists to run one thing and nothing else, so give it exactly the shape that implies: no login shell, no password, no reason for anyone to ever land a session on it.

```{bash}
#| eval: false
#| code-fold: false
sudo useradd -r -s /sbin/nologin splam
```

`-r` puts it in the system UID range rather than the normal human range, so it never collides with, or gets mistaken for, somebody's actual login. `-s /sbin/nologin` means any attempt to log in as `splam`, over SSH or locally, is rejected outright:

```{bash}
#| eval: false
#| code-fold: false
sudo su - splam
```

```{verbatim}
This account is currently not available.
```

That's the message you want to see. If `su -` into it ever succeeds, the account is doing more than its job requires.

Hand it ownership of the state it writes to, but not the code it runs. `/var/lib/splam` needs to be writable by the service; `/opt/splam` doesn't, and leaving that owned by `root` means a compromised app process can't rewrite its own source or replace the venv it's running from:

```{bash}
#| eval: false
#| code-fold: false
sudo chown -R splam:splam /var/lib/splam
```

Then scope exactly what it's allowed to escalate to. Every `systemctl` verb and unit an operator needs gets its own line, nothing broader:

```{bash}
#| eval: false
#| code-fold: false
sudo visudo -f /etc/sudoers.d/splam
```

```{verbatim}
splam ALL=(root) NOPASSWD: /bin/systemctl restart cron
splam ALL=(root) NOPASSWD: /bin/systemctl status cron
splam ALL=(root) NOPASSWD: /bin/systemctl restart nginx
splam ALL=(root) NOPASSWD: /bin/systemctl status nginx
```

`visudo` checks syntax before it lets the file save, so a typo fails loudly instead of locking `sudo` for everyone. Check it explicitly anyway:

```{bash}
#| eval: false
#| code-fold: false
sudo visudo -c -f /etc/sudoers.d/splam
```

```{verbatim}
/etc/sudoers.d/splam: parsed OK
```

Confirm the account can actually use what you just granted, without a password prompt in the way:

```{bash}
#| eval: false
#| code-fold: false
sudo -u splam sudo -n systemctl status cron
```

Status output back means the rule is wired correctly. Anything not on the list, `sudo -u splam sudo -n systemctl restart sshd` for instance, should be refused rather than prompt for a password it will never get.

```{mermaid}
%%| fig-cap: 'What the Service Account Can Escalate To'
%%| fig-width: 7.5
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace', "fontSize":"12px"}}}}%%

graph TD
    Splam(["splam account<br/><em>no shell, no password</em>"])
    Sudo{"<strong>sudo -n</strong><br/>/etc/sudoers.d/splam"}
    Allowed("<strong>restart / status</strong><br/>cron, nginx")
    Denied(["everything else<br/><em>denied</em>"])

    Splam --> Sudo
    Sudo --"unit is listed"--> Allowed
    Sudo --"unit is not listed"--> Denied

    style Splam fill:#d2562b,stroke:#fff,stroke-width:2px,color:#fff
    style Sudo fill:#e8a33d,stroke:#000,stroke-width:2px,color:#000
    style Allowed fill:#5b8c5a,stroke:#fff,stroke-width:2px,color:#fff
    style Denied fill:#666,stroke:#999,stroke-width:1px,color:#fff
```

```{r}
#| label: co_box_service_account_practices
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "b",
  look = "default",
  hsize = "1.15",
  size = "1.00",
  header = "Best practices for a service account",
  fold = TRUE,
  contents = "
A few things worth holding to for any service account, not just this one:

- No login shell and no password. The account exists to run a process, not to be logged into, so `/sbin/nologin` and a locked password are the default, not an exception you add later.
- System UID range (`-r`). Keeps it out of the human login range and out of any audit that lists who can access the box.
- Own the state, not the code. The account should be able to write to its own data directory and have no write access to the application files it's running.
- Sudoers by allowlist, never a bare `ALL` grant. One line per verb and unit. A new unit an operator needs is a new line, not a broader rule.

Decide which side of it you want to live on before you hand out logins, not after. Retrofitting a scoped account onto a team that's already used to `root` access is a far harder conversation than starting there.
  "
)
```

## Daemonizing with systemd

Running `shiny run` in an SSH session works until the session ends. Making it a `systemd` unit is what turns it into something that starts on boot and comes back after a crash, which is the whole point of the daemonizing discussion in [Lab 10](lab10.qmd).

Create the unit file:

```{bash}
#| eval: false
#| code-fold: false
sudo vim /etc/systemd/system/splam.service
```

And give it this content:

```{verbatim}
[Unit]
Description=splam
After=network.target

[Service]
Type=simple
User=splam
WorkingDirectory=/opt/splam
Environment=SPLAM_DATA_DIR=/var/lib/splam
ExecStart=/opt/splam/.venv/bin/shiny run splam.app:app --host 127.0.0.1 --port 8000
Restart=on-failure
RestartSec=5

[Install]
WantedBy=multi-user.target
```

Five lines in there are doing the work worth understanding.

`User=splam` runs the process as the scoped account from the previous section, not `root`. Anything the app needs to do beyond serving requests, restarting the units listed in `/etc/sudoers.d/splam`, comes from that sudoers rule, not from broader privilege on the process itself.

`ExecStart` uses the absolute path into the venv. There's no shell involved, no `PATH` to search, and no activation step. If this path is wrong the unit fails instantly, and it's the single most common cause of a failed first start.

`--host 127.0.0.1` binds the app to loopback only. Change it to `0.0.0.0` and the app is reachable from the internet without TLS and without the proxy, which is exactly what we don't want.

`Environment=SPLAM_DATA_DIR=/var/lib/splam` is how the service finds the credentials you just created. `systemd` doesn't read `~/.bashrc`, so this is the only place that variable gets set.

`Restart=on-failure` with `RestartSec=5` brings the app back after a crash, waiting five seconds between attempts so a fast crash loop doesn't spin the CPU.

Load the new file and start the service:

```{bash}
#| eval: false
#| code-fold: false
sudo systemctl daemon-reload
```

```{bash}
#| eval: false
#| code-fold: false
sudo systemctl enable --now splam
```

`daemon-reload` tells `systemd` to reread the unit files on disk. `enable` registers the unit to start on boot, and `--now` starts it immediately so you don't have to run `start` separately.

## Confirming it actually came up

Never trust `enable --now` to mean the app is running. It means `systemd` tried.

```{bash}
#| eval: false
#| code-fold: false
sudo systemctl status splam
```

```{verbatim}
● splam.service - splam
   Loaded: loaded (/etc/systemd/system/splam.service; enabled; vendor preset: disabled)
   Active: active (running) since Sat 2026-09-05 18:22:41 UTC; 12s ago
 Main PID: 4812 (python3.12)
   CGroup: /system.slice/splam.service
           └─4812 /opt/splam/.venv/bin/python3.12 /opt/splam/.venv/bin/shiny run splam.app:app --host 127.0.0.1 --port 8000

Sep 05 18:22:41 ip-172-31-24-9 systemd[1]: Started splam.
Sep 05 18:22:42 ip-172-31-24-9 shiny[4812]: INFO:     Started server process [4812]
Sep 05 18:22:42 ip-172-31-24-9 shiny[4812]: INFO:     Uvicorn running on http://127.0.0.1:8000
```

Two things to read here. `Active: active (running)` is the state, and `enabled` in the `Loaded:` line is the confirmation it will come back after a reboot. A unit can be running and not enabled, which works fine until the instance restarts.

Then confirm something is actually listening, from the instance itself:

```{bash}
#| eval: false
#| code-fold: false
curl -I http://127.0.0.1:8000
```

```{verbatim}
HTTP/1.1 200 OK
date: Sat, 05 Sep 2026 18:23:10 GMT
server: uvicorn
content-type: text/html; charset=utf-8
```

A `200` means the login page rendered. A redirect is fine too. Connection refused means the process died between `status` and `curl`, and the journal will say why:

```{bash}
#| eval: false
#| code-fold: false
sudo journalctl -u splam -n 50 --no-pager
```

`-u` filters to this unit, `-n 50` limits it to the last 50 lines, and `--no-pager` dumps it straight to the terminal instead of opening `less`. When you want to watch a start attempt live, `journalctl -u splam -f` follows the log the same way `tail -f` does.

```{r}
#| label: co_box_startup_fail
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "y",
  look = "default",
  hsize = "1.15",
  size = "1.00",
  header = "Reading a failed start",
  fold = TRUE,
  contents = "
Four failures account for almost every bad first start, and each has a distinct signature in `journalctl -u splam`:

| What the journal says | What's actually wrong |
|:----------------------|:----------------------|
| `Failed at step EXEC ... No such file or directory` | The `ExecStart` path is wrong. Check it with `ls /opt/splam/.venv/bin/shiny` |
| `ModuleNotFoundError: No module named 'splam'` | `pip install -e .` ran somewhere other than `/opt/splam`, or into a different venv |
| `Address already in use` | Something already holds `:8000`, often a `shiny run` you started by hand and forgot |
| `status=203/EXEC` with no other output | The venv was built from a Python that no longer exists at that path |

The pattern in all four is the same: `systemd` executes a literal path with no shell, so anything the interactive session was quietly providing has to be spelled out in the unit file.
  "
)
```

## Putting nginx in front

The app is listening on loopback, which means nothing outside the instance can reach it. Something has to accept the connection on `443`, terminate TLS, and forward the request inward.

Install `nginx`:

```{bash}
#| eval: false
#| code-fold: false
sudo amazon-linux-extras install -y nginx1
```

Create the site configuration:

```{bash}
#| eval: false
#| code-fold: false
sudo vim /etc/nginx/conf.d/splam.conf
```

```{verbatim}
server {
    listen 443 ssl;
    server_name splam.example.com;

    ssl_certificate     /etc/letsencrypt/live/splam.example.com/fullchain.pem;
    ssl_certificate_key /etc/letsencrypt/live/splam.example.com/privkey.pem;

    location / {
        proxy_pass http://127.0.0.1:8000;
        proxy_http_version 1.1;
        proxy_set_header Upgrade $http_upgrade;
        proxy_set_header Connection "upgrade";
        proxy_set_header Host $host;
    }
}
```

The certificate paths assume you've been through [Configure SSL](configure-ssl.qmd) and have a hostname pointed at the instance from [Configure DNS](configure-dns.qmd). An Elastic IP is worth allocating here, because without one a stopped and restarted instance comes back with a new address and the DNS record goes stale.

Test the configuration before reloading, since a syntax error will otherwise take `nginx` down entirely:

```{bash}
#| eval: false
#| code-fold: false
sudo nginx -t
```

```{verbatim}
nginx: the configuration file /etc/nginx/nginx.conf syntax is ok
nginx: configuration file /etc/nginx/nginx.conf test is successful
```

Then enable and start it, the same two step pattern as the app:

```{bash}
#| eval: false
#| code-fold: false
sudo systemctl enable --now nginx
```

```{r}
#| label: co_box_websocket
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default",
  hsize = "1.15",
  size = "1.00",
  header = "Shiny needs the WebSocket headers",
  fold = TRUE,
  contents = "
The `proxy_http_version`, `Upgrade`, and `Connection` lines are not boilerplate. Shiny for Python holds its connection to the browser open over a WebSocket rather than plain request and response HTTP.

Leave them out and the failure is confusing rather than obvious: the page loads, the layout renders, and then nothing responds. No error appears in the browser, and `journalctl -u splam` shows a session opening and immediately closing. Any Shiny app behind any proxy needs this, in R or in Python.
  "
)
```

Load the hostname in a browser and you should get the login screen, served over HTTPS, backed by the account you created earlier.

## Recap

1. `splam` calls `systemctl` and `journalctl` on its own host, so it runs directly on the EC2 instance rather than in a container. That single constraint is what makes this a `systemd` plus reverse proxy deployment instead of a Docker one.

2. Amazon Linux 2's Python is too old and belongs to `yum`. Build 3.12 into `/opt/python/3.12.1` with `make altinstall`, put the app in `/opt/splam`, its state in `/var/lib/splam`, and leave the system interpreter alone.

3. `systemd` runs no shell and reads no startup files. Absolute paths in `ExecStart`, `Environment=` lines for anything the app expects from the environment, and `--host 127.0.0.1` so the only way in is through the proxy.

4. Verify in three places before you call it deployed: `systemctl status` for the unit state, `curl -I http://127.0.0.1:8000` for the app, and the browser over HTTPS for the proxy. When one of them fails, `journalctl -u splam -n 50` names the reason.