# <em>Install Apps</em> {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default",
  hsize = "1.25",
  size = "1.00",
  header = "Caution",
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

## Finding your applications

Installing an application and being able to *run* it are two different problems. A package manager like `yum` puts binaries somewhere the shell already knows to look, but anything I install by hand usually lands somewhere the shell doesn't check. That's when a command I know is installed comes back as `command not found`.

### The `PATH` variable

`PATH` is an environment variable holding a colon separated list of directories. When I type a command, the shell walks that list from left to right and runs the first matching executable it finds. It never searches the whole filesystem.

Print it to see what the shell is actually checking:

```{bash}
#| eval: false
#| code-fold: false
echo $PATH
```

```{verbatim}
/usr/local/bin:/usr/bin:/usr/local/sbin:/usr/sbin:/home/ec2-user/.local/bin:/home/ec2-user/bin
```

To find out which copy wins, use `which`:

```{bash}
#| eval: false
#| code-fold: false
which python3
```

```{verbatim}
/usr/bin/python3
```

If a command isn't found, the executable is almost always on the box, just not in one of those directories. Adding it is a matter of appending the directory to `PATH` in a startup file so the change survives a new login:

```{bash}
#| eval: false
#| code-fold: false
echo 'export PATH="/opt/python/3.12.1/bin:$PATH"' >> ~/.bashrc
source ~/.bashrc
```

Putting the new directory first means it takes priority over anything with the same name further down the list.

```{mermaid}
%%| fig-cap: 'How PATH Search Works'
%%| fig-width: 8.0
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}}%%

graph TD
    Cmd(["shell command<br/><strong>python3</strong>"]) --> Check1("Check<br/><strong>/usr/local/bin</strong>")
    Check1 -->|"not found"| Check2("Check<br/><strong>/usr/bin</strong>")
    Check2 -->|"found!"| Run("Run<br/><strong>/usr/bin/python3</strong>")
    Check1 -->|"found!"| Run2("Run<br/><strong>/usr/local/bin/python3</strong><br/>(first match wins)")

    style Cmd fill:#d2562b,stroke:#fff,stroke-width:2px,color:#fff
    style Check1 fill:#e8a33d,stroke:#000,stroke-width:2px,color:#000
    style Check2 fill:#e8a33d,stroke:#000,stroke-width:2px,color:#000
    style Run fill:#5b8c5a,stroke:#fff,stroke-width:2px,color:#fff
    style Run2 fill:#5b8c5a,stroke:#fff,stroke-width:2px,color:#fff
    
```

```{r}
#| label: co_box_path_scope
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "b",
  look = "default",
  hsize = "1.15",
  size = "1.00",
  header = "PATH is per user",
  fold = TRUE,
  contents = "
Editing `~/.bashrc` only changes `PATH` for that one account. A service running under `systemd` doesn't read it at all, and neither does a different user who logs in later. If an application has to be visible to everyone (and to services), a symlink into `/usr/local/bin` is more reliable than asking every account to edit a startup file.
  "
)
```

### Symlinks with `ln`

A symlink is a pointer: a small file that redirects to a real one somewhere else. Rather than adding a new directory to everyone's `PATH`, I create a link inside a directory that's already there.

The `-s` flag makes it symbolic, and the real path always comes first:

```{bash}
#| eval: false
#| code-fold: false
sudo ln -s /opt/python/3.12.1/bin/python3 /usr/local/bin/python3
```

Confirm the link points where you expect:

```{bash}
#| eval: false
#| code-fold: false
ls -l /usr/local/bin/python3
```

```{verbatim}
lrwxrwxrwx 1 root root 30 Aug 21 14:02 /usr/local/bin/python3 -> /opt/python/3.12.1/bin/python3
```

```{mermaid}
%%| fig-cap: 'Symlink Resolution'
%%| fig-width: 7.0
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}}%%

graph TD
    User["User types:<br/><code>python3</code>"]
    Link["<strong>/usr/local/bin/python3</strong>"]
    Points["<strong>/opt/python/3.12.1/<br>bin/python3</strong>"]
    Real["<strong>Actual binary</strong>"]

    User --"Shell finds symlink"--> Link
    Link --"Symlink points to"--> Points
    Points --"Real executable runs"--> Real

    style User fill:#d2562b,stroke:#fff,stroke-width:2px,color:#fff
    style Link fill:#2a6f77,stroke:#fff,stroke-width:1px,color:#fff
    style Points fill:#e8a33d,stroke:#000,stroke-width:1px,color:#000
    style Real fill:#5b8c5a,stroke:#fff,stroke-width:2px,color:#fff
```

The other reason I reach for `ln` is versioning. When a new version of R or Python goes into `/opt` alongside the old one, repointing a single symlink switches every user over, and repointing it back is the rollback.

## Where do applications live?

Linux spreads a single application across several directories by purpose, not by product. Knowing the convention is how I find a config file or a log without searching for it.

| Location | What lives there |
|:---------|:-----------------|
| `/bin`, `/usr/bin` | Core system executables, managed by the package manager |
| `/usr/local/bin` | Executables installed outside the package manager, plus symlinks |
| `/opt` | Self contained third party software, one directory per version |
| `/etc` | Configuration files |
| `/var/log` | Log files |

: Common application locations {#tbl-app-locations}

```{mermaid}
%%| fig-cap: 'Application Filesystem Hierarchy'
%%| fig-width: 7.0
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}}%%

graph TD
    Root(["/"])
    Bin("<strong>/bin</strong> or<br><strong>/usr/bin</strong>")
    Local("<strong>/usr/local/bin</strong>")
    Opt("<strong>/opt</strong>")
    OptPy("<strong>/opt/python/3.12.1</strong>")
    OptR("<strong>/opt/R/4.4.1</strong>")
    Etc("<strong>/etc</strong>")
    Log("<strong>/var/log</strong>")

    Root --"<em>OS managed</em>"--> Bin
    Root --"<em>symlinks here</em>"--> Local
    Root --"<em>versions isolated</em>"--> Opt
    Root --"<em>config files</em>"--> Etc
    Root --"<em>log files</em>"--> Log
    Opt --"<em>lib, bin, share</em>"--> OptPy
    Opt --"<em>lib, bin, share</em>"--> OptR

    style Root fill:#d2562b,stroke:#fff,stroke-width:2px,color:#fff,font-size:12px,font-family:monospace
    style Bin fill:#2a6f77,stroke:#fff,stroke-width:1px,color:#fff,font-size:12px,font-family:monospace
    style Local fill:#5b8c5a,stroke:#fff,stroke-width:1px,color:#fff,font-size:12px,font-family:monospace
    style Opt fill:#e8a33d,stroke:#000,stroke-width:2px,color:#000,font-size:12px,font-family:monospace
    style OptPy fill:#e8a33d,stroke:#000,stroke-width:1px,color:#000,font-size:12px,font-family:monospace
    style OptR fill:#e8a33d,stroke:#000,stroke-width:1px,color:#000,font-size:12px,font-family:monospace
    style Etc fill:#666,stroke:#999,stroke-width:1px,color:#fff,font-size:12px,font-family:monospace
    style Log fill:#666,stroke:#999,stroke-width:1px,color:#fff,font-size:12px,font-family:monospace
    
```

`/opt` is the important one for data science work. Installing there keeps a version fully contained in its own folder, so nothing overwrites anything the operating system depends on.

## Language versions

The Python already on the instance belongs to the operating system. Amazon Linux 2 uses it for `yum` and other internal tooling, so upgrading it or installing packages into it with `sudo pip install` risks breaking the box itself.

The safer pattern is to install your own Python into `/opt`, leave the system copy alone, and create virtual environments from the `/opt` copy:

```{bash}
#| eval: false
#| code-fold: false
/opt/python/3.12.1/bin/python3 -m venv ~/my-app/venv
source ~/my-app/venv/bin/activate
```

R has no equivalent system dependency, so multiple versions can coexist in `/opt/R` without conflict. [`rig`](https://github.com/r-lib/rig), the R Installation Manager, handles installing and switching between them:

```{bash}
#| eval: false
#| code-fold: false
sudo rig add 4.4.1
sudo rig default 4.4.1
rig list
```

Both languages depend on system libraries that neither one installs for you. A package that compiles from source will fail with a missing header error until the matching development package is present:

```{bash}
#| eval: false
#| code-fold: false
sudo yum install -y libcurl-devel openssl-devel libxml2-devel
```

These are the three that account for most failed installs of `curl`, `openssl`, and `xml2`, and every R or Python package that depends on them.

```{mermaid}
%%| fig-cap: 'Language Runtime Installation Pattern'
%%| fig-width: 8.0
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}}%%

graph TD
    SysPy(["System Python<br/><em>owned by OS</em>"])
    OptPy(["Your Python<br/><strong>/opt/python/3.12.1</strong><br/><em>isolated version</em>"])
    Venv("Virtual Env<br/><strong>~/my-app/venv</strong><br/><em>project packages</em>")

    SysPy -->|"leave untouched"| SysTools("yum, AWS CLI,<br/>OS internal tools")
    OptPy -->|"create venv from"| Venv
    Venv -->|"install packages to"| Packages("<strong>pip install<br/>my-packages</strong>")

    R(["System R<br/><em>usually absent</em>"])
    OptR("Multiple R versions<br/><strong>/opt/R/4.4.1</strong><br/><strong>/opt/R/4.3.2</strong><br/><em>rig switches</em>")
    Libs("System libraries<br/><strong>libcurl-devel</strong><br/><strong>openssl-devel</strong><br/><strong>libxml2-devel</strong>")

    R -->|"alternative"| OptR
    OptR --> Libs
    Libs -->|"required for"| RPackages("R packages<br/>from source")

    style SysPy fill:#d2562b,stroke:#fff,stroke-width:2px,color:#fff
    style OptPy fill:#5b8c5a,stroke:#fff,stroke-width:2px,color:#fff
    style Venv fill:#5b8c5a,stroke:#fff,stroke-width:2px,color:#fff
    style Packages fill:#2a6f77,stroke:#fff,stroke-width:1px,color:#fff
    style R fill:#d2562b,stroke:#fff,stroke-width:2px,color:#fff
    style OptR fill:#5b8c5a,stroke:#fff,stroke-width:2px,color:#fff
    style Libs fill:#e8a33d,stroke:#000,stroke-width:2px,color:#000
    style RPackages fill:#2a6f77,stroke:#fff,stroke-width:1px,color:#fff
    style SysTools fill:#2a6f77,stroke:#fff,stroke-width:1px,color:#fff
```

## Recap

1. `PATH` decides what the shell can find, and `which` tells you which copy is winning. A `command not found` error usually means a directory is missing from the list, not that the software is missing.

2. Install language runtimes into `/opt` and symlink them into `/usr/local/bin`. That keeps the system Python untouched, makes the version visible to every account and service, and turns a version change into a one line edit.