# <strong>Lab 1: Environments as Code</strong> {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: false
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

The [first lab](https://do4ds.com/chapters/sec1/1-1-env-as-code.html#lab01) includes two Quarto documents: 

```{r}
#| collapse: true
#| eval: false
#| echo: false
fs::dir_tree(path = "_labs/lab01/", recurse = 1)
```

```{verbatim}
_labs/lab01/
├── eda-basic.qmd
└── model-basic-py.qmd
```


## Project-oriented workflows

Key takeaways: 

1. Each project should have an **isolated** package environment that preserves the names and versions. 

2. The project's dependencies/requirements should be captured in a way that **makes it exportable/transferable to another machine**.[^renv-venv]

3. Each project should have a **virtual environment** in a **standalone directory** that documents everything the project needs (usually in a GitHub or GitLab repo).[^rstats-wtf-projects]

[^rstats-wtf-projects]: Read more in [What They Forgot to Teach You About R: Chapter 3, Project-oriented workflow](https://rstats.wtf/projects)

[^renv-venv]: In R, this is [`renv`](https://rstudio.github.io/renv/index.html). In Python, this is [`venv`](https://packaging.python.org/en/latest/guides/installing-using-pip-and-virtual-environments/#creating-a-virtual-environment).

## Virtual environments

`venv` and `renv`, "*keep user-level package caches of the actual packages and use symlinks, so only one copy of the package is installed.*"

### `renv`

First, read the documentation on [project level management with `renv`](https://rstudio.github.io/renv/index.html) and [using renv with Quarto](https://quarto.org/docs/projects/virtual-environments.html#using-renv).

I’ve summarized the steps below:

```r
# create a new R project-local environment
install.packages("renv")
renv::init()

# install packages into the renv-managed environment
install.packages(c("tidyverse", "shiny", "ggplot2"))

# snapshot the state of the environment and package versions in 
# the lock file (renv.lock)
renv::snapshot()

# to deactivate the environment, restart R or close the session
# renv projects are automatically reactivated the time a project
# is opened

# restore the environment on a new system with
renv::restore()
```

To demonstrate how to use `renv`, I've copied the `eda-basic.qmd` file into it's own 'project' folder (`renv-eda-basic`):

```{verbatim}
_labs/lab01/renv-eda-basic/
└── eda-basic.qmd
```

Then I ran `renv::init()`: 

```{r}
#| eval: false
renv::init()
```

```{r}
#| collapse: true
#| eval: false
#| echo: false
fs::dir_tree(
  path = "_labs/lab01/renv-eda-basic/", 
  all = TRUE, 
  recurse = 1)
```

Initiating `renv` in a project root directory will create the `.Rprofile` and `renv.lock` files and a `renv/` folder:

```{verbatim}
_labs/lab01/renv-eda-basic/
├── .Rprofile
├── eda-basic.qmd
├── renv/
├── renv-eda-basic.Rproj
└── renv.lock
```

`renv::init()` will also create `renv-eda-basic.Rproj`, which helps enforce project-oriented workflows (but these are specific to RStudio).  

To install the packages, we can use `install.packages()` or `pak::pak()` for more flexible installations:

```{r}
#| eval: false
install.packages("pak")
pak::pak(c("tidyverse", "shiny", "ggplot2"))
```

```{r}
#| label: co_box_renv_install
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "b",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "**Note**", 
  fold = FALSE,
  contents = "`renv::install()` will install cached packages if they are available if not, it will install the latest version from CRAN.")
```

Create a snapshot (i.e., the current versions of R and any installed packages, along with the R repositories) with `renv::snapshot()`:

```{r}
#| eval: false
renv::snapshot()
```

The dependencies are saved in the `renv.lock` file, which should be committed to version control. 

To make sure the dependencies in the `renv.lock` file have been installed (i.e., after cloning), run: 

```{r}
#| eval: false
renv::restore()
```

```{mermaid}
%%| fig-cap: 'renv'
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%

flowchart TD
    %% External sources
    CRAN("CRAN/<br>GitHub/...")
    
    %% Local components
    SysLib("system<br>library")
    ProjLib("project<br>library")
    RenvCache("renv<br>cache")
    Lockfile("lockfile")
    
    %% Subgraph for "your computer" vs "the internet"
    subgraph Internet ["<strong>Internet</strong>"]
        CRAN
    end
    
    subgraph Computer ["<strong>Computer</strong>"]
        SysLib
        ProjLib
        RenvCache
        Lockfile
    end
    
    %% Connections with function labels
    SysLib -->|"<strong>init()<strong>"| ProjLib
    
    CRAN -->|"<strong>install()</strong><br><strong>update()</strong>"| RenvCache
    RenvCache -->|"renv"| ProjLib
    
    ProjLib <-->|"<strong>status()</strong>"| Lockfile
    ProjLib -->|"<strong>snapshot()</strong>"| Lockfile
    Lockfile -->|"<strong>restore()</strong>"| ProjLib
    
    %% styling
    %% classDef default fill:#F2E6C2,stroke:#2A6F77,color:#1B2A41
    style CRAN fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style SysLib fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    style RenvCache fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style ProjLib fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    style Lockfile fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style Internet fill:#fbf7ec,stroke:#5B8C5A,color:#1B2A41
    style Computer fill:#fbf7ec,stroke:#2A6F77,color:#1B2A41
    
    
```


### `venv`

Read the [documentation on Python virtual environments](https://packaging.python.org/en/latest/guides/installing-using-pip-and-virtual-environments/#creating-a-virtual-environment) and using [venv with Quarto](https://quarto.org/docs/projects/virtual-environments.html#using-venv). 

I've summarized the steps below: 

```bash
# create new Python 3 virtual environment in the '.venv' directory 
python3 -m venv .venv
# activate the environment
source .venv/bin/activate
# use pip to install packages into your environment
python3 -m pip install jupyter matplotlib pandas
# deactivate an environment
deactivate
# save environment
python3 -m pip freeze > requirements.txt
# restoring environment
python3 -m pip install -r requirements.txt
```

Use the **Terminal** to navigate to the `venv-model-basic-py` folder and create the Python virtual environment: 

```bash
python3 -m venv .venv
```

This will create a `.venv/` folder inside `venv-model-basic-py`.

```{r}
#| collapse: true
#| eval: false
#| echo: false
fs::dir_tree(
  path = "_labs/lab01/venv-model-basic-py/", 
  all = TRUE, 
  recurse = 1)
```

```{verbatim}
_labs/lab01/venv-model-basic-py/
├── .venv
│   ├── bin
│   ├── include
│   ├── lib
│   └── pyvenv.cfg
└── model-basic-py.qmd
```


To activate the virtual environment, run the following in the **Terminal**:

```bash
source .venv/bin/activate
```

This should change the **Terminal** prompt from: 

```bash
Bash@MacBook venv-model-basic-py$ 
```

to: 

```bash
(.venv) Bash@MacBook venv-model-basic-py$
```

Now we can install required Python packages (we include `jupyter` for interactive outputs) 

```bash
pip install palmerpenguins pandas numpy scikit-learn jupyter
```

To store the installed packages 

```{mermaid}
%%| fig-cap: 'venv'
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%

flowchart TD
    %% External sources
    PyPI("PyPI/<br>GitHub/...")
    
    %% Local components
    SysPython("system<br>Python")
    VirtEnv("virtual<br>environment")
    PipCache("pip<br>cache")
    Requirements("requirements.txt")
    
    %% Subgraph for "your computer" vs "the internet"
    subgraph Internet ["<strong>Internet</strong>"]
        PyPI
    end
    
    subgraph Computer ["<strong>Computer</strong>"]
        SysPython
        VirtEnv
        PipCache
        Requirements
    end
    
    %% Connections with function labels
    SysPython -->|"<strong>python -m venv</strong>"| VirtEnv
    
    PyPI -->|"<strong>pip install</strong><br><strong>pip upgrade</strong>"| PipCache
    PipCache -->|"pip"| VirtEnv
    
    VirtEnv <--->|"<strong>pip list</strong><br><strong>pip show</strong>"| Requirements
    VirtEnv -->|"<strong>pip freeze</strong>"| Requirements
    Requirements -->|"<strong>pip install -r</strong>"| VirtEnv
    
    %% Additional workflow arrows
    VirtEnv -.->|"<strong>source activate</strong><br><strong>deactivate</strong>"| VirtEnv
    
    %% Styling
    style PyPI fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style SysPython fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style VirtEnv fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    style PipCache fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style Requirements fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style Internet fill:#fbf7ec,stroke:#5B8C5A,color:#1B2A41
    style Computer fill:#fbf7ec,stroke:#2A6F77,color:#1B2A41
    
```

#### Rendering with Python

To render, use the following in the **Terminal**: 

```bash
quarto preview model-basic-py.qmd --no-browser --no-watch-inputs
```

To tell Quarto to use the virtual environment, we can place the following in the `model-basic-py.qmd` file:

```yaml
execute:
  python: .venv/bin/python
```

Or we can set the Python interpreter with the **Terminal**:[^venv-win]

```bash
quarto use python .venv/bin/python3
```

[^venv-win]: On Windows try `.venv\Scripts\python.exe`


If you're working inside **RStudio or Posit Workbench**, consider letting `reticulate` manage the Python environment:

```{r}
#| eval: false
library(reticulate)
use_virtualenv("myenv", required = TRUE)
virtualenv_install(
  envname = "myenv", 
  packages = c("palmerpenguins", "pandas", "numpy", 
               "scikit-learn", "jupyter"))
```

This ensures RStudio uses a known-good Python environment.

Now, if we open the `model-basic-py.qmd` file in RStudio and click the **Render** button in the toolbar we should see the rendered output. 




