# Lab 6: Demystifying Docker {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

## Comprehension questions

The questions below come from the [Demystifying Docker](https://do4ds.com/chapters/sec1/1-6-docker.html) chapter.

### Docker mental map

> "*Draw a mental map of the relationship between the following: Dockerfile, Docker Image, Docker Registry, Docker Container.*"

1.  A **Dockerfile** is a plain-text recipe that describes exactly what goes into an image.
2.  Running `docker build` executes the **Dockerfile** instructions and produces a **Docker Image** (an immutable, layered snapshot of the filesystem and configuration).
3.  **Docker Images** can be pushed to a **Docker Registry** (such as Docker Hub or GitHub Container Registry) for storage and sharing.
4.  Running `docker run` on an image creates a **Docker Container**, a live, isolated process that uses the image as its read-only base.

```{mermaid}
%%| fig-cap: ‘Docker components’
%%| fig-width: 7.0
%%| fig-align: 'center'
%%{init: {‘theme’: ‘base’, ‘themeVariables’: {‘fontFamily’: ‘monospace’}}}%%

graph LR
    subgraph Build["<strong>Build</strong>"]
        DF("<strong>Dockerfile</strong>")
        Img("<strong>Docker Image</strong>")
        DF -->|"<code>docker<br>build</code>"| Img
    end

    subgraph Share["<strong>Share</strong>"]
        Reg("<strong>Docker<br>Registry<strong><br>(Docker Hub, GHCR)")
        Img -->|"<code>docker<br>push</code>"| Reg
        Reg -->|"<code>docker<br>pull</code>"| Img
    end

    subgraph Run["<strong>Run</strong>"]
        Con("<strong>Docker<br>Container</strong>")
        Img -->|"<code>docker<br>run</code>"| Con
    end

    style Build fill:#fbf7ec,stroke:#5B8C5A,color:#1B2A41
    style Share fill:#fbf7ec,stroke:#2A6F77,color:#1B2A41
    style Run fill:#fbf7ec,stroke:#2A6F77,color:#1B2A41
    style DF fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style Img fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style Reg fill:#1B2A41,stroke:#000000,stroke-width:1px,color:#ffffff
    style Con fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
```

The key distinction is that an image is **static** (it never changes after it is built) while a container is **ephemeral** (it runs, does work, and can be stopped and removed). Multiple containers can be created from the same image simultaneously, each isolated from the others.

### Docker run flags

> "*When would you want to use each of the following flags for docker run? When wouldn’t you? `-p`, `--name`, `-d`, `--rm`, `-v`*"

| Flag | Use | Don't Use |
|--------------------|--------------------------|------------------------|
| `-p <host>:<container>` | The container runs a service (API, Shiny app, database) that needs to be reachable from the host or network | The container does not expose a network service and runs entirely internally |
| `--name <name>` | You need to reference the container by name in other commands (`docker logs`, `docker stop`, `docker exec`), or when running multiple containers that need to find each other | Running a quick one-off command where the auto-generated name is fine |
| `-d` | Running a long-lived background service (a web server, an API) where you want the terminal back immediately | You need to watch live output, or the container runs a short job and you want to see when it finishes |
| `--rm` | Short-lived, disposable jobs: running a script, rendering a document, or any task where you do not need to inspect the container afterward | Long-lived services you want to restart later, or when you may need to examine the stopped container to debug a failure |
| `-v <host_path>:<container_path>` | Persisting data beyond the container’s lifetime, sharing config files or secrets, or mounting code during development so changes are reflected without rebuilding | The container should be fully self-contained and stateless; mounting volumes couples the container to the host filesystem and can undermine reproducibility |

### Dockerfile commands

> "*What are the most important Dockerfile commands?*"

Every Dockerfile starts with `FROM`, which names the base image (e.g., `FROM rocker/r-ver:4.5`). All subsequent instructions build on top of it.

| Command | Purpose |
|--------------|---------------------------------------------------|
| `FROM` | Set the base image — always the first instruction |
| `RUN` | Execute shell commands during the build (install packages, create directories) |
| `COPY` | Copy files from the build context into the image |
| `WORKDIR` | Set the working directory for all subsequent `RUN`, `COPY`, `CMD`, and `ENTRYPOINT` instructions |
| `ENV` | Set environment variables available at both build time and runtime |
| `EXPOSE` | Document which port the container listens on (informational — does not publish the port) |
| `CMD` | Default command to run when the container starts; can be overridden at `docker run` |
| `ENTRYPOINT` | Main executable for the container; `CMD` provides its default arguments |

The most important distinction is between `CMD` and `ENTRYPOINT`. `CMD` is the default command and is easily replaced (`docker run my-image Rscript other.R`). `ENTRYPOINT` locks in the executable so the container always behaves like that program (`docker run my-api --port 8080` passes `--port 8080` as arguments to the entrypoint). For data science services, `ENTRYPOINT` is the right choice when the container should always start the API or app.

## Docker Containers for Model Prediction API

This lab involves creating Docker containers for the APIs we created in R and Python in [lab 2](https://do4ds.com/chapters/sec1/1-2-proj-arch.html#lab2) (see my solutions [here](https://github.com/mjfrigaard/do4ds-labs/tree/main/_labs/lab02)).

## Python 

## R 
