# <strong>Lab 3: Databases and APIs</strong> {.unnumbered} 

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

The original files in [`_labs/lab03`](https://github.com/alexkgold/do4ds/tree/main/_labs/lab3) are below: 

```{bash}
#| code-fold: false
#| eval: false
├── app-api.py #<1>
├── app-api.R #<2>
├── eda-db.qmd #<3>
├── model-db.qmd #<3>
├── py-data-load.py #<4>
└── r-data-load.R #<5>
```
1. Python Shiny app 
2. R Shiny app 
3. EDA files (quarto)
4. Database connection via R   
5. Database connection via Python   

In the previous labs, we build a Python model API with [`vetiver`](https://rstudio.github.io/vetiver-python/stable/) (I also repeated this [lab in R](https://rstudio.github.io/vetiver-r/)). This lab continues with accessing databases and using APIs. 

## Comprehension questions

### Databases

> ***1. Draw two mental maps for connecting to a database, one using a database driver (in a Python or R package) vs. an ODBC or JDBC driver. You should (at a minimum) include the nodes: database package, DBI (R only), driver, system driver, ODBC, JDBC, and database.***

#### Python and databases 

In Python, there is a large database ecosystem (i.e., many driver options). ORM support with  [SQLAlchemy](https://www.sqlalchemy.org/) and it's flexible (i.e., it can bypass standards if needed). 

```{mermaid}
%%| fig-cap: 'Python DB Connections'
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%

graph TB
    subgraph PyApp["Python App"]
        PyCode("<strong>app-api.py</strong>")
    end

    
    subgraph PyPkg["<strong>Python Packages</strong>"]
        DBPkg("DB Package<br><strong>psycopg2</strong>, <strong>pymysql</strong>, <strong>sqlite3</strong>")
        ODBCPkg("ODBC Package<br><strong>pyodbc</strong>")
        JDBCPkg("JDBC Package<br><strong>jaydebeapi</strong>")
    end
    
    %%% Tints of #35605a

    subgraph Driver["<strong>Driver Layer</strong>"]
        NatDrv("Native Driver<br>libpq, MySQL C API")
        ODBCDrv("ODBC Driver<br>Vendor-specific")
        JDBCDrv("JDBC Driver<br>.jar files")
    end

    %% Tints of #6b818c
    
    subgraph System["<strong>System Layer</strong>"]
        ODBCSys("ODBC Manager<br>System ODBC")
        JAVASys("Java Runtime<br>JVM/JRE")
    end

    %% Tints of #00120b
    
    subgraph DB["Database"]
        DBServ("<strong>Database Server</strong>")
    end

    
    %% Direct pathway (bypasses system layer)
    PyCode --> DBPkg
    DBPkg --> NatDrv
    NatDrv --"`Bypass <strong>System Layer</strong>`"--> DBServ
    
    %% ODBC pathway
    PyCode --> ODBCPkg
    ODBCPkg --> ODBCDrv
    ODBCDrv --> ODBCSys
    ODBCSys --> DBServ
    
    %% JDBC pathway
    PyCode --> JDBCPkg
    JDBCPkg --> JDBCDrv
    JDBCDrv --> JAVASys
    JAVASys --> DBServ
    style PyApp fill:#fbf7ec,stroke:#5B8C5A,color:#1B2A41
    style PyPkg fill:#fbf7ec,stroke:#5B8C5A,color:#1B2A41
    style Driver fill:#fbf7ec,stroke:#2A6F77,color:#1B2A41
    style System fill:#fbf7ec,stroke:#2A6F77,color:#1B2A41
    style DB fill:#fbf7ec,stroke:#2A6F77,color:#1B2A41
    style PyCode fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style DBPkg fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style ODBCPkg fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style JDBCPkg fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style NatDrv fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style ODBCDrv fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style JDBCDrv fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style ODBCSys fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style JAVASys fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style DBServ fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    

```


#### R and databases 

In R, the **`DBI` package provides a universal interface** (same code works across drivers). This is a simpler model because all paths converge on `DBI` package. There is also better integration with R's data structures (i.e., `data.frame`s).

```{mermaid}
%%| fig-cap: 'R Database Connection Pathways'
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%

graph TB
    subgraph RApp["<strong>R App</strong>"]
        A("<strong>app-api.R</strong>")
    end

    
    subgraph RPkgs["<strong>R Packages</strong>"]
        B("DB Packages<br><strong>RPostgres</strong>, <strong>RMySQL</strong>, <strong>RSQLite</strong>")
        C("ODBC Package<br><strong>odbc</strong>")
        D("JDBC Package<br><strong>RJDBC</strong>")
    end

    %%% Tints of #35605a
    
    subgraph DBI["<strong>R Standard Interface</strong>"]
        E("DBI<br>Database Interface")
    end

    %% Tints of #6b818c

    
    subgraph Driver["<strong>Driver Layer</strong>"]
        F("Native Driver<br>libpq, MySQL C API")
        G("ODBC Driver<br>Vendor-specific")
        H("JDBC Driver<br>.jar files")
    end

    %% Tints of #6b818c
    
    subgraph System["<strong>System Layer</strong>"]
        I("ODBC Manager<br>System ODBC")
        J("Java Runtime<br>JVM/JRE")
    end

    
    subgraph DB["<strong>Database</strong>"]
        K("<strong>Database Server</strong>")
    end

    
    %% All paths go through DBI
    A --> B
    A --> C  
    A --> D
    B --> E
    C --> E
    D --> E
    
    %% Direct pathway (bypasses system layer)
    E --> F
    F --"bypasses system layer"--> K
    
    %% ODBC pathway  
    E --> G
    G --> I
    I --> K
    
    %% JDBC pathway
    E --> H
    H --> J
    J --> K
    style RApp fill:#fbf7ec,stroke:#5B8C5A,color:#1B2A41
    style RPkgs fill:#fbf7ec,stroke:#5B8C5A,color:#1B2A41
    style DBI fill:#fbf7ec,stroke:#5B8C5A,color:#1B2A41
    style Driver fill:#fbf7ec,stroke:#2A6F77,color:#1B2A41
    style System fill:#fbf7ec,stroke:#2A6F77,color:#1B2A41
    style DB fill:#fbf7ec,stroke:#2A6F77,color:#1B2A41
    style A fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style B fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style C fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style D fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style E fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style F fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style G fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style H fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style I fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style J fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style K fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    
```

Both R and Python native database drivers bypass the system layer and connect directly to databases. The system layer is only involved when using ODBC or JDBC connections.

```{mermaid}
%%| fig-cap: 'DB Connection Mental Model Differences'
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%

graph LR
    subgraph RPhil["R Philosophy"]
        A("DBI as Universal<br>Interface") --> B("All paths<br>converge to<br>DBI")
        B --> C("Consistent API<br>regardless of<br>driver")
    end
    
    subgraph PyPhil["Python Philosophy"]
        D("DB-API 2.0 as<br>Standard") --> E("Multiple<br>competing<br>approaches")
        E --> F("Choose best<br>tool for job")
    end
    style RPhil fill:#fbf7ec,stroke:#5B8C5A,color:#1B2A41
    style PyPhil fill:#fbf7ec,stroke:#5B8C5A,color:#1B2A41
    style A fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style B fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style C fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    style D fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style E fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style F fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    

    
```

### APIs

> ***2. Draw a mental map for using an API from R and Python. You should (at a minimum) include nodes for `{requests}`/`{httr2}`, request, HTTP verb/request method, headers, query parameters, body, JSON, response, and response code.***

In general, APIs are like the digital 'plumbing' that makes connections between modern software ecosystems possible. 

```{mermaid}
%%| fig-cap: 'APIs'
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%

graph LR
    DS("Data Scientist") --> Req("API Request")
    Req --> ApiServ("API Server")
    ApiServ --> DataProcMod("Data Processing/Model")
    DataProcMod --> Res("Results")
    Res --> ApiServ
    ApiServ --> APIResp("API Response")
    APIResp --> DS
    style DS fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style Req fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style ApiServ fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    style DataProcMod fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style Res fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    style APIResp fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    
    
```

APIs have standardized language that allows different software systems to communicate and share functionality (without needing to understand how the internal implementation works).

**What do they communicate?**

1. What you can request (endpoints)   
2. How to make the request (HTTP methods, parameters)   
3. What you'll get back (response format)   
4. When things go wrong (error codes)  

#### Python APIs

```{mermaid}
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%

graph LR
	subgraph "Python APIs"
		A2("<strong>Python Code</strong>") --> B2("<strong>requests</strong> package")
		B2 --> C2("<strong>requests.method</strong> call")
		C2 --> D2("<strong>method</strong> parameter<br>GET/POST/PUT/DELETE")
		C2 --> E2("<strong>headers</strong> parameter<br>dict of headers")
		C2 --> F2("<strong>params</strong> parameter<br>URL query parameters")
		C2 --> G2("<strong>json</strong> parameter<br>Request payload")
		D2 --> H2("HTTP Request")
		E2 --> H2
		F2 --> H2
		G2 --> H2
		H2 --> I2("API Server")
		I2 --> J2("HTTP Response")
		J2 --> K2("<strong>response.status_code</strong><br>200, 404, 500, etc.")
		J2 --> L2("<strong>response.headers</strong><br>Response metadata")
		J2 --> M2("<strong>response.json</strong><br>JSON parsing")
		K2 --> N2("<strong>Python Response Object</strong>")
		L2 --> N2
		M2 --> N2
	end
    style A2 fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style B2 fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style C2 fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style D2 fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style E2 fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style F2 fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style G2 fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style H2 fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style I2 fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    style J2 fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    style K2 fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style L2 fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style M2 fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style N2 fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff

    
```

#### R APIs

```{mermaid}
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%

graph LR
    subgraph RApi["R APIs"]
        A1("R Code") --> B1("<strong>httr2</strong><br>package")
        B1 --> C1("<strong>request()</strong> object")
        C1 --> D1("<strong>req_method()</strong><br>GET/POST/PUT/DELETE")
        C1 --> E1("<strong>req_headers()</strong><br>Content-Type, Auth, etc.")
        C1 --> F1("<strong>req_url_query()</strong><br>?param=value&key=data")
        C1 --> G1("<strong>req_body_json()</strong><br>Request payload")
        D1 --> H1("HTTP Request")
        E1 --> H1
        F1 --> H1
        G1 --> H1
        H1 --> I1("API Server")
        I1 --> J1("HTTP Response")
        J1 --> K1("<strong>resp_status()</strong><br>200, 404, 500, etc.")
        J1 --> L1("<strong>resp_headers()</strong><br>Response metadata")
        J1 --> M1("<strong>resp_body_json()</strong><br>JSON parsing")
        K1 --> N1("R Response Object")
        L1 --> N1
        M1 --> N1
    end
    style RApi fill:#fbf7ec,stroke:#5B8C5A,color:#1B2A41
    style A1 fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style B1 fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style C1 fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style D1 fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style E1 fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style F1 fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style G1 fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style H1 fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style I1 fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    style J1 fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    style K1 fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style L1 fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style M1 fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style N1 fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff

```

> ***3. How can environment variables be used to keep secrets secure in your code?***




### Previously, in lab 2

We previously turned our Python `scikit-learn` model into a `VetiverModel()`, saved it to a `pins` board, and then created a `VetiverAPI()`.[^lab02-with-py] 

  - In R, we created a linear model (with `lm()` and `fastDummies`), converted it to a `vetiver_model()`, put it in a `board_folder()`, 'pinned' it with `vetiver_pin_write()`, then converted it into an API with `plumber::pr()` and `vetiver_api()`.[^lab02-with-r] 

[^lab02-with-r]: This can be found in [`_labs/lab02/R/model-vetiver-r.qmd`](https://github.com/mjfrigaard/do4ds-labs/tree/main/_labs/lab02/R/model-vetiver-r.qmd). 

[^lab02-with-py]: This can be found in
[/_labs/lab02/model-vetiver/model-vetiver.qmd](https://github.com/mjfrigaard/do4ds-labs/blob/main/_labs/lab02/model-vetiver/model-vetiver.qmd)

The sections below cover how to create and communicate with APIs.  I've split each lab into separate `Python/` and `R/` folders (and included RStudio project files). 

## <strong>Python Shiny App</strong> 

The original Python App files can be found in the [`_labs/lab03/` folder](https://github.com/mjfrigaard/do4ds-labs/tree/main/_labs/lab03/). 

```{bash}
#| code-fold: false
#| eval: false
_labs/lab03/
├── app-api.py
├── model-db.qmd
└──  py-data-load.py
```

### Model EDA {#sec-model-db}

The [model-db.qmd](_labs/lab03/model-db.qmd) file contains the EDA from [lab 1](lab01.qmd).

```{python}
#| eval: false
#| code-fold: false
import duckdb
from pandas import get_dummies
import numpy as np
from sklearn.linear_model import LinearRegression
from sklearn import preprocessing
```

```{python}
#| eval: false
#| code-fold: false
con = duckdb.connect('my-db.duckdb')
df = con.execute("SELECT * FROM penguins").fetchdf().dropna()
con.close()

df.head(3)
```

```{python}
#| eval: false
#| code-fold: false
X = get_dummies(df[['bill_length_mm', 'species', 'sex']], drop_first = True)
y = df['body_mass_g']

model = LinearRegression().fit(X, y)
```

```{python}
#| eval: false
#| code-fold: false
print(f"R^2 {model.score(X,y)}")
print(f"Intercept {model.intercept_}")
print(f"Columns {X.columns}")
print(f"Coefficients {model.coef_}")
```

### Original app file {#sec-app-api-py}

`app-api.py` creates an Python Shiny application that calls the `vetiver` API.

#### UI 

```{python}
#| eval: false
#| code-fold: false
from shiny import App, render, ui, reactive
import requests

api_url = 'http://127.0.0.1:8080/predict'

app_ui = ui.page_fluid(
    ui.panel_title("Penguin Mass Predictor"), 
    ui.layout_sidebar(
        ui.panel_sidebar(
            [ui.input_slider("bill_length", "Bill Length (mm)", 30, 60, 45, step = 0.1),
            ui.input_select("sex", "Sex", ["Male", "Female"]),
            ui.input_select("species", "Species", ["Adelie", "Chinstrap", "Gentoo"]),
            ui.input_action_button("predict", "Predict")]
        ),
        ui.panel_main(
            ui.h2("Penguin Parameters"),
            ui.output_text_verbatim("vals_out"),
            ui.h2("Predicted Penguin Mass (g)"), 
            ui.output_text("pred_out")
        )
    )   
)
```


#### Server

```{python}
#| eval: false
#| code-fold: false
def server(input, output, session):
    @reactive.Calc
    def vals():
        d = {
            "bill_length_mm" : input.bill_length(),
            "sex_Male" : input.sex() == "Male",
            "species_Gentoo" : input.species() == "Gentoo", 
            "species_Chinstrap" : input.species() == "Chinstrap"

        }
        return d
    
    @reactive.Calc
    @reactive.event(input.predict)
    def pred():
        r = requests.post(api_url, json = vals())
        return r.json().get('predict')[0]

    @output
    @render.text
    def vals_out():
        return f"{vals()}"

    @output
    @render.text
    def pred_out():
        return f"{round(pred())}"
```

```{python}
#| eval: false
#| code-fold: false
app = App(app_ui, server)
```


### Data load {#sec-py-data-load-py}

```{python}
#| eval: false
#| code-fold: false
import duckdb
from palmerpenguins import penguins

con = duckdb.connect('my-db.duckdb')
df = penguins.load_penguins()
con.execute('CREATE TABLE penguins AS SELECT * FROM df')
con.close()
```


## <strong>R Shiny App</strong>

The R App API files can be found in the [`_labs/lab03/` folder](https://github.com/mjfrigaard/do4ds-labs/tree/main/_labs/lab03/). 

```{bash}
#| collapse: true
#| eval: false
_labs/lab03/
├── app-api.R
├── eda-db.qmd
└── r-data-load.R
```

### Model EDA {#sec-eda-db.qmd}

The [`eda-db.qmd`](_labs/lab03/eda-db.qmd) file has the EDA R code from lab 1. 

It begins by loading the libraries: 

```{r}
#| eval: false
library(DBI) # add this
library(duckdb) # add this
library(palmerpenguins) # add this
library(dplyr)
library(ggplot2)
library(dbplyr)
```

The code below creates a connection to the database, but needs to be changed slightly so the `penguins` data is registered with `duckdb` (before the queries can be run).

```{r}
#| eval: false
# load penguins data first
penguins_data <- palmerpenguins::penguins
# then connect
con <- DBI::dbConnect(
  duckdb::duckdb(), 
  dbdir = "my-db.duckdb"
)
# register table
duckdb_register(con, "penguins_data", penguins_data)
# create table
dbExecute(con, "CREATE OR REPLACE TABLE penguins AS SELECT * FROM penguins_data")
# create df 
df <- dbGetQuery(con, "SELECT * FROM penguins") |> na.omit()
```

The table below displays the means of the measurements (body mass, flipper length, and bill length/depth) by species and sex:

```{r}
#| eval: false
df %>%
  group_by(species, sex) %>%
  summarise(
    across(
        ends_with("mm") | ends_with("g"),
      \(x) mean(x, na.rm = TRUE)
      )
    ) %>%
  dplyr::collect() %>%
  knitr::kable()
```

The graph displays the best-fit linear relationship between `bill_length_mm` and `body_mass_g` by `species`:

```{r}
#| eval: false
df %>%
  ggplot(
    aes(
    x = bill_length_mm, 
    y = body_mass_g, 
    color = species)
    ) +
  geom_point() + 
  geom_smooth(method = "lm")
```

And we finish by closing the connection to the database.

```{r}
#| eval: false
DBI::dbDisconnect(con)
```

### Original app file

`app-api.R` creates an R Shiny application that calls the `vetiver` API. The top of the file creates an `api_url`:

```{r}
#| eval: false
api_url <- "http://127.0.0.1:8080/predict"
```

#### UI

The UI definition is below: 

```{r}
#| eval: false
ui <- fluidPage(
  titlePanel("Penguin Mass Predictor"),

  # Model input values
  sidebarLayout(
    sidebarPanel(
      sliderInput(
        "bill_length",
        "Bill Length (mm)",
        min = 30,
        max = 60,
        value = 45,
        step = 0.1
      ),
      selectInput(
        "sex",
        "Sex",
        c("Male", "Female")
      ),
      selectInput(
        "species",
        "Species",
        c("Adelie", "Chinstrap", "Gentoo")
      ),
      # Get model predictions
      actionButton(
        "predict",
        "Predict"
      )
    ),

    mainPanel(
      h2("Penguin Parameters"),
      verbatimTextOutput("vals"),
      h2("Predicted Penguin Mass (g)"),
      textOutput("pred")
    )
  )
)
```

#### Server

The server definition is below: 

```{r}
#| eval: false
server <- function(input, output) {
  # Input params
  vals <- reactive(
    list(
      bill_length_mm = input$bill_length,
      species_Chinstrap = input$species == "Chinstrap",
      species_Gentoo = input$species == "Gentoo",
      sex_male = input$sex == "Male"
    )
  )

  # Fetch prediction from API
  pred <- eventReactive(
    input$predict,
    httr2::request(api_url) |>
      httr2::req_body_json(vals()) |>
      httr2::req_perform() |>
      httr2::resp_body_json(),
    ignoreInit = TRUE
  )

  # Render to UI
  output$pred <- renderText(pred()$predict[[1]])
  output$vals <- renderPrint(vals())
}
```

```{r}
#| eval: false
shinyApp(ui = ui, server = server)
```

### Data load {#sec-r-data-load-r}

```{r}
#| eval: false
con <- DBI::dbConnect(duckdb::duckdb(), dbdir = "my-db.duckdb")

DBI::dbWriteTable(con, "penguins", palmerpenguins::penguins)

DBI::dbDisconnect(con)
```

