# <strong>Lab 5: Deployments and Code Promotion</strong> {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: false
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

## Comprehension questions

The questions below come from the [Deployments and Code Promotion](https://do4ds.com/chapters/sec1/1-5-deployments.html) chapter. 

### Data Science Environments 

> ***Write down a mental map of the relationship between the three environments for data science. Include the following terms: Git Promote, CI/CD, automation, deployment, dev, test, prod.***

The three environments for data science are: development, test, and production (not to be confused with the three *layers* of data science environments[^layers]).

The figures from the [chapter](https://do4ds.com/chapters/sec1/1-5-deployments.html#separate-the-prod-environment) do an excellent job illustrating these environments, but I've added a mindmap:

[^layers]: See [Environments Have Layers](https://do4ds.com/chapters/sec1/1-1-env-as-code.html#environments-have-layers) in the do4ds book.

```{=html}
<style>

.codeStyle span:not(.nodeLabel) {
  font-family: monospace;
  font-size: 1.5em;
  font-weight: bold;
  color: #d80808 !important;
  background-color: #ffffff;
  padding: 0.2em;
}
</style>
```

```{mermaid}
%%| fig-cap: 'Data Science Environments'
%%| fig-width: 7.0
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%
mindmap
  root((Code<br>Promotion))
    Dev
      Code Writing
      Experimentation
      Git Promote
        Push to feature<br>branch
        Pull request<br>triggers CI/CD
    Test
      Automation
        Unit tests
        Integration<br>tests
        UAT
      CI/CD
        Validates<br>build
        Runs test<br>suite
    Prod
      Deployment
        Automated<br>release
        Version<br>tagged
      Monitoring
        Stability<br>checks
        Access control

```

#### Development Environments 

The development environment (**dev**) is the primary location where data scientists write new code. It's typically used for iteration and experimentation, and it's connected to version control (Git).  

#### Test Environments 

The testing environment (**test**) is a mirror of the production environment with realistic data (often anonymized). Test environments are designed to check integration and validation with other systems. The test environment also typically includes user acceptance testing and/or system tests.  

#### Production Environments 

The 'live' environment that serves the end users is the production environment (**prod**). The production branch has strictly controlled access with a focus on stability and monitoring.  


### Git & Data Science 

> ***Why is Git so important to a good code promotion strategy?***

Code promotion is the process of tagging Git commits as ‘ready for production.’ This creates a reference point that can trigger tests (in the `test` branch) or a production deployment pipeline (in the `prod` branch). Without version control, untracked changes have no canonical record. Git and code promotion together ensure consistency across all three environments (dev, test, and prod) and reduce human error in deployment.

#### Artifact immutability 

With Git, all changes are signed via specific commit SHAs or tagged releases. Before any changes can be merged into a `prod` branch, approval is required (the branch is protected). This **artifact immutability** connects builds as they are promoted across environments, ensuring untested or unreviewed code never reaches production.

#### Environment drift

Environment drift is when the production code differs from the tested code. Without a code promotion strategy and version control, the code would need to be rebuilt in each environment, so the tested code wouldn't necessarily be the code deployed into production. In DevOps, environment drift is the enemy because it can lead to unpredictable behavior and makes debugging difficult. 

#### Environment parity

Git enables environment parity. When used properly, our three environments (`dev`, `test`, and `prod`) differ only by data and credentials. Changes are promoted, not re-created. When production code breaks, developers can use Git to roll back to a previous version (via commit SHAs or tagged releases), avoiding ad hoc changes directly in `prod`.


> ***Can you have a code promotion strategy without Git?***

Code promotion can be done without Git, but you would need to find a way to:

* Create an 'immutable versioned artifact' (some way of tracking who did what, and when)
* Set up blocks and gates requiring tests and approvals before code is promoted
* Build a promotion mechanism to copy, move, or retag each versioned artifact between environments

You would also need substitutes for:

1. A way to view and review diffs, either through code review or document-style review
2. The ability to keep previous artifacts and switch between environments
3. Some means of traceability connecting artifacts, manifests, approvals, and build logs


### Git & CI/CD

> ***What is the relationship between Git and CI/CD?***

A CI/CD pipeline is an automated process that builds, tests, and deploys code. In CI/CD systems, Git events define the code promotion workflow, not the pipeline itself. The pipeline simply reacts to Git events, while the promotion logic lives in the branches, tags, and merge policies. This enables a "build once, promote many times" strategy where the same tested artifact moves across environments.

Git provides the code change and approval history, with pipeline results tied to specific commits. Each release is tagged, making it straightforward to track what changed and when. That traceability is critical for debugging and accountability. Git serves as the gatekeeper for CI/CD policies and integrates directly with enforcement mechanisms.

Using Git turns CI/CD into a system, not a script.

```{mermaid}
%%| fig-cap: 'Git & CI/CD'
%%| fig-width: 7.0
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%
flowchart LR
    GitEv("Git Event<br>(push, PR, merge)") -->|"triggers"| CICD("CI/CD Pipeline")
    CICD --> CI("Continuous<br>Integration (CI)<br><em>Build & Test</em>")
    CICD --> CD("Continuous<br>Delivery (CD)<br><em>Deploy</em>")
    style GitEv fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style CICD fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style CI fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    style CD fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff

```

Think of Git as the source of truth and CI/CD as an automated response to changes in that truth. Git events (pushes, pull requests, merges) can fire a CI/CD pipeline.


**Key distinctions between Git and CI/CD:**

| | Git | CI/CD |
|---|---|---|
| **Role** | Tracks & versions code | Automates build, test, deploy |
| **Trigger** | Human action (commit, merge) | Git events |
| **Output** | A versioned history | A tested, deployed artifact |




> ***What’s the benefit of using Git and CI/CD together?***

Without Git, CI/CD has no reliable source to build from. Without CI/CD, Git changes require manual promotion between environments, which is slow and error-prone.

```{mermaid}
%%| fig-cap: 'Git & CI/CD'
%%| fig-width: 7.0
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%

gitGraph
   commit id: "Initial codebase"

   branch dev
   checkout dev
   commit id: "Experiment: model v1"
   commit id: "Iterate: model v2"
   commit id: "Iterate: model v3"

   branch test
   checkout test
   commit id: "Deploy to Test env"
   commit id: "Integration checks"
   commit id: "System tests"
   commit id: "UAT approved"

   checkout main
   merge test id: "Merge to main"
   commit id: "Tag release v1.0"

   branch prod
   checkout prod
   commit id: "Deploy to Prod"
   commit id: "Monitoring active"
```

In the `gitGraph` above: 

- **`dev`** is where all experimentation happens and never touches `main` directly
- **`test`** branches from `dev` once code is stable enough to validate, keeping conditions isolated from `prod`
- **`main`** only receives code that has passed full UAT and system testing, acting as the **gatekeeper**
- **`prod`** branches from `main` after a release is tagged, representing exactly what is live at any given time and making rollbacks straightforward

### Git, GitHub, CI/CD, Actions, & Version Control 

> ***Write out a mental map of the relationship of the following terms: Git, GitHub, CI/CD, GitHub Actions, Version Control.***


```{mermaid}
%%| fig-cap: 'Git, GitHub, CI/CD, Actions, & Version Control'
%%| fig-width: 7.0
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%

mindmap
  root((Code<br>Management))
    <strong>Version Control</strong>
      <em>Tracks changes<br>over time</em>
      <em>Enables<br>rollback</em>
      <em>Supports<br>collaboration</em>
        <strong>Git</strong>
          <em>Local<br>repositories</em>
          <em>Branching<br>and merging</em>
          <em>Commit history</em>
            <strong>GitHub</strong>
              <em>Remote<br>repositories</em>
              <em>Pull<br>requests</em>
              <em>Code<br>review</em>
                <strong>GitHub Actions</strong>
                  <em>Workflow<br>files</em>
                  <em>Triggered by<br>Git events</em>
                    <strong>CICD</strong>
                      <em>Continuous<br>Integration</em>
                        <em>Automated<br>testing</em>
                        <em>Build<br>validation</em>
                      <em>Continuous<br>Delivery</em>
                        <em>Automated<br>deployment</em>
                        <em>Environment<br>promotion</em>

```



