# <strong>Lab 7: Getting started with AWS</strong> {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

## Comprehension questions


> *What are two reasons you should consider going to the cloud?*

Two primary reasons to move to the cloud include:

* The ability to rapidly increase/decrease computing resources to match demand without investing in physical hardware (i.e., scalability)    
* The cloud provider handles much of the underlying infrastructure management, such as hardware repairs, power, and cooling (i.e., reduce maintenance).  


> *What’s one reason you shouldn’t?*

One reason we might avoid the cloud is *cost unpredictability*. While we only pay for what we use, unmonitored usage or unexpected spikes in demand can lead to much higher monthly bills than anticipated.


> *What is the difference between PaaS, IaaS, and SaaS?* 

The differences lie in how much of the technology stack we manage versus how much the provider manages:

```{mermaid}
%%| fig-cap: ‘Cloud acronyms’
%%| fig-width: 7.0
%%| fig-align: 'center'
%%{init: {‘theme’: ‘base’, ‘themeVariables’: {‘fontFamily’: ‘monospace’}}}%%

graph LR
    subgraph "User Management Level"
        SaaS["SaaS: You manage<br>only usage"]
        PaaS["PaaS: You manage<br>apps & data"]
        IaaS["IaaS: You manage<br>OS, apps, &<br>data"]
    end
    
    Provider(["Provider manages<br>infrastructure"]) --> SaaS
    Provider --> PaaS
    Provider --> IaaS
    
    style Provider fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style IaaS fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style SaaS fill:#1B2A41,stroke:#000000,stroke-width:1px,color:#ffffff
    style PaaS fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    
```

The table below covers other differences: 

| Acronym | Model                       | Responsibility                        |
| ------- | :-------------------------- | :------------------------------------ |
| IaaS    | Infrastructure as a Service | User manages OS, middleware, and apps |
| PaaS    | Platform as a Service       | User manages applications and data    |
| SaaS    | Software as a Service       | Provider manages everything           |

> *What’s an example of each that you’re familiar with?*

| Acronym | Description                                                                           | Example                     |
| :------ | ------------------------------------------------------------------------------------- | :-------------------------- |
| IaaS    | Provides raw computing resources like virtual servers and networking.                 | AWS EC2                     |
| PaaS    | Provides a framework for developers to build and deploy apps without managing the OS. | Heroku or Google App Engine |
| SaaS    | Delivers finished software applications over the internet.                            | Google Workspace or Slack   |


> *What are the names of AWS’s services for: renting a server, filesystem storage, blob storage?*

*   **Renting a server:** Amazon EC2 (Elastic Compute Cloud)
*   **Filesystem storage:** Amazon EFS (Elastic File System)
*   **Blob storage:** Amazon S3 (Simple Storage Service)

