# <strong>Lab 9: Linux Admin</strong> {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: false
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

## Comprehension questions

> *What are the parts of a bash command?*

A bash command has several key parts:

```{mermaid}
%%| fig-cap: 'Bash Command Structure'
%%| fig-width: 8.0
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}}%%

graph TD
    A["command<br/><code>ls</code>, <code>cd</code>, <code>echo</code>"] --> B["options/flags<br/><code>-l</code>, <code>-a</code>, <code>-r</code>"]
    B --> C["arguments<br/><code>filenames</code>, <code>paths</code>"]
    C --> D["redirects<br/><code>&gt;</code>, <code>&lt;</code>, <code>|</code>"]

    style A fill:#d2562b,stroke:#fff,stroke-width:2px,color:#fff
    style B fill:#2a6f77,stroke:#fff,stroke-width:2px,color:#fff
    style C fill:#5b8c5a,stroke:#fff,stroke-width:2px,color:#fff
    style D fill:#e8a33d,stroke:#000,stroke-width:2px,color:#000
```

|  |  |  |
|------------------------|------------------------|------------------------|
| Part | Description | Example |
| Command | The program to run | `ls`, `cd`, `echo`, `grep` |
| Options/Flags | Modify behavior, prefixed with (single) or (double) | `-` or `--` |
| Arguments | What the command operates on | files, paths, text |
| Redirects | Send output elsewhere | `>` file, `<` input, `|` pipe |

**Example:** `ls -la /home | grep .txt`

Command: `ls` \| Flags: `-la` \| Argument: `/home` \| Pipe: `|` \| Chain: `grep .txt`

> *What is the difference between a relative and an absolute path?*

The main difference is that absolute paths always start with `/` (and works from anywhere), while relative paths do not start with `/` and are interpreted from the current working directory (you can use `..` to go up one level, `.` for current directory).

```{mermaid}
%%| fig-cap: 'Absolute vs Relative Paths'
%%| fig-width: 8.0
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}}%%

graph LR
    Path["Filesystem<br>Paths"]

    Absolute["<strong>Absolute Path</strong><br/>Starts from root<br>directory /"]
    AbEx1["<code>/home/user/documents/file.txt</code>"]
    AbEx2["<code>/etc/passwd</code>"]
    AbDescr["Always same location<br>regardless of current<br>directory"]

    Relative["<strong>Relative Path</strong><br/>Relative to current<br>directory"]
    RelEx1["<code>documents/file.txt</code>"]
    RelEx2["<code>../parent-folder/file.txt</code>"]
    RelDescr["Changes meaning based<br>on where you are<br>in filesystem"]

    Path -->|"Full from root"| Absolute
    Absolute --> AbEx1
    Absolute --> AbEx2
    Absolute --> AbDescr

    Path -->|"From current location"| Relative
    Relative --> RelEx1
    Relative --> RelEx2
    Relative --> RelDescr

    style Path fill:#e8a33d,stroke:#000,stroke-width:2px,color:#000
    style Absolute fill:#d2562b,stroke:#fff,stroke-width:2px,color:#fff
    style Relative fill:#2a6f77,stroke:#fff,stroke-width:2px,color:#fff
    
```

> *What are some ways to direct them to run in a particular place on the filesystem?*

**You can run use relative/absolute paths to:**

1.  Change to a directory, then run the commands: `cd /path/to/directory`
2.  Run the script with the absolute path directly: `/absolute/path/to/file.sh`
3.  Run the script relative to current location: `relative/path/file.sh`
4.  Run the script in the current directory (explicit): `./file.sh`

```{mermaid}
%%| fig-cap: 'Ways to Control Execution Location'
%%| fig-width: 8.0
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}}%%

graph LR
    Control["Ways to Control<br>Execution Location"]

    CD["Use <code>cd</code>to change<br>directory first"]
    ABS["Use the absolute path<br/>(specify full path)"]
    REL["Use the relative path<br/>(but specify from<br>current location)"]
    Dot["Run from current directory with <code>./file.sh</code>"]

    Control --> CD
    Control --> ABS
    Control --> REL
    Control --> Dot

    style Control fill:#d2562b,stroke:#fff,stroke-width:2px,color:#fff
    style CD fill:#2a6f77,stroke:#fff,stroke-width:1px,color:#fff
    style ABS fill:#2a6f77,stroke:#fff,stroke-width:1px,color:#fff
    style REL fill:#2a6f77,stroke:#fff,stroke-width:1px,color:#fff
    style Dot fill:#5b8c5a,stroke:#fff,stroke-width:1px,color:#fff
```

> *How can you copy, move, or delete a file? What about to or from a server?*

**Local operations include `cp`, `mv` and `rm`:**

| Command                            | Description                            |
|----------------------------------|--------------------------------------|
| `cp /source/destination`           | copy a file                            |
| `cp -r source_dir destination_dir` | copy a directory recursively           |
| `mv source destination`            | move or rename a file                  |
| `rm filename`                      | delete a file (permanent, no recovery) |
| `rm -r directory`                  | delete a directory and it's contents   |

**Server operations (via SSH) include `scp`, `rsync`, and `sftp`**

| Command | Description |
|------------------------------------|------------------------------------|
| `scp local_file user@server:/remote/path` | copy to server |
| `scp user@server:/remote/file local_path` | copy from server |
| `scp -r local_dir user@server:/remote/path` | copy directory to server |
| `rsync -av local_dir user@server:/remote/path` | sync files (better for large operations) |
| `sftp user@server` | interactive file transfer session |

**Key note:** `rm` is permanent! There is no trash can in Linux. Use the `-i` flag to prompt before deletion: `rm -i filename`

> *Mind map of: operating system, Windows, MacOS, Unix, Linux, Distro, Ubuntu*

```{mermaid}
%%| fig-cap: 'Operating System Family Tree'
%%| fig-width: 8.0
%%| fig-align: 'center'
%%{init: {'theme': 'neutral', 'themeVariables': {'fontFamily': 'monospace'}}}%%

mindmap
  root((Operating System))
    Commercial
      Windows
        GUI-focused
        Proprietary<br>kernel
      MacOS
        Unix-based
        Proprietary<br>kernel
        Desktop &<br>laptop
    Open Source<br>Unix-like
      Unix
        Original<br>design
        POSIX<br>standard
        BSD, Solaris
      Linux
        Free & open<br>source
        POSIX<br>compatible
        Kernel +<br>utilities
        Distros
          Ubuntu
            Debian-based
            User-friendly
            Server &<br>desktop
          CentOS
            RedHat-based
            Enterprise
          Fedora
            Cutting<br>edge
            RedHat-based
        Distro
          Linux kernel +<br>tools
          Different package<br>managers
          Different<br>defaults
          Community<br>focused
```

**Key relationships between operating systems:**

The **Operating System** isthesoftware that manages hardware and runs programs. **Windows and MacOS** are commercial proprietary operating systems, while Unix is the original open standard from 1970s (influenced modern systems).

**Linux** is a free, open-source Unix-like kernel created by Linus Torvalds, and the **distro (Distribution)** is the Linux kernel packaged with utilities, package manager, defaults. **Ubuntu** is a popular Linux distro based on Debian.

> *What are the 3x3 options for Linux file permissions? How are they indicated in ls -l?*

The 3×3 permission matrix:

| Owner (User)        | Group               | Other               |
|---------------------|---------------------|---------------------|
| **Read (r)** = 4    | **Read (r)** = 4    | **Read (r)** = 4    |
| **Write (w)** = 2   | **Write (w)** = 2   | **Write (w)** = 2   |
| **Execute (x)** = 1 | **Execute (x)** = 1 | **Execute (x)** = 1 |

**How they appear in `ls -l`:**

```         
-rw-r--r-- 1 user group 1234 Jan 15 10:30 file.txt
^         ^^^^^^ ^^^^^^ ^^^^^^
|         |      |      |
|         |      |      +-- Other: read only
|         |      +--------- Group: read only
|         +-------------- User: read & write
+----------------------- File type (- = regular file, d = directory)
```

**Breaking down permissions:**

- **Position 1:** File type (`-` file, `d` directory, `l` link)
- **Positions 2-4:** Owner permissions (rwx)
- **Positions 5-7:** Group permissions (rwx)
- **Positions 8-10:** Other permissions (rwx)

**Octal notation (chmod):**

- `chmod 755 file`: Owner: rwx (7), Group: rx (5), Other: rx (5)
- `chmod 644 file`: Owner: rw (6), Group: r (4), Other: r (4)
- `chmod 700 file`: Owner: rwx (7), Group: (0), Other: (0)
- `chmod 600 file`: Owner: rw (6), Group: (0), Other: (0): typical for SSH keys
