# <strong>Lab 8: The Command Line</strong> {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: false
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

## Comprehension questions

> "*Draw a mental map that includes the following: terminal, shell, theme manager, operating system, my laptop.*"

```{mermaid}
%%| fig-cap: 'System Architecture: Layers and Components'
%%| fig-width: 7.0
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%

graph TD
    Laptop(["<strong>Laptop</strong><br/><em>(Hardware)</em>"])
    OS("<strong>Operating System</strong><br/><em>(macOS, Linux,<br>Windows)</em>")
    Shell("<strong>Shell</strong><br/><em>(bash, zsh, fish)</em>")
    Terminal{{"<strong>Terminal</strong><br/><em>(Application)</em>"}}
    Theme("<strong>Theme Manager</strong><br/><em>(Appearance settings)</em>")

    Laptop --> OS
    OS --> Shell
    OS --> Theme
    Shell --> Terminal
    Theme --> Terminal

    style Laptop fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style OS fill:#1B2A41,stroke:#000000,stroke-width:1px,color:#ffffff
    style Shell fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    style Terminal fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style Theme fill:#E8A33D,stroke:#000000,stroke-width:1px,color:#ffffff
    
```

> "*Under what circumstances should you move or share your SSH private key? (Hint: this is a trick question.)*"

**Never.** This is the trick–don't ever move or share an SSH private key under any circumstances. A private key is like the password to our entire identity on remote systems. If someone gains access to it, they can impersonate us, access our servers, and compromise our systems.

The only place a private key should exist is on machines we own and control. If we suspect a private key has been compromised, generate a new key pair immediately and update the corresponding public key on all remote systems.

> "*What is it about SSH public keys that makes them safe to share?*"

SSH public keys are mathematically safe to share because of the asymmetric encryption algorithm they're based on (typically RSA or ECDSA). The public key can only encrypt messages that the private key can decrypt—it cannot be reversed to derive the private key. The mathematical relationship is one-way: given a public key, it's computationally infeasible to calculate the corresponding private key. This asymmetry is what makes public-key cryptography work. We can safely distribute our public key anywhere (GitHub, servers, etc.) because knowing the public key provides no information about the private key.

## SSH Key Authentication Flow

Here's how SSH keys work in practice when we authenticate to a remote server. The server never sees our private key. It only uses our public key to encrypt a challenge that only our private key can decrypt. This proves we own the private key without ever transmitting it:

```{mermaid}
%%| fig-cap: 'SSH Key Authentication Sequence'
%%| fig-width: 8.0
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}}%%

sequenceDiagram
    actor User as Developer
    participant Client as Client Machine
    participant Server as Remote Server

    User->>Client: ssh user@server.com
    Client->>Server: Initiate connection
    Server->>Client: Send server public key
    Client->>Client: Verify server identity
    Client->>Server: Offer client public key
    Server->>Server: Look up authorized_keys
    alt Public key found
        Server->>Client: Challenge (encrypt with public key)
        Client->>Client: Decrypt with private key
        Client->>Server: Send response
        Server->>Server: Verify response
        Server->>Client: ✓ Authentication successful
        Client->>User: Connected
    else Public key not found
        Server->>Client: ✗ Authentication failed
        Client->>User: Permission denied
    end
```
