---
editor: 
  markdown: 
    wrap: 72
---

# Lab 4: Logging and Monitoring {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

The original files in `do4ds/_labs/lab04` are below:

```{r}
#| collapse: true
#| eval: false
fs::dir_tree(path = "_labs/lab04/")
## _labs/lab04/
## ├── app-log.R #<1>
## └── app-log.py #<2>
```
1.  R Shiny app   
2.  Python Shiny app

## Comprehension questions

Below are my answers to the comprehension questions in this chapter. 

### Monitoring vs. logging

> ***What is the difference between monitoring and logging? What are the two halves of the monitoring and logging process?***

Logging is a record of "***What happened***"--logs capture discrete events and detailed narratives of what occurred in the system at specific points in time.

Monitoring is usually a display of "***How is it performing***"--tracking quantitative metrics and system health indicators over time to understand trends and current state.

```{mermaid}
%%| fig-cap: 'Logging vs. Monitoring'
%%| fig-align: center
%%| fig-width: 6.5
%%{init: {'theme': 'base', 'look': 'handDrawn', 'themeVariables': {'fontFamily': 'monospace'}}}%%

graph LR
    subgraph "LOG: events"
        L1("Discrete<br>Events")
        L2("Detailed<br>Context")
        L3("Historical<br>Investigation")
        L4("Text-Based<br>Messages")
        L5("<strong>When/What/Who/Why</strong>")
    end
    
    subgraph "MONITOR: metrics"
        M1("Quantitative<br>Metrics")
        M2("Real-time<br>Status")
        M3("Trend<br>Analysis")
        M4("Numerical<br>Data")
        M5("<strong>How Much/How Fast/<br>How Often</strong>")
    end
    
    subgraph "Common Use Cases"
        LC("Root Cause<br>Analysis")
        LC2("Debugging Specific<br>Issues")
        LC3("Compliance<br>Auditing")
        LC4("Security<br>Investigation")
        
        MC("Performance<br>Optimization")
        MC2("Capacity<br>Planning")
        MC3("SLA Monitoring")
        MC4("Real-time<br>Alerting")
    end
    
    L1 --> LC
    L2 --> LC2
    L3 --> LC3
    L4 --> LC4
    
    M1 --> MC
    M2 --> MC2
    M3 --> MC3
    M4 --> MC4
    
    classDef logging fill:#d8e4ff,stroke:#333,stroke-width:2px,rx:6,ry:6,text-align:center,font-size:13px
    classDef monitoring fill:#d8e4ff,stroke:#333,stroke-width:2px,rx:6,ry:6,text-align:center,font-size:13px
    classDef usecase fill:#31e981,stroke:#333,stroke-width:1px,rx:6,ry:6,text-align:center,font-size:13px
    
    class L1,L2,L3,L4 logging
    class M1,M2,M3,M4 monitoring
    class LC,LC2,LC3,LC4,MC,MC2,MC3,MC4 usecase
    classDef default fill:#F2E6C2,stroke:#2A6F77,color:#1B2A41
```

| Aspect | Logging | Monitoring |
|-----------------|----------------------------|----------------------------|
| **Data Type** | Discrete events, text messages | Metrics, counters, gauges |
| **When Used** | After something happens | Continuously |
| **Purpose** | Debugging, investigation | Performance tracking, alerting |
| **Format** | Structured/unstructured text | Numerical time-series data |
| **Storage** | Log files, log databases | Time-series databases |
| **Analysis** | Search, filter, correlate | Aggregate, trend, threshold |

### What not to log

> ***Logging is generally good, but what are some things you should be careful not to log?***

Critical data we should make sure not to log includes:

1.  Sensitive Personal Information (PII)    
    -   Email addresses, phone numbers, names     
    -   Social Security Numbers, passport numbers   
    -   Addresses, birth dates    
    -   Medical information (HIPAA protected)   
2.  Authentication Credentials    
    -   Passwords (plain text or hashed)   
    -   API keys and tokens     
    -   Database connection strings with credentials    
    -   OAuth tokens and refresh tokens   
3.  Financial Information   
    -   Credit card numbers     
    -   Bank account numbers    
    -   Routing numbers   

Industry specific guidelines on logging are available below:

1.  Financial Services:   
    -   Never log account numbers in full   
    -   Mask payment card data    
    -   Avoid logging transaction details in plain text   
        -   Source: [FFIEC IT Examination Handbook](https://ithandbook.ffiec.gov/it-booklets/information-security.aspx)   
2.  Healthcare    
    -   De-identify all patient data in logs    
    -   Remove [18 HIPAA identifiers](https://www.luc.edu/its/aboutus/itspoliciesguidelines/hipaainformation/the18hipaaidentifiers/)      
    -   Implement access controls for log files   
        -   Source: [HHS Security Risk Assessment](https://www.hhs.gov/hipaa/for-professionals/security/guidance/cybersecurity/index.html)
3.  Government/Defense    
    -   Classify log data appropriately    
    -   Implement need-to-know access   
    -   Encrypt logs containing sensitive information     
        -   Source: [NIST SP800-53](https://csrc.nist.gov/publications/detail/sp/800-53/rev-5/final)

### Log levels

> ***At what level would you log each of the following events:***

The answers below are provided for the [R `logger` package](https://daroczig.github.io/logger/) and [Python's `logging` facility](https://docs.python.org/3/library/logging.html#). 

In R, the `logger` package has the following levels:

```{mermaid}
%%| fig-cap: 'R logger'
%%| fig-align: center
%%| fig-width: 6.5
%%{init: {'theme': 'base', 'look': 'handDrawn', 'themeVariables': {'fontFamily': 'monospace'}}}%%

graph TD
    LogLevels("Logging Levels<br>Least -> Most Severe") --> TRACE("TRACE")
    TRACE --> DEBUG("DEBUG")
    DEBUG --> INFO("INFO")
    INFO --> SUCCESS("SUCCESS")
    SUCCESS --> WARN("WARN")
    WARN --> ERROR("ERROR")
    ERROR --> FATAL("FATAL")
    
    TRACE -.Fine-grained details.-> TraceEx[Variable states<br>Loop iterations]
    DEBUG -.Diagnostic information.-> DebugEx[Function entry/exit<br>Parameter values]
    INFO -.Informational messages.-> InfoEx[Process started<br>Configuration loaded]
    SUCCESS -.Positive confirmations.-> SuccessEx[Task completed<br>Data saved successfully]
    WARN -.Potential issues.-> WarnEx[Deprecated functions<br>Missing optional params]
    ERROR -.Recoverable errors.-> ErrorEx[Failed validation<br>File not found]
    FATAL -.Critical failures.-> FatalEx[Cannot proceed<br>System unrecoverable]
    classDef default fill:#F2E6C2,stroke:#2A6F77,color:#1B2A41
    

```

In Python, the `logging` facility has the following log levels:

```{mermaid}
%%| fig-cap: 'Python logging'
%%| fig-align: center
%%| fig-width: 6.5
%%{init: {'theme': 'base', 'look': 'handDrawn', 'themeVariables': {'fontFamily': 'monospace'}}}%%

graph TD
    A("Logging Levels<br>Least -> Most Severe") --> B("DEBUG")
    B --> C("INFO")
    C --> D("WARNING")
    D --> E("ERROR")
    E --> F("CRITICAL")
    
    B -.Verbose details.-> B1[Function calls<br>Variable values]
    C -.Normal events.-> C1[Operations succeeded<br>State changes]
    D -.Unexpected but handled.-> D1[Slow responses<br>Retries needed]
    E -.Failures.-> E1[Operations failed<br>Connections refused]
    F -.System failure.-> F1[Cannot continue<br>Data corruption]
    classDef default fill:#F2E6C2,stroke:#2A6F77,color:#1B2A41
    
    
```

> ***Someone clicks on a particular tab in your Shiny app.***

:::{layout="[40,60]" layout-valign="top"}

**R `logger`:** `TRACE` or `DEBUG`

Tab clicks are fine-grained UI interaction details, so we could use `TRACE` if we want extremely detailed user behavior tracking, or `DEBUG` if we only need this during development/troubleshooting. In production, we typically wouldn't log every tab click unless tracking user flow.

:::

:::{layout="[40,60]" layout-valign="top"}

**Python `logging`:** `DEBUG`

Similar reasoning - this diagnostic information for user interactions are primarily useful during development (or when debugging specific user behavior issues). 

:::

*In most production apps, we'd disable `TRACE`/`DEBUG` levels and only enable them when investigating specific issues to avoid log bloat.*

> ***Someone puts an invalid entry into a text entry box.***

:::{layout="[40,60]" layout-valign="top"}

**R `logger`:** `WARN`

The user provided invalid input, but the app can handle it gracefully (show validation message). This is a potential issue worth noting but doesn't break functionality.

:::

:::{layout="[40,60]" layout-valign="top"}

**Python `logging`:** `WARNING`

Same reasoning - unexpected but handled situation. We'd want to track validation failures for UX improvements without treating them as errors.

:::

*Some developers use `INFO` for validation failures if they're expected and very common. Use `WARN` if you want to monitor frequency or patterns of invalid inputs.*

> ***An HTTP call your app makes to an external API fails.***

:::{layout="[40,60]" layout-valign="top"}

**R `logger`:** `ERROR`

The API call failed, which means our app cannot complete an intended operation. Even if we have fallback logic, the primary operation failed and needs attention.

:::

:::{layout="[40,60]" layout-valign="top"}

**Python `logging`:** `ERROR`

Same reasoning - this is a genuine failure that prevents the app from functioning as designed.

:::

*Nuances:*

-   *We should use `ERROR` if this breaks core functionality*     
-   *We might consider `WARN` if we have robust fallback mechanisms and the app continues normally*   
-   *Use `FATAL` only if the API is so critical that the entire app becomes unusable without it*   

*Include the HTTP status code, error message, and endpoint in the log:*

```{r}
#| eval: false 
#| code-fold: false
log_error("API call failed: {endpoint} returned {status_code} - {error_msg}")
```

> ***The numeric values that are going into your computational function.***

:::{layout="[40,60]" layout-valign="top"}

**R `logger`:** `TRACE`

These are extremely granular details about data flowing through our functions, so we'd use `TRACE` for this level of variable inspection. We might use `DEBUG` if these values are critical decision points.

:::

:::{layout="[40,60]" layout-valign="top"}

**Python `logging`:** `DEBUG`

Python doesn't have `TRACE`, so `DEBUG` is appropriate for detailed variable state information used in troubleshooting computational logic.

:::

*When to log this:*  

-   *During development to verify calculations*      
-   *When debugging reported computational issues*     
-   *Should almost never be enabled in production*       


#### Log Levels Summary


| Event          | `logger`        |`logging`         | Key Factor                  |
|----------------|-----------------|------------------|-----------------------------|
| Tab click      | `TRACE`/`DEBUG` | `DEBUG`          | Fine-grained UI interaction |
| Invalid input  | `WARN`          | `WARNING`        | Expected, handled issue     |
| API failure    | `ERROR`         | `ERROR`          | Operation failed            |
| Numeric values | `TRACE`         | `DEBUG`          | Variable state inspection   |

**General Rules of Thumb:**

-   **`TRACE`/`DEBUG`** = "I need to see what's happening inside"

-   **`INFO`/`SUCCESS`** = "Normal operations worth recording"

-   **`WARN`/`WARNING`** = "Something unexpected but manageable"

-   **`ERROR`** = "Something broke but app continues"

-   **`FATAL`/`CRITICAL`** = "Cannot continue operating"

## Observability

We can't prevent code failures (or when they happen), but we *can* create [**observable systems**](https://en.wikipedia.org/wiki/Observability_(software)) that 1) **accept failure is inevitable** (rather than trying to prevent it), 2) **implement comprehensive monitoring and logging**, and 3) **enable rapid diagnosis** (i.e., using logs and metrics to reconstruct what failed and why).

To do this, data scientist and developers need to generate meaningful logs and metrics from their code. System admins need to ensure new data science work can be easily integrated within the organizational infrastructure (i.e., into the existing log aggregation and monitoring tools).

```{mermaid}
%%| fig-width: 8.0
%%| fig-align: center
%%{init: {'theme': 'base', 'look': 'handDrawn', 'themeVariables': {'fontFamily': 'monospace'}}}%%

graph LR
    subgraph DatSci["Data Scientist"]
        A("Generate Logs")
        B("Emit Metrics")
    end
    
    subgraph SysAdmin["System Admins"]
        C("Collect &<br>Aggregate")
        D("Store &<br>Process")
        E("Monitor &<br>Alert")
    end
    
    subgraph Out["Outcomes"]
        F("Detect<br>Issues")
        G("Diagnose<br>Problems")
        H("Rapid<br>Recovery")
    end
    
    A ---> C
    B ---> C
    C --> D
    D --> E
    E --> F & G & H

    
    
    classDef ds fill:#d8e4ff,stroke:#000,stroke-width:1px
    class A,B ds
    
    classDef outcome fill:#31e981,stroke:#000,stroke-width:1px
    class F,G,H outcome
    classDef default fill:#F2E6C2,stroke:#2A6F77,color:#1B2A41
    style DatSci fill:#d4e7e4,stroke:#2A6F77,color:#1B2A41
    style SysAdmin fill:#d4e7e4,stroke:#2A6F77,color:#1B2A41
    style Out fill:#d4e7e4,stroke:#2A6F77,color:#1B2A41
    
    
```

This will give us:

1)  Operational peace of mind (the ability to detect and resolve issues quickly)

2)  Value demonstration (shows decision makers who uses the work and how)

3)  Team validation (prove the team's impact and justify resources)

## Logging in R

<details>

<summary>Expand to view app-log.R</summary>

```{r}
#| label: print-app-log-R
#| echo: false
#| code-overflow: wrap
library(readr)
readr::read_lines("_labs/lab04/app-log.R") |> 
  base::noquote() |> 
  writeLines()
```

</details>

## Logging in Python

<details>

<summary>Expand to view app-log.py</summary>

```{r}
#| label: print-app-log-py
#| echo: false
#| code-overflow: wrap
library(readr)
readr::read_lines("_labs/lab04/app-log.py") |> 
  base::noquote() |> 
  writeLines()
```

</details>
