# Lab 2: Project Architecture {.unnumbered}


```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

Lab 2 introduces the[ `vetiver` package](https://rstudio.github.io/vetiver-r/) for modeling in [`model-vetiver.qmd`](model-vetiver.qmd). The `vetiver` package comes in both flavors: [R](https://vetiver.rstudio.com/) and [Python](https://rstudio.github.io/vetiver-python/). 

```{r}
#| collapse: true
#| eval: false
fs::dir_tree(path = "_labs/lab02/")
## _labs/lab02/
## └── model-vetiver.qmd
```


If you'd like to see the same process using R, check out the [R version](model-vetiver-r.qmd).

## Comprehension Questions

1. What are the layers of a three-layer application architecture? What libraries could you use to implement a three-layer architecture in R or Python?

```{=html}
<style>

.codeStyle span:not(.nodeLabel) {
  font-family: monospace;
  font-size: 1.5em;
  font-weight: bold;
  color: #9753b8 !important;
  background-color: #f6f6f6;
  padding: 0.2em;
}
</style>
```

```{mermaid}
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%
flowchart TD
    Present("Presentation<br>Layer") -.-> Process("Processing<br>Layer")
    Process -.-> Data("Data<br>Layer")
    
    subgraph UI["<strong>User Interface</strong>"]
        Present
    end
    
    subgraph BL["<strong>Business Logic</strong>"]
        Process
    end
    
    subgraph DS["<strong>Data Storage</strong>"]
        Data
    end
    
    %% #b5c0c5 is a tint of #6b818c
    style Present fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style Process fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style Data fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    style UI fill:#fbf7ec,stroke:#5B8C5A,color:#1B2A41
    style BL fill:#fbf7ec,stroke:#2A6F77,color:#1B2A41
    style DS fill:#fbf7ec,stroke:#2A6F77,color:#1B2A41
    
```

*2. What are some questions you should explore to reduce the data requirements for your project?*

- Problem Definition    
  - What is the minimum viable outcome?   
  - Can we solve a simpler version first?   
  - What is the cost of being wrong?    
- Feature Engineering   
  - Which features are actually predictive?   
  - Can we derive features from less data?      
  - What is redundant?    
- Sampling Strategy   
  - How much data do we actually need?    
  - Can we stratify smartly?    
  - What about active learning?   
- External Resources    
  - What data already exists?   
  - Can we use pre-trained models?    
  - Are there proxies available?    

*3. What are some patterns you can use to make big data smaller?*

- Minimize data movement by doing as much work as possible where the data already sits.

- Don't load all your data at once if you don't need to (keep a live connection to your database and only pull specific pieces of data when you actually need them). 

- Working with samples can dramatically speed up your analysis while maintaining accuracy for most types of questions

- If your data has natural groupings, process one group at a time instead of loading everything.

*4. Where can you put intermediate artifacts in a data science project?*

Flat files, `.csv` files, or `duckdb` (which only loads the data you need into memory).

*5. What does it mean to “take data out of the bundle”?*

Take data out of the presentation bundle (i.e., separate it from the code) if it's updated more frequently than the app code. 

## Data preparation {#sec-pkgs-ilbs}

We'll use the `reticulate` package below to install the other packages in the `model-vetiver.qmd` lab. 

```{r}
#| results: hide
#| message: false
#| warning: false
library(reticulate)
if (!"myenv" %in% virtualenv_list()) {
  virtualenv_create("myenv")
}
use_virtualenv("myenv", required = TRUE)
virtualenv_install(
  envname = "myenv",
  packages = c("palmerpenguins", "pandas", "numpy",
               "scikit-learn", "duckdb", "vetiver",
                "pins"))
```

Now we can import the Python packages/functions. 

```{python}
from palmerpenguins import load_penguins
from pandas import get_dummies
import numpy as np
from sklearn.linear_model import LinearRegression
from sklearn import preprocessing
import duckdb
```

### Using `duckdb`

```{r}
#| label: co_box_duckdb
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "r",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "`duckdb` error", 
  fold = TRUE,
  contents = "This originally returned an error: 
  
\`\`\`
Traceback (most recent call last):
  File \"<string>\", line 1, in <module>
duckdb.duckdb.InvalidInputException: Invalid Input Error: Python Object 
\"penguins\" of type \"module\" found on line \"<string>:1\" not suitable for
replacement scans.
Make sure that \"penguins\" is either a pandas.DataFrame, 
duckdb.DuckDBPyRelation, pyarrow Table, Dataset, RecordBatchReader, 
Scanner, or NumPy ndarrays with supported format

Traceback (most recent call last):
  File \"<string>\", line 1, in <module>
NameError: name 'df' is not defined
\`\`\`
  
  "
)
```

Below we **load the penguins data**, connect to `duckdb`, then create the table:

```{python}
penguins_data = load_penguins()
con = duckdb.connect('my-db.duckdb')
```

The `load_penguins()` function put the penguins data into a DataFrame, then we can insert it into DuckDB *before* querying:

```{python}
con.execute("CREATE OR REPLACE TABLE penguins AS SELECT * FROM penguins_data")
df = con.execute("SELECT * FROM penguins").fetchdf().dropna()
con.close()
```

Confirm the data: 

```{python}
df.head(3)
```

### Create a `model`

The model is created below (like lab 1). 

```{python}
X = get_dummies(df[['bill_length_mm', 'species', 'sex']], drop_first = True)
y = df['body_mass_g']

model = LinearRegression().fit(X, y)
```

### Print `model` information

This will print the model information. 

```{python}
print(f"R^2 {model.score(X,y)}")
print(f"Intercept {model.intercept_}")
print(f"Columns {X.columns}")
print(f"Coefficients {model.coef_}")
```

## `model` -> `VetiverModel`

The model is created with `VetiverModel()`, which is a **wrapper class** that takes our trained machine learning model and adds important metadata and functionality for deployment and versioning.

```{python}
from vetiver import VetiverModel
v = VetiverModel(model, model_name='penguin_model', prototype_data=X)
```

```{=html}
<style>

.codeStyle span:not(.nodeLabel) {
  font-family: monospace;
  font-size: 1.5em;
  font-weight: bold;
  color: #9753b8 !important;
  background-color: #f6f6f6;
  padding: 0.2em;
}
</style>
```

```{mermaid}
%%| fig-cap: 'VetiverModel()'
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%

flowchart LR
    RawMod("<strong>Raw sklearn model</strong>") --> VMod("<strong>VetiverModel()</strong><br>wrapper")
    VMod --> ModMeta("<strong>Model + Metadata</strong>")
    ModMeta --> Dep("<strong>Deployment Ready</strong>")
    
    subgraph Vet["<em>What it creates</em>"]
        InSch("Input Schema")
        ModTypeInfo("Model Type Info")
        Ver("Versioning")
        Api("API Generation")
        Ser("Serialization")
    end
    
    VMod --> InSch
    VMod --> ModTypeInfo 
    VMod --> Ver
    VMod --> Api
    VMod --> Ser

    %% Style for outputs %%
    style RawMod fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style VMod fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style ModMeta fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style Dep fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    style InSch fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style ModTypeInfo fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style Ver fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style Api fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style Ser fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style Vet fill:#fbf7ec,stroke:#2A6F77,color:#1B2A41
    
```

### What are `pins` boards? 

In this version, we're going to save our `pins` board to a temporary location (`f"{temp_dir}/models"`), but in the lab, we'll create the directory before creating the board. 

```{mermaid}
%%| fig-cap: 'board_folder()'
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%

flowchart LR
    BrdFldr("<strong>board_folder()</strong>") --> FSI("File System Interface")
    FSI --> VerStor("<strong>Versioned Storage</strong>")
    VerStor --> MetaTr("<strong>Metadata Tracking</strong>")
    
    subgraph BF["<em>What it creates</em>"]
        DirStr("Directory Structure")
        VerCont("Version Control")
        PinMngMnt("Pin Management")
    end
    
    FSI --> DirStr
    FSI --> VerCont
    FSI --> PinMngMnt

    %% Style for outputs %%
    style BrdFldr fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style FSI fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style VerStor fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style MetaTr fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    style DirStr fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style VerCont fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style PinMngMnt fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style BF fill:#fbf7ec,stroke:#2A6F77,color:#1B2A41


```


```{python}
import tempfile
from pins import board_folder
from vetiver import vetiver_pin_write

temp_dir = tempfile.mkdtemp() #<1>
model_board = board_folder(f"{temp_dir}/models", allow_pickle_read=True) #<2> 
vetiver_pin_write(model_board, v)
```
1. Use temp directory  
2. `allow_pickle_read` allows reading Python pickle files (needed for `scikit-learn` models) 

-  **`board_folder()`** creates the `pins` board that sets up a storage interface to a local temporary folder. This creates a standardized structure for storing "pins" (data objects/models).

-  **`vetiver_pin_write()`** receives the `model_board`, which is our `board_folder` object pointing to `temp_dir/models` and `v` (our `VetiverModel` wrapping the `LinearRegression`). 

```{python}
print(f"Model saved to: {temp_dir}/models")
```

```{python}
#| code-fold: true 
#| code-summary: 'show/hide print_tree()'
import os

def print_tree(directory, prefix="", max_depth=None, current_depth=0):
    """Print a tree structure of directory contents."""
    if max_depth is not None and current_depth > max_depth:
        return
    
    items = sorted(os.listdir(directory))
    for i, item in enumerate(items):
        path = os.path.join(directory, item)
        is_last = i == len(items) - 1
        
        current_prefix = "└── " if is_last else "├── "
        print(f"{prefix}{current_prefix}{item}")
        
        if os.path.isdir(path):
            extension = "    " if is_last else "│   "
            print_tree(path, prefix + extension, max_depth, current_depth + 1)
```

Below is a tree of the model folder: 

```{python}
print_tree(f"{temp_dir}/models")
```

### What is in `penguin_model/`?

The `penguin_model` board folder is described below: 

```{verbatim}
└── penguin_model #<1> 
    └── CURRENT-DATE-TIME-aba30/ #<2> 
        ├── data.txt  #<3> 
        └── penguin_model.joblib #<4>
```
1. `pin` name directory   
2. Version directory timestamp + hash   
3. Model metadata (what, when, how big)   
4. Serialized model (the actual trained model + `VetiverModel` wrapper with all the prediction logic)

The board folder handles versioning automatically and manages metadata about what's stored.

```{mermaid}
%%| fig-cap: 'vetiver_pin_write()'
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%

flowchart LR
    Mod("<strong>model</strong>") --> VetMod("<strong>VetiverModel(model)</strong>")
    VetMod --> VPW("<strong>vetiver_pin_write()</strong>")
    VPW --> Joblib("Serialized <strong>.joblib</strong>")
    VPW --> Meta("Metadata <strong>.txt</strong>")
    
    Rep("To<br>reproduce...") --> VPR("<strong>vetiver_pin_read()</strong>")
    VPR --> Joblib
    VPR --> Meta
    VPR --> Recon("Reconstructed<br><strong>VetiverModel()</strong>")

    %% Style for outputs %%
    style Mod fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style VetMod fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style VPW fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style Joblib fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    style Meta fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    style Rep fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style VPR fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style Recon fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff


```

## `VetiverModel` -> `VetiverAPI`

Finally, we'll turn the model into an API. 

```{python}
from vetiver import VetiverAPI
app = VetiverAPI(v, check_prototype = True)
```

**`VetiverAPI()`** creates a production-ready web API that wraps the trained model, validates inputs, and returns predictions via HTTP. 

```{mermaid}
%%| fig-cap: 'VetiverAPI()'
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%

flowchart LR
    VetMod("<strong>VetiverModel()</strong>") --> VetAPI("<strong>VetiverAPI()</strong>")
    VetAPI --> REST("<strong>REST</strong> API<br>Server")
    REST --> HttpEP("<strong>HTTP</strong><br>Endpoints")
    
    HttpEP --> Post("<strong>POST</strong><br>/predict")
    HttpEP --> GetPing("<strong>GET</strong><br>/ping")
    HttpEP --> GetProto("<strong>GET</strong><br>/prototype")
    HttpEP --> GetMeta("<strong>GET</strong><br>/metadata")

    %% Style for outputs %%
    style VetMod fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style VetAPI fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style REST fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style HttpEP fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    style Post fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style GetPing fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style GetProto fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff
    style GetMeta fill:#485466,stroke:#000000,stroke-width:1px,color:#ffffff

```

This makes our model accessible to any application that can make web requests, which we will cover in the next lab. 
