# Lab 1 {.unnumbered}

The [first lab](https://do4ds.com/chapters/sec1/1-1-env-as-code.html#lab1) includes two Quarto documents: 

```{r}
#| collapse: true
#| eval: false
#| echo: false
fs::dir_tree(path = "_labs/lab1/", recurse = 1)
```

```{verbatim}
_labs/lab1/
├── eda-basic.qmd
└── model-basic-py.qmd
```


## Project-oriented workflows

Key takeaways: 

1. Each project should have an **isolated** package environment that preserves the names and versions. 

2. The package environment should be captured in a way that makes is exportable/transferable to another machine. 
    - In R, this is `renv`. In Python, this is `virtualenv` or `venv`

3. Each project should have a virtual environment in a standalone directory that documents everything the project needs.[^rstats-wtf-projects]
    - This usually means a GitHub or GitLab repo. 

[^rstats-wtf-projects]: Read more in [What They Forgot to Teach You About R: Chapter 3, Project-oriented workflow](https://rstats.wtf/projects)

## Virtual environments

`venv` and `renv`, "*keep user-level package caches of the actual packages and use symlinks, so only one copy of the package is installed.*"

### `renv`

I've copied the `eda-basic.qmd` file into the `renv-eda-basic` folder and run `renv::init()`: 

```{r}
#| collapse: true
#| eval: false
#| echo: false
fs::dir_tree(
  path = "_labs/lab1/renv-eda-basic/", 
  all = TRUE, 
  recurse = 1)
```

```{verbatim}
_labs/lab1/renv-eda-basic/
├── .Rprofile
├── eda-basic.qmd
├── renv
│   ├── activate.R
│   ├── library
│   └── settings.json
├── renv-eda-basic.Rproj
└── renv.lock
```

As we can see, `renv` creates `.Rprofile`, `renv/` and `renv.lock` in the project's root directory. The `renv-eda-basic.Rproj` is specific to RStudio, but it also helps maintain the project-oriented workflow.  

### `venv`

Use the **Terminal** to navigate to the `venv-model-basic-py` folder and create the Python virtual environment: 

```bash
python3 -m venv .venv
```

This will create a `.venv/` folder inside `venv-model-basic-py`.

```{r}
#| collapse: true
#| eval: false
#| echo: false
fs::dir_tree(
  path = "_labs/lab1/venv-model-basic-py/", 
  all = TRUE, 
  recurse = 1)
```

```{verbatim}
_labs/lab1/venv-model-basic-py/
├── .venv
│   ├── bin
│   ├── include
│   ├── lib
│   └── pyvenv.cfg
└── model-basic-py.qmd
```


Activate the virtual environment:

```bash
source .venv/bin/activate
```

This should change the **Terminal** prompt from: 

```bash
Bash@MacBook venv-model-basic-py$ 
```

to: 

```bash
(.venv) Bash@MacBook venv-model-basic-py$
```

Now we can install required Python packages. 

```bash
pip install palmerpenguins pandas numpy scikit-learn
```

We will add `jupyter` for interactive outputs

```bash
pip install jupyter
```

To tell Quarto to use the virtual environment, we can create a `quarto.yml` file in the `venv-model-basic-py/` folder and add:

```yaml
project:
  type: default

execute:
  python: .venv/bin/python  
  # or .venv\Scripts\python.exe on windows
```

Or we can set the Python interpreter with the terminal:[^venv-win]

```bash
quarto use python .venv/bin/python
```

[^venv-win]: On Windows try `.venv\Scripts\python.exe`

Now, if we open the `model-basic-py.qmd` file in RStudio and click the **Render** button in the toolbar we should see the rendered output. 




