# Lab 2 {.unnumbered}


```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

Lab 2 introduces the[ `vetiver` package](https://rstudio.github.io/vetiver-r/) for modeling. 

```{r}
#| collapse: true
#| eval: false
fs::dir_tree(path = "_labs/lab2/")
## _labs/lab2/
## └── model-vetiver.qmd
```

## vetiver basics {#sec-vetiver-basics}

The `vetiver` package comes in both flavors: [R](https://vetiver.rstudio.com/) and [Python.](https://rstudio.github.io/vetiver-python/). We'll use the reticulate package below to install the other packages in the `model-vetiver.qmd` lab. 

```{r}
#| results: hide
#| message: false 
#| warning: false
library(reticulate)
use_virtualenv("myenv", required = TRUE)
virtualenv_install(
  envname = "myenv", 
  packages = c("palmerpenguins", "pandas", "numpy", 
               "scikit-learn", "duckdb", "vetiver", 
                "pins"))
```

```{python}
from palmerpenguins import load_penguins
from pandas import get_dummies
import numpy as np
from sklearn.linear_model import LinearRegression
from sklearn import preprocessing
import duckdb
```

## Using `duckdb`

The code below establishes a connection to a `duckdb` database. 

```{r}
#| label: co_box_duckdb
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "r",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "`duckdb` error", 
  fold = TRUE,
  contents = "This originally returned an error: 
  
\`\`\`
Traceback (most recent call last):
  File \"<string>\", line 1, in <module>
duckdb.duckdb.InvalidInputException: Invalid Input Error: Python Object 
\"penguins\" of type \"module\" found on line \"<string>:1\" not suitable for
replacement scans.
Make sure that \"penguins\" is either a pandas.DataFrame, 
duckdb.DuckDBPyRelation, pyarrow Table, Dataset, RecordBatchReader, 
Scanner, or NumPy ndarrays with supported format

Traceback (most recent call last):
  File \"<string>\", line 1, in <module>
NameError: name 'df' is not defined
\`\`\`
  
  "
)
```

Below we **load the penguins data** and **connect to DuckDB and create the table**:

```{python}
penguins_data = load_penguins()
con = duckdb.connect('my-db.duckdb')
```

The `load_penguins()` function put the penguins data into a DataFrame, then we can insert it into DuckDB *before* querying:

```{python}
con.execute("CREATE OR REPLACE TABLE penguins AS SELECT * FROM penguins_data")
df = con.execute("SELECT * FROM penguins").fetchdf().dropna()
con.close()
```

Confirm the data: 

```{python}
df.head(3)
```

## Create `model`

The model is created below (like lab 1). 

```{python}
X = get_dummies(df[['bill_length_mm', 'species', 'sex']], drop_first = True)
y = df['body_mass_g']

model = LinearRegression().fit(X, y)
```

## Model information

This will print the model information. 

```{python}
print(f"R^2 {model.score(X,y)}")
print(f"Intercept {model.intercept_}")
print(f"Columns {X.columns}")
print(f"Coefficients {model.coef_}")
```

## Turn into `VetiverModel`

The model is created with `VetiverModel()`, which is a **wrapper class** that takes our trained machine learning model and adds important metadata and functionality for deployment and versioning.

```{python}
from vetiver import VetiverModel
v = VetiverModel(model, model_name='penguin_model', prototype_data=X)
```

```{=html}
<style>

.codeStyle span:not(.nodeLabel) {
  font-family: monospace;
  font-size: 1.5em;
  font-weight: bold;
  color: #9753b8 !important;
  background-color: #f6f6f6;
  padding: 0.2em;
}
</style>
```

```{mermaid}
%%| fig-cap: 'VetiverModel()'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%
flowchart LR
    RawMod[Raw sklearn model] --> VMod["<code>VetiverModel()</code><br>wrapper"]
    VMod --> ModMeta["Model + Metadata"]
    ModMeta --> Dep["Deployment Ready"]
    
    subgraph Vet["<em>What it creates</em>"]
        InSch("Input Schema")
        ModTypeInfo("Model Type Info")
        Ver("Versioning")
        Api("API Generation")
        Ser("Serialization")
    end
    
    VMod --> InSch
    VMod --> ModTypeInfo 
    VMod --> Ver
    VMod --> Api
    VMod --> Ser
    
    %% style %%
    style RawMod fill:#FFF,stroke:#4CBB9D,color:#000,stroke-width:3px,rx:10,ry:10,font-size:16px
    style VMod fill:#FFF,stroke:#4CBB9D,color:#000,stroke-width:3px,rx:10,ry:10,font-size:16px
    style Vet fill:#FFF,stroke:#4CBB9D,color:#000,stroke-width:3px,rx:10,ry:10,font-size:16px
    
    style InSch fill:#FFF,stroke:#ff6f69,color:#000,stroke-width:2px,rx:5,ry:5,font-size:15px
    style ModTypeInfo fill:#FFF,stroke:#ff6f69,color:#000,stroke-width:2px,rx:5,ry:5,font-size:15px
    style Ver fill:#FFF,stroke:#ff6f69,color:#000,stroke-width:2px,rx:5,ry:5,font-size:15px
    style Api fill:#FFF,stroke:#ff6f69,color:#000,stroke-width:2px,rx:5,ry:5,font-size:15px
    style Ser fill:#FFF,stroke:#ff6f69,color:#000,stroke-width:2px,rx:5,ry:5,font-size:15px
    
    style ModMeta fill:#FFF,stroke:#ffcc5c,color:#000,stroke-width:3px,rx:10,ry:10,font-size:16px
    style Dep fill:#FFF,stroke:#ffcc5c,color:#000,stroke-width:3px,rx:10,ry:10,font-size:16px
    
```

## `pins` boards

In this version, we're going to save our `pins` board to a temporary location, but in the lab, we'll create the directory before creating the board. 

```{mermaid}
%%| fig-cap: 'board_folder()'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%
flowchart LR
    BrdFldr["<code>board_folder()</code>"] --> FSI[File System Interface]
    FSI --> VerStor[Versioned Storage]
    VerStor --> MetaTr[Metadata Tracking]
    
    subgraph BF["What it creates"]
        DirStr[Directory Structure]
        VerCont[Version Control]
        PinMngMnt[Pin Management]
    end
    
    FSI --> DirStr
    FSI --> VerCont
    FSI --> PinMngMnt
    
    style BrdFldr fill:#FFF,stroke:#4CBB9D,color:#000,stroke-width:3px,rx:10,ry:10,font-size:16px
    style FSI fill:#FFF,stroke:#4CBB9D,color:#000,stroke-width:3px,rx:10,ry:10,font-size:16px
    style BF fill:#FFF,stroke:#4CBB9D,color:#000,stroke-width:3px,rx:10,ry:10,font-size:16px
    
    style DirStr fill:#FFF,stroke:#ff6f69,color:#000,stroke-width:2px,rx:5,ry:5,font-size:15px
    style VerCont fill:#FFF,stroke:#ff6f69,color:#000,stroke-width:2px,rx:5,ry:5,font-size:15px
    style PinMngMnt fill:#FFF,stroke:#ff6f69,color:#000,stroke-width:2px,rx:5,ry:5,font-size:15px
    
    style VerStor fill:#FFF,stroke:#ffcc5c,color:#000,stroke-width:3px,rx:10,ry:10,font-size:16px
    style MetaTr fill:#FFF,stroke:#ffcc5c,color:#000,stroke-width:3px,rx:10,ry:10,font-size:16px

```

In `board_folder()`:  

- The **Path** (`f"{temp_dir}/models"`) is where to store everything    

- **`allow_pickle_read=True`** allows reading Python pickle files (needed for `scikit-learn` models)   

```{python}
import tempfile
from pins import board_folder
from vetiver import vetiver_pin_write

# use temp dir
temp_dir = tempfile.mkdtemp()
model_board = board_folder(f"{temp_dir}/models", allow_pickle_read=True)
vetiver_pin_write(model_board, v)
```

`board_folder()` creates a **pins board** that sets up a storage interface to a local temporary folder. This creates a standardized structure for storing "pins" (data objects/models). 

```{python}
print(f"Model saved to: {temp_dir}/models")
```

```{python}
#| code-fold: true 
#| code-summary: 'show/hide print_tree()'
import os

def print_tree(directory, prefix="", max_depth=None, current_depth=0):
    """Print a tree structure of directory contents."""
    if max_depth is not None and current_depth > max_depth:
        return
    
    items = sorted(os.listdir(directory))
    for i, item in enumerate(items):
        path = os.path.join(directory, item)
        is_last = i == len(items) - 1
        
        current_prefix = "└── " if is_last else "├── "
        print(f"{prefix}{current_prefix}{item}")
        
        if os.path.isdir(path):
            extension = "    " if is_last else "│   "
            print_tree(path, prefix + extension, max_depth, current_depth + 1)
```

```{python}
print_tree(f"{temp_dir}/models")
```

The board folder handles versioning automatically and manages metadata about what's stored.

```{verbatim}
└── penguin_model # pin name directory
    └── CURRENT-DATE-TIME-aba30/ # version directory timestamp + hash
        ├── data.txt  # model metadata
        └── penguin_model.joblib # serialized model
```

## Turn model into `VetiverAPI`

```{python}
from vetiver import VetiverAPI
app = VetiverAPI(v, check_prototype = True)
```