# Lab 3 {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

The original files in `_labs/lab3` are below: 

```{r}
#| collapse: true
#| eval: false
fs::dir_tree(path = "_labs/lab3/")
## _labs/lab3/
## ├── app-api.R
## ├── app-api.py
## ├── eda-db.qmd
## ├── model-db.qmd
## ├── py-data-load.py
## └── r-data-load.R
```

In the previous labs, we build a model API with [`vetiver`](https://rstudio.github.io/vetiver-python/stable/) (also [available in R](https://rstudio.github.io/vetiver-r/)). This lab continues with accessing databases and using APIs. 

## APIs 

Let's back up a bit and get an understanding of what an API does. In general, APIs are the digital plumbing that makes modern software ecosystems possible. 

```{mermaid}
%%| fig-cap: 'APIs'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%
graph LR
    DS["Data Scientist"] --> Req["API Request"]
    Req --> ApiServ["API Server"]
    ApiServ --> DataProcMod["Data Processing/Model"]
    DataProcMod --> Res["Results"]
    Res --> ApiServ
    ApiServ --> APIResp["API Response"]
    APIResp --> DS
    
    style ApiServ fill:#e1f5fe
    style DataProcMod fill:#f3e5f5
```

APIs have standardized language that allows different software systems to communicate and share functionality (without needing to understand how the internal implementation works).

**What do they communicate?**

1. What you can request (endpoints)   
2. How to make the request (HTTP methods, parameters)   
3. What you'll get back (response format)   
4. When things go wrong (error codes)  

In the previous lab we turned our model into a `VetiverModel()`, saved it to a `pins` board, and then created a `VetiverAPI()`:

```{mermaid}
%%| fig-cap: 'Our pipeline'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"14px"}}}%%
graph LR
    subgraph DataPrep["Data Preparation"]
        Penguins["<code>penguins</code><br>data"] --> DuckDB["put in<br><code>duckdb</code>"]
        DuckDB
    end
    
    subgraph ModTrain["Linear Regression"]
        DuckDB --> LinReg["create<br><code>model</code>"]
        LinReg
    end
    
    subgraph MLOps["MLOps"]
        LinReg --> VetMod["create<br><code>VetiverModel()</code>"]
        VetMod --> Pins["save to<br><code>pins</code> board"]
        Pins --> VetAPI["create<br><code>VetiverAPI()</code>"]
    end
    
    style ModTrain fill:#e1f5fe
    style MLOps fill:#f3e5f5
    style VetAPI fill:#e8f5e8
```

I've also completed these steps in R, which can be found in [`_labs/lab2/R/model-vetiver-r.qmd`](https://github.com/mjfrigaard/do4ds-labs/tree/main/_labs/lab2/R/model-vetiver-r.qmd). 

The sections below cover how to create and communicate with APIs.  I've split each lab into separate `R/` and `Python/` folders and included RStudio project files. 

## R App API

The R App API files can be found in the [`_labs/lab3/R/` folder](https://github.com/mjfrigaard/do4ds-labs/tree/main/_labs/lab3/R). 

```{r}
#| collapse: true
#| eval: false
fs::dir_tree(path = "_labs/lab3/R/")
## _labs/lab3/R/
## ├── R.Rproj
## ├── app-api.R
## ├── eda-db.qmd
## └── r-data-load.R
```

### eda-db.qmd {#sec-eda-db.qmd}

The [`eda-db.qmd`](_labs/lab3/eda-db.qmd) file has the EDA R code from lab 1. 

It begins by loading the libraries: 

```{r}
#| eval: false
library(DBI) # add this
library(duckdb) # add this
library(palmerpenguins) # add this
library(dplyr)
library(ggplot2)
library(dbplyr)
```

The code below creates a connection to the database, but needs to be changed slightly so the `penguins` data is registered with `duckdb` (before the queries can be run).

```{r}
#| eval: false
# load penguins data first
penguins_data <- palmerpenguins::penguins
# then connect
con <- DBI::dbConnect(
  duckdb::duckdb(), 
  dbdir = "my-db.duckdb"
)
# register table
duckdb_register(con, "penguins_data", penguins_data)
# create table
dbExecute(con, "CREATE OR REPLACE TABLE penguins AS SELECT * FROM penguins_data")
# create df 
df <- dbGetQuery(con, "SELECT * FROM penguins") |> na.omit()
```

The table below displays the means of the measurements (body mass, flipper length, and bill length/depth) by species and sex:

```{r}
#| eval: false
df %>%
  group_by(species, sex) %>%
  summarise(
    across(
        ends_with("mm") | ends_with("g"),
      \(x) mean(x, na.rm = TRUE)
      )
    ) %>%
  dplyr::collect() %>%
  knitr::kable()
```

The graph displays the best-fit linear relationship between `bill_length_mm` and `body_mass_g` by `species`:

```{r}
#| eval: false
df %>%
  ggplot(
    aes(
    x = bill_length_mm, 
    y = body_mass_g, 
    color = species)
    ) +
  geom_point() + 
  geom_smooth(method = "lm")
```

And we finish by closing the connection to the database.

```{r}
#| eval: false
DBI::dbDisconnect(con)
```

To view the original code in `eda-db.qmd`, expand the section below:

<details>
  <summary>Expand to view eda-db.qmd</summary>
  
```{r}
#| label: print-eda-db.qmd
#| echo: false
library(readr)
readr::read_lines("_labs/lab3/eda-db.qmd") |> 
  base::noquote() |> 
  writeLines()
```

</details>

### app-api.R 

`app-api.R` creates an R Shiny application that calls the `vetiver` API. The top of the file creates an `api_url`:

```{r}
#| eval: false
api_url <- "http://127.0.0.1:8080/predict"
```

#### UI

The UI definition is below: 

```{r}
#| eval: false
ui <- fluidPage(
  titlePanel("Penguin Mass Predictor"),

  # Model input values
  sidebarLayout(
    sidebarPanel(
      sliderInput(
        "bill_length",
        "Bill Length (mm)",
        min = 30,
        max = 60,
        value = 45,
        step = 0.1
      ),
      selectInput(
        "sex",
        "Sex",
        c("Male", "Female")
      ),
      selectInput(
        "species",
        "Species",
        c("Adelie", "Chinstrap", "Gentoo")
      ),
      # Get model predictions
      actionButton(
        "predict",
        "Predict"
      )
    ),

    mainPanel(
      h2("Penguin Parameters"),
      verbatimTextOutput("vals"),
      h2("Predicted Penguin Mass (g)"),
      textOutput("pred")
    )
  )
)
```

#### Server

The server definition is below: 

```{r}
#| eval: false
server <- function(input, output) {
  # Input params
  vals <- reactive(
    list(
      bill_length_mm = input$bill_length,
      species_Chinstrap = input$species == "Chinstrap",
      species_Gentoo = input$species == "Gentoo",
      sex_male = input$sex == "Male"
    )
  )

  # Fetch prediction from API
  pred <- eventReactive(
    input$predict,
    httr2::request(api_url) |>
      httr2::req_body_json(vals()) |>
      httr2::req_perform() |>
      httr2::resp_body_json(),
    ignoreInit = TRUE
  )

  # Render to UI
  output$pred <- renderText(pred()$predict[[1]])
  output$vals <- renderPrint(vals())
}
```

```{r}
#| eval: false
shinyApp(ui = ui, server = server)
```

### r-data-load.R {#sec-r-data-load-r}

```{r}
#| eval: false
con <- DBI::dbConnect(duckdb::duckdb(), dbdir = "my-db.duckdb")

DBI::dbWriteTable(con, "penguins", palmerpenguins::penguins)

DBI::dbDisconnect(con)
```


To view the original code in `r-data-load.R`, expand the section below:

<details>
  <summary>Expand to view r-data-load.R</summary>

```{r}
#| label: print-r-data-load-R
#| echo: false
#| code-overflow: wrap
library(readr)
readr::read_lines("_labs/lab3/r-data-load.R") |> 
  base::noquote() |> 
  writeLines()
```

</details>

## Python App API

The Python App API files can be found in the [`_labs/lab3/Python/` folder](https://github.com/mjfrigaard/do4ds-labs/tree/main/_labs/lab3/Python). 

```{r}
#| collapse: true
#| eval: false
fs::dir_tree(path = "_labs/lab3/Python/")
## _labs/lab3/Python/
## ├── Python.Rproj
## ├── app-api.py
## ├── model-db.qmd
## └── py-data-load.py
```

### model-db.qmd

### app-api.py

`app-api.py` creates an Python Shiny application that calls the `vetiver` API. 

### py-data-load.py 




