# Lab 3 {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

The original files in `_labs/lab3` are below: 

```{r}
#| collapse: true
#| eval: false
fs::dir_tree(path = "_labs/lab3/")
## _labs/lab3/
## ├── app-api.R
## ├── app-api.py
## ├── eda-db.qmd
## ├── model-db.qmd
## ├── py-data-load.py
## └── r-data-load.R
```

In the previous labs, we build a model API with [`vetiver`](https://rstudio.github.io/vetiver-python/stable/) (also [available in R](https://rstudio.github.io/vetiver-r/)). This lab continues with accessing databases and using APIs. I've split these files into separate folders (i.e., projects) and included RStudio project files: 

```{r}
#| collapse: true
#| eval: true
fs::dir_tree(path = "_labs/lab3/Python/")
```

```{r}
#| collapse: true
#| eval: true
fs::dir_tree(path = "_labs/lab3/R/")
```

## APIs 

Let's back up a bit and get an understanding of what an API does. In general, APIs are the digital plumbing that makes modern software ecosystems possible. 

```{mermaid}
%%| fig-cap: 'APIs'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"16px"}}}%%
graph LR
    DS["Data Scientist"] --> Req["API Request"]
    Req --> ApiServ["API Server"]
    ApiServ --> DataProcMod["Data Processing/Model"]
    DataProcMod --> Res["Results"]
    Res --> ApiServ
    ApiServ --> APIResp["API Response"]
    APIResp --> DS
    
    style ApiServ fill:#e1f5fe
    style DataProcMod fill:#f3e5f5
```

APIs have standardized language that allows different software systems to communicate and share functionality (without needing to understand how the internal implementation works).

**What do they communicate?**

1. What you can request (endpoints)   
2. How to make the request (HTTP methods, parameters)   
3. What you'll get back (response format)   
4. When things go wrong (error codes)  

In the previous lab we turned our model into a `VetiverModel()`, saved it to a `pins` board, and then created a `VetiverAPI()`:

```{mermaid}
%%| fig-cap: 'Our pipeline'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"14px"}}}%%
graph LR
    subgraph DataPrep["Data Preparation"]
        Penguins["<code>penguins</code><br>data"] --> DuckDB["put in<br><code>duckdb</code>"]
        DuckDB
    end
    
    subgraph ModTrain["Linear Regression"]
        DuckDB --> LinReg["create<br><code>model</code>"]
        LinReg
    end
    
    subgraph MLOps["MLOps"]
        LinReg --> VetMod["create<br><code>VetiverModel()</code>"]
        VetMod --> Pins["save to<br><code>pins</code> board"]
        Pins --> VetAPI["create<br><code>VetiverAPI()</code>"]
    end
    
    style ModTrain fill:#e1f5fe
    style MLOps fill:#f3e5f5
    style VetAPI fill:#e8f5e8
```

## Apps 

Both `app-api.R` and `app-api.py` create Shiny applications. 

### app-api.R 

When we initially launch the app, we'll see the following error: 

```{verbatim}
Warning: Error in httr2::req_perform: Failed to perform HTTP request.
Caused by error in `curl::curl_fetch_memory()`:
! Could not connect to server [127.0.0.1]:
Failed to connect to 127.0.0.1 port 8080 after 0 ms: Could not connect to server
  163: <Anonymous>
  162: signalCondition
  161: signal_abort
  160: cnd_signal
  159: handle_resp
  158: httr2::req_perform
  154: eventReactiveValueFunc
  102: pred
  101: renderText
  100: func
   87: renderFunc
   86: output$pred
    1: runApp
```

### app-api.py

## EDA

## eda-db.qmd

## model-db.qmd

## Loading data 

### py-data-load.py 

### r-data-load.R
