# Lab 4: Logging and Monitoring {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

The original files in `_labs/lab4` are below: 

```{r}
#| collapse: true
#| eval: false
fs::dir_tree(path = "_labs/lab4/")
## _labs/lab4/
## ├── app-log.R #<1>
## └── app-log.py #<2>
```
1. R Shiny app    
2. Python Shiny app    

## Comprehension questions



### Monitoring vs. logging

> 1. *What is the difference between monitoring and logging? What are the two halves of the monitoring and logging process?*



### What not to log

> 2. *Logging is generally good, but what are some things you should be careful not to log?*



### Log levels 

> 3. *At what level would you log each of the following events:*
> *Someone clicks on a particular tab in your Shiny app.*



> *Someone puts an invalid entry into a text entry box.*



> *An HTTP call your app makes to an external API fails.*



> *The numeric values that are going into your computational function.*

## Logging in R 


<details>
  <summary>Expand to view app-log.R</summary>

```{r}
#| label: print-app-log-R
#| echo: false
#| code-overflow: wrap
library(readr)
readr::read_lines("_labs/lab4/app-log.R") |> 
  base::noquote() |> 
  writeLines()
```

</details>

## Logging in Python 

<details>
  <summary>Expand to view app-log.py</summary>

```{r}
#| label: print-app-log-py
#| echo: false
#| code-overflow: wrap
library(readr)
readr::read_lines("_labs/lab4/app-log.py") |> 
  base::noquote() |> 
  writeLines()
```

</details>