# Lab 4: Logging and Monitoring {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

The original files in `_labs/lab4` are below: 

```{r}
#| collapse: true
#| eval: false
fs::dir_tree(path = "_labs/lab4/")
## _labs/lab4/
## ├── app-log.R #<1>
## └── app-log.py #<2>
```
1. R Shiny app    
2. Python Shiny app    

## Comprehension questions



### Monitoring vs. logging

> *What is the difference between monitoring and logging? What are the two halves of the monitoring and logging process?*

Logging is a record of "*What Happened*"--it captures discrete events and detailed narratives of what occurred in the system at specific points in time.

Monitoring is usually a dashboard of "*How Is It Performing*"--tracking quantitative metrics and system health indicators over time to understand trends and current state.

```{mermaid}
%%| fig-cap: 'Logging vs. Monitoring'
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace'}}}%%

graph LR
    subgraph "LOG: events"
        L1[/"Discrete Events"/]
        L2[/"Detailed Context"/]
        L3[/"Historical Investigation"/]
        L4[/"Text-Based Messages"/]
        L5(["<strong>When/What/Who/Why</strong>"])
    end
    
    subgraph "MONITOR: metrics"
        M1[/"Quantitative Metrics"/]
        M2[/"Real-time Status"/]
        M3[/"Trend Analysis"/]
        M4[/"Numerical Data"/]
        M5(["<strong>How Much/How Fast/How Often</strong>"])
    end
    
    subgraph "Common Use Cases"
        LC[\"Root Cause Analysis"\]
        LC2[\"Debugging Specific Issues"\]
        LC3[\"Compliance Auditing"\]
        LC4[\"Security Investigation"\]
        
        MC[\"Performance Optimization"\]
        MC2[\"Capacity Planning"\]
        MC3[\"SLA Monitoring"\]
        MC4[\"Real-time Alerting"\]
    end
    
    L1 --> LC
    L2 --> LC2
    L3 --> LC3
    L4 --> LC4
    
    M1 --> MC
    M2 --> MC2
    M3 --> MC3
    M4 --> MC4
    
    classDef logging fill:#e3f2fd
    classDef monitoring fill:#f3e5f5
    classDef usecase fill:#f1f8e9
    
    class L1,L2,L3,L4,L5 logging
    class M1,M2,M3,M4,M5 monitoring
    class LC,LC2,LC3,LC4,MC,MC2,MC3,MC4 usecase
```


| Aspect        | Logging                        | Monitoring                     |
| ------------- | ------------------------------ | ------------------------------ |
| **Data Type** | Discrete events, text messages | Metrics, counters, gauges      |
| **When Used** | After something happens        | Continuously                   |
| **Purpose**   | Debugging, investigation       | Performance tracking, alerting |
| **Format**    | Structured/unstructured text   | Numerical time-series data     |
| **Storage**   | Log files, log databases       | Time-series databases          |
| **Analysis**  | Search, filter, correlate      | Aggregate, trend, threshold    |



### What not to log

> *Logging is generally good, but what are some things you should be careful not to log?*

Critical data we should make sure not to log includes: 

1. Sensitive Personal Information (PII)   
    - Email addresses, phone numbers, names   
    - Social Security Numbers, passport numbers   
    - Addresses, birth dates    
    - Medical information (HIPAA protected)   

2. Authentication Credentials     
    - Passwords (plaintext or hashed)     
    - API keys and tokens   
    - Database connection strings with credentials    
    - OAuth tokens and refresh tokens   

3. Financial Information    
    - Credit card numbers     
    - Bank account numbers    
    - Routing numbers   

### Log levels 

> *At what level would you log each of the following events:*
> *Someone clicks on a particular tab in your Shiny app.*



> *Someone puts an invalid entry into a text entry box.*



> *An HTTP call your app makes to an external API fails.*



> *The numeric values that are going into your computational function.*

## Observability

While we cannot prevent code failures (or when they occur), we can create [**observable systems**](https://en.wikipedia.org/wiki/Observability_(software)) that 1) **accept failure as inevitable** (rather than trying to prevent it), 2) **implement comprehensive monitoring and logging**, and 3) **enable rapid diagnosis** (by using logs to reconstruct failure pathways).


To do this, data scientist and developers need to emit meaningful logs and metrics from their code. As system admins, we need to ensure new data science work can be easily integrated within our organizational infrastructure (i.e., into the existing log aggregation and monitoring tools).

```{mermaid}
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"18px"}}}%%

sequenceDiagram
    participant DS as Data Science<br>Code
    participant Emit as Emit
    participant Collect as Collect
    participant Process as Process
    participant Store as Store
    participant Alert as Alert
    participant Display as Display
    participant Team as DevOps<br>Team
    
    Note over DS: User makes<br>prediction<br>request
    
    DS->>Emit: log_info("prediction started")
    DS->>Emit: emit_metric("response_time", 1.2)
    
    Collect->>Emit: Scrape logs<br>& metrics
    Collect->>Process: Logs & metrics to<br>processing pipeline
    
    Process->>Store: Parse &<br>enrich data
    Store->>Alert: Check alert<br>conditions
    Store->>Display: Update real-time<br>displays
    
    alt Response time > 5 seconds
        Alert->>Team: Send notifications
        Team->>Display: Investigate<br>using logs
    end
    
    Display-->>Team: Show trends<br>and KPIs
```

This provides 1) operational peace of mind (the ability to detect and resolve issues quickly), 2) value demonstration (shows decision makers who uses your work and how), and 3) team validation (prove the team's impact and justify resources).

## Logging in R 


<details>
  <summary>Expand to view app-log.R</summary>

```{r}
#| label: print-app-log-R
#| echo: false
#| code-overflow: wrap
library(readr)
readr::read_lines("_labs/lab4/app-log.R") |> 
  base::noquote() |> 
  writeLines()
```

</details>

## Logging in Python 

<details>
  <summary>Expand to view app-log.py</summary>

```{r}
#| label: print-app-log-py
#| echo: false
#| code-overflow: wrap
library(readr)
readr::read_lines("_labs/lab4/app-log.py") |> 
  base::noquote() |> 
  writeLines()
```

</details>