# Lab 4: Logging and Monitoring {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

The original files in `_labs/lab4` are below: 

```{r}
#| collapse: true
#| eval: false
fs::dir_tree(path = "_labs/lab4/")
## _labs/lab4/
## ├── app-log.R #<1>
## └── app-log.py #<2>
```
1. R Shiny app    
2. Python Shiny app    

## Comprehension questions



### Monitoring vs. logging

> *What is the difference between monitoring and logging? What are the two halves of the monitoring and logging process?*

Logging is a record of "***What happened***"--logs capture discrete events and detailed narratives of what occurred in the system at specific points in time.

Monitoring is usually a display of "***How is it performing***"--tracking quantitative metrics and system health indicators over time to understand trends and current state.

```{mermaid}
%%| fig-cap: 'Logging vs. Monitoring'
%%| fig-align: center
%%| fig-width: 6.5
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace'}}}%%

graph LR
    subgraph "LOG: events"
        L1(["Discrete<br>Events"])
        L2(["Detailed<br>Context"])
        L3(["Historical<br>Investigation"])
        L4(["Text-Based<br>Messages"])
        L5("<strong>When/What/Who/Why</strong>")
    end
    
    subgraph "MONITOR: metrics"
        M1(["Quantitative<br>Metrics"])
        M2(["Real-time<br>Status"])
        M3(["Trend<br>Analysis"])
        M4(["Numerical<br>Data"])
        M5("<strong>How Much/How Fast/<br>How Often</strong>")
    end
    
    subgraph "Common Use Cases"
        LC(["Root Cause<br>Analysis"])
        LC2(["Debugging Specific<br>Issues"])
        LC3(["Compliance<br>Auditing"])
        LC4(["Security<br>Investigation"])
        
        MC(["Performance<br>Optimization"])
        MC2(["Capacity<br>Planning"])
        MC3(["SLA Monitoring"])
        MC4(["Real-time<br>Alerting"])
    end
    
    L1 --> LC
    L2 --> LC2
    L3 --> LC3
    L4 --> LC4
    
    M1 --> MC
    M2 --> MC2
    M3 --> MC3
    M4 --> MC4
    
    classDef logging fill:#d8e4ff,stroke:#333,stroke-width:2px,rx:6,ry:6,text-align:center,font-size:13px
    classDef monitoring fill:#d8e4ff,stroke:#333,stroke-width:2px,rx:6,ry:6,text-align:center,font-size:13px
    classDef usecase fill:#31e981,stroke:#333,stroke-width:1px,rx:6,ry:6,text-align:center,font-size:13px
    
    class L1,L2,L3,L4 logging
    class M1,M2,M3,M4 monitoring
    class LC,LC2,LC3,LC4,MC,MC2,MC3,MC4 usecase
```


| Aspect        | Logging                        | Monitoring                     |
| ------------- | ------------------------------ | ------------------------------ |
| **Data Type** | Discrete events, text messages | Metrics, counters, gauges      |
| **When Used** | After something happens        | Continuously                   |
| **Purpose**   | Debugging, investigation       | Performance tracking, alerting |
| **Format**    | Structured/unstructured text   | Numerical time-series data     |
| **Storage**   | Log files, log databases       | Time-series databases          |
| **Analysis**  | Search, filter, correlate      | Aggregate, trend, threshold    |



### What not to log

> *Logging is generally good, but what are some things you should be careful not to log?*

Critical data we should make sure not to log includes: 

1. Sensitive Personal Information (PII)   
    - Email addresses, phone numbers, names   
    - Social Security Numbers, passport numbers   
    - Addresses, birth dates    
    - Medical information (HIPAA protected)   

2. Authentication Credentials     
    - Passwords (plaintext or hashed)     
    - API keys and tokens   
    - Database connection strings with credentials    
    - OAuth tokens and refresh tokens   

3. Financial Information    
    - Credit card numbers     
    - Bank account numbers    
    - Routing numbers   

Industry specific guidelines on logging are available below: 

1. Financial Services:  
    - Never log account numbers in full     
    - Mask payment card data    
    - Avoid logging transaction details in plain text   
        - Source: [FFIEC IT Examination Handbook](https://ithandbook.ffiec.gov/it-booklets/information-security.aspx)

2. Healthcare   
    - De-identify all patient data in logs      
    - Remove [18 HIPAA identifiers](https://www.luc.edu/its/aboutus/itspoliciesguidelines/hipaainformation/the18hipaaidentifiers/)      
    - Implement access controls for log files   
        - Source: [HHS Security Risk Assessment](https://www.hhs.gov/hipaa/for-professionals/security/guidance/cybersecurity/index.html)

3. Government/Defense     
    - Classify log data appropriately     
    - Implement need-to-know access       
    - Encrypt logs containing sensitive information     
        - Source: [NIST SP 800-53](https://csrc.nist.gov/publications/detail/sp/800-53/rev-5/final)



### Log levels 

> *At what level would you log each of the following events:*
> *Someone clicks on a particular tab in your Shiny app.*



> *Someone puts an invalid entry into a text entry box.*



> *An HTTP call your app makes to an external API fails.*



> *The numeric values that are going into your computational function.*

## Observability

We can't prevent code failures (or when they happen), but we *can* create [**observable systems**](https://en.wikipedia.org/wiki/Observability_(software)) that 1) **accept failure is inevitable** (rather than trying to prevent it), 2) **implement comprehensive monitoring and logging**, and 3) **enable rapid diagnosis** (i.e., using logs and metrics to reconstruct what failed and why).

To do this, data scientist and developers need to generate meaningful logs and metrics from their code. System admins need to ensure new data science work can be easily integrated within the organizational infrastructure (i.e., into the existing log aggregation and monitoring tools).

```{mermaid}
%%| fig-width: 8.0
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace'}}}%%

graph LR
    subgraph DatSci["Data Scientist"]
        A(["Generate Logs"])
        B(["Emit Metrics"])
    end
    
    subgraph SysAdmin["System Admins"]
        C("Collect &<br>Aggregate")
        D("Store &<br>Process")
        E("Monitor &<br>Alert")
    end
    
    subgraph Out["Outcomes"]
        F(["Detect<br>Issues"])
        G(["Diagnose<br>Problems"])
        H(["Rapid<br>Recovery"])
    end
    
    A ---> C
    B ---> C
    C --> D
    D --> E
    E --> F & G & H

    
    
    classDef ds fill:#d8e4ff,stroke:#000,stroke-width:1px
    class A,B ds
    
    classDef outcome fill:#31e981,stroke:#000,stroke-width:1px
    class F,G,H outcome
    
    
```

This will give us: 

1) Operational peace of mind (the ability to detect and resolve issues quickly)

2) Value demonstration (shows decision makers who uses the work and how) 

3) Team validation (prove the team's impact and justify resources)

## Logging in R 


<details>
  <summary>Expand to view app-log.R</summary>

```{r}
#| label: print-app-log-R
#| echo: false
#| code-overflow: wrap
library(readr)
readr::read_lines("_labs/lab4/app-log.R") |> 
  base::noquote() |> 
  writeLines()
```

</details>

## Logging in Python 

<details>
  <summary>Expand to view app-log.py</summary>

```{r}
#| label: print-app-log-py
#| echo: false
#| code-overflow: wrap
library(readr)
readr::read_lines("_labs/lab4/app-log.py") |> 
  base::noquote() |> 
  writeLines()
```

</details>