# Lab 5: Deployments and Code Promotion {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

## Comprehension questions

The questions below come from the [Deployments and Code Promotion](https://do4ds.com/chapters/sec1/1-5-deployments.html) chapter. 

### Data Science Environments 

> ***Write down a mental map of the relationship between the three environments for data science. Include the following terms: Git Promote, CI/CD, automation, deployment, dev, test, prod.***

The three environments for data science are: development, test, and production (not to be confused with the three *layers* of data science environments[^layers]).

The figures from the [chapter](https://do4ds.com/chapters/sec1/1-5-deployments.html#separate-the-prod-environment) do an excellent job illustrating these environments, but I've added a mindmap:

[^layers]: These [three layers](https://do4ds.com/chapters/sec1/1-1-env-as-code.html#environments-have-layers) are Packages (Python + R Packages), System (Python + R Packages, Other System Libraries, and Operating System), and Hardware (Virtual Hardware and Physical Hardware).

```{=html}
<style>

.codeStyle span:not(.nodeLabel) {
  font-family: monospace;
  font-size: 1.5em;
  font-weight: bold;
  color: #d80808 !important;
  background-color: #ffffff;
  padding: 0.2em;
}
</style>
```

```{mermaid}
%%| fig-cap: 'Data Science Environments'
%%| fig-width: 7.0
%%| fig-align: center
%%{init: {'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"18px"}}}%%
mindmap
  root((Code<br>Promotion))
    Dev
      Code Writing
      Experimentation
      Git Promote
        Push to feature branch
        Pull request triggers CI/CD
    Test
      Automation
        Unit tests
        Integration tests
        UAT
      CI/CD
        Validates build
        Runs test suite
    Prod
      Deployment
        Automated release
        Version tagged
      Monitoring
        Stability checks
        Access control

```

#### DEV Environments 

The development environment (**dev**) is the primary location where data scientists write new code. It's typically used for iteration and experimentation, and it's connected to version control (Git).  

#### Test Environments 

The testing environment (**test**) is a mirror of the production environment with realistic data (often anonymized). Test environments are integrated and validates with other systems, and user acceptance testing occurs here.  

#### Production Environments 

The 'live' environment that serves the end users is the production environment (**prod**). Prod has strictly controlled access with a focus on stability and monitoring.  


### Git & Data Science 

> ***Why is Git so important to a good code promotion strategy?***

Code promotion is the process of tagging a Git commit as ready for production. This creates a reference point, and if desired, can trigger a production deployment pipeline. If a change isn’t represented as a Git commit, it literally didn’t happen. Code promotion ensures consistency across all three environments and reduces human error in deployment. 

All commits are signed--we know who made the changes--and the prod branch requires approval to merge (i.e., it's protected). This artifact immutability connects 'builds' to specific commit SHAs or by tagging releases. The same code is promoted across environments so we can be sure untested code never reaches production. 

Without Git, the code would need to be rebuilt in each environment, and the tested code wouldn't necessarily be the deployed code, causing environment drift. Environment drift is when the code in production differs from the code that was tested. In DevOps, environment drift is the enemy because it leads to unpredictable behavior and makes debugging difficult. 

Git enables environment parity. The three environments (dev, test, and prod) differ only by data and credentials, and changes are promoted, not re-created. When production inevitably breaks, we can use Git to roll back by redeploying previous commits/tags. This avoids making ad-hoc changes in a production server.

-----------------------------------------------------------

> ***Can you have a code promotion strategy without Git?***

Sure — code promotion can be done without Git, but you'd need to find a way to:

* Create some kind of 'immutable versioned artifact' (i.e., something to be promoted)  
* Blocks/gates requiring tests/approvals that must be passed before promotion    
* A promotion mechanism to copy/move/retag the immutable versioned artifact between environments       

Also need substitutes for:

1. A way to view/review 'diffs', either by performing code review (or document-style review) on release bundles 
2. The ability to keep previous artifacts and switch the environments     
3. Some means of traceability that connects the artifacts, manifests, approvals, and  build logs   


### Git & CI/CD

> ***What is the relationship between Git and CI/CD?***

A CI/CD Pipeline is an automated process that builds, tests, and deploys the code. In CI-CD systems, Git defines the code promotion workflow, but not the pipeline. The pipeline simply *reacts* to Git events. The code promotion logic lives in the Git branches, tags, merge policies, but not in brittle pipeline scripts. This enables a "build once, promote many times" strategy, where the same tested artifact is promoted across environments.

Git provides the code change/approval history, and the pipeline results tied to commits. Each release is also tagged, making it easy to track commits and introduced changes. This traceability is crucial for debugging and accountability. Git serves as the gatekeeper for CI-CD policies and integrates directly with enforcement mechanisms. 

Using Git turns CI-CD into a system, not a script. 

```{mermaid}
%%| fig-cap: 'Git & CI/CD'
%%| fig-width: 7.0
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"18px"}}}%%
flowchart LR
    GitEv["Git Event<br>(push, PR, merge)"] -->|"triggers"| CICD["CI/CD Pipeline"]
    CICD --> CI["Continuous<br>Integration (CI)<br>Build & Test"]
    CICD --> CD["Continuous<br>Delivery (CD)<br>Deploy"]

    style GitEv fill:#F9E79F,stroke:#D4AC0D,color:#000
    style CICD fill:#D4E6F1,stroke:#2E86C1,color:#000
    style CI fill:#D5F5E3,stroke:#1E8449,color:#000
    style CD fill:#FADBD8,stroke:#C0392B,color:#000
```

Think of Git as the source of truth and CI/CD as an automated response to changes in that truth. Git events — pushes, pull requests, merges — can fire a CI/CD pipeline.


```{mermaid}
%%| fig-cap: 'Git & CI/CD'
%%| fig-width: 7.0
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"18px"}}}%%
sequenceDiagram
    actor Dev as Developer
    participant Git as Git
    participant CI as CI Pipeline
    participant CD as CD Pipeline
    participant Test as Test Env
    participant Prod as Prod Env

    Dev->>Git: Push to feature branch
    Git->>CI: Triggers build & tests
    CI-->>Dev: ✅ Tests pass / ❌ Fail

    Dev->>Git: Merge PR to main
    Git->>CD: Triggers deployment
    CD->>Test: Deploy to test env
    CD-->>Dev: ✅ Test deployment confirmed

    Dev->>Git: Tag release (v1.0)
    Git->>CD: Triggers production deployment
    CD->>Prod: Deploy to prod env
```

**Key distinctions between Git and CI/CD:**

| | Git | CI/CD |
|---|---|---|
| **Role** | Tracks & versions code | Automates build, test, deploy |
| **Trigger** | Human action (commit, merge) | Git events |
| **Output** | A versioned history | A tested, deployed artifact |


-----------------------------------------------------------


> ***What’s the benefit of using Git and CI/CD together?***

Without Git, CI/CD has no reliable source to build from. Without CI/CD, Git changes require manual promotion between environments — which is slow and error-prone.

```{mermaid}
%%| fig-cap: 'Git & CI/CD'
%%| fig-width: 7.0
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"18px"}}}%%

gitGraph
   commit id: "Initial commit"
   
   branch feature
   checkout feature
   commit id: "New feature code"
   commit id: "Unit tests pass ✅"
   
   checkout main
   merge feature id: "Merge PR → triggers CI"
   commit id: "Deploy to Test ✅"
   commit id: "UAT approved ✅"
   commit id: "Tag release v1.0 → triggers CD"
   
   branch prod
   checkout prod
   commit id: "Deploy to Prod 🚀"
```

In the `gitGraph` above: 

1. The `feature` branch is where the developer writes and tests code locally    
2. When a pull request is merged to `main`, this fires the CI pipeline (build & test) and deploys to `test`     
3. `v1.0` is tagged to a release on the `main` branch, which fires the CD pipeline and deploys to `prod`       
4. The `prod` branch represents the exact code running in the live environment, providing a clear snapshot for auditing and rollback

### Git, GitHub, CI/CD, Actions, & Version Control 

> ***Write out a mental map of the relationship of the following terms: Git, GitHub, CI/CD, GitHub Actions, Version Control.***


```{mermaid}
%%| fig-cap: 'Git, GitHub, CI/CD, Actions, & Version Control'
%%| fig-width: 7.0
%%| fig-align: center
%%{init: {'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"18px"}}}%%

mindmap
  root((Code<br>Management))
    Version Control
      Tracks changes over time
      Enables rollback
      Supports collaboration
        Git
          Local repositories
          Branching and merging
          Commit history
            GitHub
              Remote repositories
              Pull requests
              Code review
                GitHubActions
                  Workflow files
                  Triggered by Git events
                    CICD
                      Continuous Integration
                        Automated testing
                        Build validation
                      Continuous Delivery
                        Automated deployment
                        Environment promotion

```



