# Lab 5: Deployments and Code Promotion {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

## Comprehension questions

### Data Science Environments 

> ***Write down a mental map of the relationship between the three environments for data science. Include the following terms: Git Promote, CI/CD, automation, deployment, dev, test, prod.***

### Git and Data Science 

> ***Why is Git so important to a good code promotion strategy? Can you have a code promotion strategy without Git***

### Git and CI/CD

> ***What is the relationship between Git and CI/CD? What’s the benefit of using Git and CI/CD together?***

### Git <> GitHub <> CI/CD <> GitHub Actions <> Version Control 

> ***Write out a mental map of the relationship of the following terms: Git, GitHub, CI/CD, GitHub Actions, Version Control.***

