# Lab 5: Deployments and Code Promotion {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

## Comprehension questions

### Data Science Environments 

> ***Write down a mental map of the relationship between the three environments for data science. Include the following terms: Git Promote, CI/CD, automation, deployment, dev, test, prod.***

The figures from the [chapter](https://do4ds.com/chapters/sec1/1-5-deployments.html#separate-the-prod-environment) do an excellent job illustrating these environments, but I've added a sequence diagram: 

```{=html}
<style>

.codeStyle span:not(.nodeLabel) {
  font-family: monospace;
  font-size: 1.5em;
  font-weight: bold;
  color: #d80808 !important;
  background-color: #ffffff;
  padding: 0.2em;
}
</style>
```

```{mermaid}
%%| fig-cap: 'Data Science Environments'
%%| fig-width: 7.0
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"18px"}}}%%


sequenceDiagram
    participant Dev as Dev
    participant Git as Git
    participant CI as CI/CD<br>Pipeline
    participant Test as Test
    participant Prod as Prod
    
    Dev->>Dev: Write code & test locally
    Dev->>Git: Commit & push changes
    Git->>CI: Trigger automation
    CI->>CI: Run tests & checks
    CI->>Test: Deploy to Test environment
    Test->>Test: Integration testing
    Test->>Test: User acceptance testing
    
    Note over Test,CI: Manual approval or<br/>automated promotion
    
    Test->>Git: Tag release (Git Promote)
    Git->>CI: Trigger production deployment
    CI->>CI: Build & validate
    CI->>Prod: Deploy to Production
    Prod->>Prod: Monitor & validate
    
    Note over Prod: Rollback<br>if needed
```

```{mermaid}
%%| fig-cap: 'Data Science Environments'
%%| fig-width: 7.0
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"18px"}}}%%

flowchart TB
    Dev[Development Environment]
    Git[Git Repository]
    CI[CI: Continuous Integration]
    Test[Test Environment]
    Promote[Git Promote]
    CD[CD: Continuous Deployment]
    Prod[Production Environment]
    
    Dev -->|commit & push| Git
    Git -->|webhook triggers| CI
    CI -->|tests pass| Test
    
    Test -->|validation complete| Promote
    Promote -->|create tag v1.0| Git
    Git -->|tag triggers| CD
    CD -->|deploy| Prod
    
    Prod -.->|rollback needed| Git
    Git -.->|deploy previous tag| CD
    
    style Dev fill:#e3f2fd,stroke:#1976d2,stroke-width:3px
    style Test fill:#fff3e0,stroke:#f57c00,stroke-width:3px
    style Prod fill:#e8f5e9,stroke:#388e3c,stroke-width:3px
    style Git fill:#fce4ec,stroke:#c2185b,stroke-width:2px
    style CI fill:#f3e5f5,stroke:#7b1fa2,stroke-width:2px
    style CD fill:#f3e5f5,stroke:#7b1fa2,stroke-width:2px
    style Promote fill:#fff9c4,stroke:#f9a825,stroke-width:3px
```

- The development environment (**Dev**) is where data scientists write and test code locally. 
    - The Dev environment is used for rapid iteration and experimentation, and it's connected to version control (Git).     
- The testing environment (**Test**) is a mirror of the production environment with realistic data (often anonymized). 
    - The Test environment is integrated and validates with other systems, and user acceptance testing occurs here.      
- The 'live' environment that serves the end users is the production environment (**Prod**). 
    - Prod has strictly controlled access with a focus on stability and monitoring.       
- **Git promotion** is the process of tagging a commit as ready for production. 
    - This creates an immutable reference point and triggers production deployment pipeline.    
- The **CI/CD Pipeline** is the automated process that builds, tests, and deploys the code. 
    - This ensures consistency across environments and reduces human error in deployment. CI/CD automates testing at each stage, deployment between environments, rollback on failures, and monitoring and alerting. 



### Git & Data Science 

> ***Why is Git so important to a good code promotion strategy?***

Git gives 

> ***Can you have a code promotion strategy without Git?***

### Git & CI/CD

> ***What is the relationship between Git and CI/CD?***

> ***What’s the benefit of using Git and CI/CD together?***

### Git, GitHub, CI/CD, Actions, & Version Control 

> ***Write out a mental map of the relationship of the following terms: Git, GitHub, CI/CD, GitHub Actions, Version Control.***



