# Lab 5: Deployments and Code Promotion {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

## Comprehension questions

The questions below come from the [Deployments and Code Promotion](https://do4ds.com/chapters/sec1/1-5-deployments.html) chapter. 

### Data Science Environments 

> ***Write down a mental map of the relationship between the three environments for data science. Include the following terms: Git Promote, CI/CD, automation, deployment, dev, test, prod.***

The figures from the [chapter](https://do4ds.com/chapters/sec1/1-5-deployments.html#separate-the-prod-environment) do an excellent job illustrating these environments, but I've added a sequence diagram: 

```{=html}
<style>

.codeStyle span:not(.nodeLabel) {
  font-family: monospace;
  font-size: 1.5em;
  font-weight: bold;
  color: #d80808 !important;
  background-color: #ffffff;
  padding: 0.2em;
}
</style>
```

```{mermaid}
%%| fig-cap: 'Data Science Environments'
%%| fig-width: 7.0
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"18px"}}}%%


sequenceDiagram
    participant Dev as Dev
    participant Git as Git
    participant CI as CI/CD<br>Pipeline
    participant Test as Test
    participant Prod as Prod
    
    Dev->>Dev: Write code & test locally
    Dev->>Git: Commit & push changes
    Git->>CI: Trigger automation
    CI->>CI: Run tests & checks
    CI->>Test: Deploy to Test environment
    Test->>Test: Integration testing
    Test->>Test: User acceptance testing
    
    Note over Test,CI: Manual approval or<br/>automated promotion
    
    Test->>Git: Tag release (Git Promote)
    Git->>CI: Trigger production deployment
    CI->>CI: Build & validate
    CI->>Prod: Deploy to Production
    Prod->>Prod: Monitor & validate
    
    Note over Prod: Rollback<br>if needed
```

#### DEV Environments 

The development environment (**Dev**) is the primary location where data scientists write and test their code. It's typically used for iteration and experimentation, and it's connected to version control (Git).  

#### Test Environments 

The testing environment (**Test**) is a mirror of the production environment with realistic data (often anonymized). Test environments are integrated and validates with other systems, and user acceptance testing occurs here.  

#### Production Environments 

The 'live' environment that serves the end users is the production environment (**Prod**). Prod has strictly controlled access with a focus on stability and monitoring.       

```{mermaid}
%%| fig-cap: 'Data Science Environments'
%%| fig-width: 7.0
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"18px"}}}%%

flowchart TB
    Dev[Development Environment]
    Git[Git Repository]
    CI[CI: Continuous Integration]
    Test[Test Environment]
    Promote[Git Promote]
    CD[CD: Continuous Deployment]
    Prod[Production Environment]
    
    Dev -->|commit & push| Git
    Git -->|webhook triggers| CI
    CI -->|tests pass| Test
    
    Test -->|validation complete| Promote
    Promote -->|create tag v1.0| Git
    Git -->|tag triggers| CD
    CD -->|deploy| Prod
    
    Prod -.->|rollback needed| Git
    Git -.->|deploy previous tag| CD
    
    style Dev fill:#e3f2fd,stroke:#1976d2,stroke-width:3px
    style Test fill:#fff3e0,stroke:#f57c00,stroke-width:3px
    style Prod fill:#e8f5e9,stroke:#388e3c,stroke-width:3px
    style Git fill:#fce4ec,stroke:#c2185b,stroke-width:2px
    style CI fill:#f3e5f5,stroke:#7b1fa2,stroke-width:2px
    style CD fill:#f3e5f5,stroke:#7b1fa2,stroke-width:2px
    style Promote fill:#fff9c4,stroke:#f9a825,stroke-width:3px
```


**Git promotion** is the process of tagging a commit as ready for production. This creates an immutable reference point and triggers production deployment pipeline. The **CI/CD Pipeline** is the automated process that builds, tests, and deploys the code. 

This ensures consistency across environments and reduces human error in deployment. CI/CD automates testing at each stage, deployment between environments, rollback on failures, and monitoring and alerting. 



### Git & Data Science 

> ***Why is Git so important to a good code promotion strategy?***

There is a canonical DevOps rule of thumb, 

*If a change isn’t represented as a Git commit, it didn’t happen.*

In mature CI-CD systems, Git defines the promotion workflow, not the pipeline. For example, if we have the following Git branching model:

* `feature`: new capability (e.g., new model, new preprocessing step, new data connector). Branch off develop or `main`, open pull request early, iterate.    
* `develop`: integration branch for ongoing work. `feature` branches merge here first; periodically promoted to `main` (release).    
* `staging`: mirrors production (`main`), but for final validation. `main` → `staging` (or vice versa) depending on release process. Used for end-to-end pipeline runs on near-prod data/infra.    
* `main`: production-ready code and (often) reproducible pipeline definitions. Only `merge` via pull request; every commit should pass tests, lint, data checks. Often corresponds to what’s deployed (or what can be deployed).

Changes are committed to `feature*` → merged to `develop` → merged to `main`:

```{mermaid}
%%| fig-cap: 'Typical CI-CD promotion pattern'
%%| fig-width: 7.0
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"18px"}}}%%

flowchart TB
  F("<code>feature/*</code>") -->|"pull request"| D("<code>develop</code>")
  D -->|release pull request| M("<code>main</code>")
  M -->|deploy| P["<strong>production</strong>"]
  M -->|promote| S("<code>staging</code>")
  S -->|approve| P
    
```


The pipeline simply *reacts to Git events*. The code promotion logic lives in the branches, tags, merge policies, and not in brittle pipeline scripts.

A core CI-CD principle:

> **Build once, promote many times.**

All commits are signed--we know who made the changes--and the `main` branch requires approval to merge (i.e., it's protected). This artifact immutability connects 'builds' to specific commit SHAs or by tagging releases (like `v1.4.1`).  

```{mermaid}
%%| fig-cap: 'Git artifact immutability'
%%| fig-width: 7.0
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"18px"}}}%%
gitGraph
    commit tag: "Initial commit" type: HIGHLIGHT
    branch test
    checkout test
    commit tag: "test-setup"

    branch feature
    checkout feature
    commit tag: "develop-feature"
    commit
    commit
    commit tag: "test-feature"

    checkout test
    merge feature tag: "merge"
    commit
    commit tag: "integration-tests"

    checkout main
    merge test id: "v1.4.1" tag: "Release" type: REVERSE
```

The same code is promoted across environments so we can be sure untested code never reaches production. 

Without Git, the code we need to be rebuilt in each environment. The tested code wouldn't necessarily be the deployed code, and environment drift would be inevitable.

Git is the gatekeeper for CI-CD policies and integrates directly with enforcement mechanisms.

> **If it’s not enforced in Git, it *will* be bypassed**

**Git enables environment parity**. In DevOps, **environment drift is the enemy**. Dev, test, and prod differ only by data and credentials, and changes are promoted, not re-created.

From a DevOps audit perspective, Git provides the code change/approval history, and the pipeline results tied to commits. Each release is also tagged. This makes it easy to answer the inevitable question, 

> **“Which commit introduced this behavior in production?”**

When production inevitably breaks, we can use Git to roll back by redeploying previous commits/tags. This avoids making ad-hoc changes in a production server. 

Git promotion mechanics are standardized, so pipelines become reusable templates. This turns CI-CD into a system, not a script. 

Git is like a **state machine**, controlling the flow of changes.

With Git-centered promotion:

* Git = **intent**
* Pipeline = **execution**
* Artifact = **immutable**
* Promotion = **merge or tag**


> ***Can you have a code promotion strategy without Git?***

Yes — you can do code promotion without Git, as long as you still have three things:

* Some way of creating 'versioned artifacts' (i.e., something immutable that can be promoted)   
* Blocks/gates requiring tests/approvals that must be passed before promotion.   
* Some kind of promotion mechanism to copy/move/retag the versioned artifact between environments.       

Without Git you still need substitutes for:

1. Diff / review: code review tooling in your platform, or document-style review on “release bundles”   
2. Rollback: keep old artifacts and switch the environment pointer back   
3. Traceability: manifests + checksums + approvals + build logs   
A good minimum is: artifact hash + signed manifest + CI logs + approval record.


### Git & CI/CD

> ***What is the relationship between Git and CI/CD?***

> ***What’s the benefit of using Git and CI/CD together?***

### Git, GitHub, CI/CD, Actions, & Version Control 

> ***Write out a mental map of the relationship of the following terms: Git, GitHub, CI/CD, GitHub Actions, Version Control.***



