# <em>Linux Users</em> {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: false
co_box(
  color = "o",
  look = "default",
  hsize = "1.25",
  size = "1.00",
  header = "Caution",
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

## Log in

Start by logging into the AWS EC2 instance. You can refer back to the [AWS Server Login](aws-ssh-ec2.qmd) and [Managing SSH key pairs](manage-ssh-keys.qmd) chapters if you need a refresher.

### SSH commands

The commands below connect to the EC2 instance as `ec2-user` with verbose output:

```{bash}
#| eval: false
#| code-fold: false
ssh -v -i ~/.ssh/your-key.pem ec2-user@<ec2-public-ip>
```

The `-i` flag suggests an identity file, but SSH will still try keys from your `ssh-agent` first. If the agent has a key the server accepts, that key wins and the `pem` is never used.

Or you can force SSH to authenticate using only the specified `pem` file:[^force-ssh-pem]

```{bash}
#| eval: false
#| code-fold: false
ssh -v -i ~/.ssh/your-key.pem -o IdentitiesOnly=yes -o IdentityAgent=none ec2-user@<ec2-public-ip>
```

- `IdentitiesOnly=yes` restricts SSH to the key given by `-i`, ignoring other keys it might otherwise try.

- `IdentityAgent=none` disables the `ssh-agent` for this session, preventing agent-held keys from being offered.

[^force-ssh-pem]: *Use this when you need to verify a specific key works, or when the agent would otherwise shadow the intended key.*

After logging in, you'll see something like this:

```{verbatim}
Last login: Fri Jul 17 17:42:40 2026 from ip<your-ip-address.net>
   ,     #_
   ~\_  ####_        Amazon Linux 2
  ~~  \_#####\
  ~~     \###|       AL2 End of Life is 2026-06-30.
  ~~       \#/ ___
   ~~       V~' '->
    ~~~         /    A newer version of Amazon Linux is available!
      ~~._.   _/
         _/ _/       Amazon Linux 2023, GA and supported until 2029-06-30.
       _/m/'           https://aws.amazon.com/linux/amazon-linux-2023/

21 package(s) needed for security, out of 24 available
Run "sudo yum update" to apply all updates.
[ec2-user@ip-your-ip-address]$
```

Be sure to run `sudo yum update` to keep all the packages updated.

## The bash shell

There are entire books written on using the command line, so I'll just cover the commands I use the most often (or find to be the most useful). The most basic of these is `whoami`, which reports the account the shell is running as:

```{bash}
#| eval: false
#| code-fold: false
whoami
```

```{verbatim}
ec2-user
```

### Where am I, what's here

Right after logging in, get your bearings:

```{bash}
#| eval: false
#| code-fold: false
pwd
```

```{verbatim}
/home/ec2-user
```

List everything, including hidden files, with permissions:

```{bash}
#| eval: false
#| code-fold: false
ls -la
```

```{verbatim}
total 16
drwx------ 4 ec2-user ec2-user 110 Jul 17 19:09 .
drwxr-xr-x 3 root     root      22 Jul  8 13:28 ..
-rw------- 1 ec2-user ec2-user 566 Aug  2 07:45 .bash_history
-rw-r--r-- 1 ec2-user ec2-user  18 Jul 15  2020 .bash_logout
-rw-r--r-- 1 ec2-user ec2-user 193 Jul 15  2020 .bash_profile
-rw-r--r-- 1 ec2-user ec2-user 231 Jul 15  2020 .bashrc
drwx------ 3 ec2-user ec2-user  18 Jul 17 19:09 .config
drwx------ 2 ec2-user ec2-user  29 Jul  8 13:28 .ssh
```

`ec2-user` owns everything here, and `.ssh` is already locked down to `700` before I've touched it. That's Amazon Linux 2's default, not something I set.

### Who am I, what can I do

`id` shows the groups behind the `sudo` access I've been using without thinking about it:

```{bash}
#| eval: false
#| code-fold: false
id
```

```{verbatim}
uid=1000(ec2-user) gid=1000(ec2-user) groups=1000(ec2-user),4(adm),10(wheel),190(systemd-journal)
```

`wheel` is the group that grants `sudo` on Amazon Linux 2. Ubuntu calls the equivalent group `sudo` - same idea, different name. If you're ever unsure which convention an AMI uses, `id` will tell you. I'll use `wheel` again below when I create a second user.

### Checking disk space

Worth doing right after logging into any new instance, before you forget:

```{bash}
#| eval: false
#| code-fold: false
df -h
```

```{verbatim}
Filesystem      Size  Used Avail Use% Mounted on
devtmpfs        462M     0  462M   0% /dev
tmpfs           471M     0  471M   0% /dev/shm
tmpfs           471M  428K  470M   1% /run
tmpfs           471M     0  471M   0% /sys/fs/cgroup
/dev/nvme0n1p1  8.0G  2.6G  5.5G  32% /
tmpfs            95M     0   95M   0% /run/user/1000
```

The default EC2 root volume is small. Installing R packages or pulling a Docker image is usually what finally runs it out.

### Piping output

A pipe (`|`) feeds the output of one command into the next. Counting files in `/etc` without opening it:

```{bash}
#| eval: false
#| code-fold: false
ls /etc | wc -l
```

```{verbatim}
191
```

### Reading the manual

Every command's flags are documented with `man`. Before creating a user below, it's worth checking what `useradd` does on this particular AMI, since its defaults aren't the same on every distro:

```{bash}
#| eval: false
#| code-fold: false
man useradd
```

Arrow keys scroll, `q` quits back to the shell.

## Create a non-root user

Everything so far has run as `ec2-user`, which already has `sudo`. That's fine for completing the labs in DO4DS, but on a real life shared server we'd want a personal account for daily work, with `sudo` granted on purpose (instead of handed out by default).

Create the account:

```{bash}
#| eval: false
#| code-fold: false
sudo useradd test-user
```

Unlike Ubuntu's `adduser`, Amazon Linux 2's `useradd` doesn't prompt for anything - it just creates the account and home directory silently. 

Set a password separately:

```{bash}
#| eval: false
#| code-fold: false
sudo passwd test-user
```

```{verbatim}
Changing password for user test-user.
New password:
Retype new password:
passwd: all authentication tokens updated successfully.
```

Add `test-user` to `wheel` so has `sudo`:

```{bash}
#| eval: false
#| code-fold: false
sudo usermod -aG wheel test-user
```

`-aG` appends the group without disturbing any groups the user already belongs to - leaving off `-a` would replace them. 

Confirm it took:

```{bash}
#| eval: false
#| code-fold: false
id test-user
```

```{verbatim}
uid=1001(test-user) gid=1001(test-user) groups=1001(test-user),10(wheel)
```

## Add an SSH key for your new user

`test-user` has a password now, but password authentication over SSH is disabled on this instance, same as it is for `ec2-user`. The account needs its own `.ssh` directory and an `authorized_keys` file before it can log in at all.

I already have a public key trusted for `ec2-user` (from [Managing SSH key pairs](manage-ssh-keys.qmd)), so the fastest path is reusing it rather than generating a new pair just for this exercise:

First I create a `.ssh/` folder in `test-user`'s home directory and copy over `ec2-user`'s `authorized_keys`:

```{bash}
#| eval: false
#| code-fold: false
sudo mkdir -p /home/test-user/.ssh
sudo cp ~/.ssh/authorized_keys /home/test-user/.ssh/authorized_keys
```

Both commands ran via `sudo`, so the new files belong to `root`, not `test-user`. Hand them over and lock them down:

```{bash}
#| eval: false
#| code-fold: false
# change ownership
sudo chown -R test-user:test-user /home/test-user/.ssh
# change permissions
sudo chmod 700 /home/test-user/.ssh
sudo chmod 600 /home/test-user/.ssh/authorized_keys
```

```{r}
#| label: co_box_chown_chmod
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "r",
  look = "default",
  hsize = "1.15",
  size = "1.00",
  header = "Ownership & Permissions",
  fold = TRUE,
  contents = "
`chown` and `chmod` solve two different problems, and both showed up in the commands above.

- **`chown user:group file`** changes *who owns* a file or directory - the user and the group. `sudo chown -R test-user:test-user /home/test-user/.ssh` handed the `.ssh` folder from `root` (who created it via `sudo`) to `test-user`, recursively (`-R`) so the `authorized_keys` file inside came along with it.

- **`chmod mode file`** changes *what the owner, group, and everyone else are allowed to do* with a file - read, write, execute. `chmod 700` restricts a directory to the owner only (read/write/execute); `chmod 600` restricts a file to owner read/write, no execute. SSH enforces these exact permissions on `.ssh` and `authorized_keys` and silently refuses to use anything looser.

In short: `chown` decides who the file belongs to, `chmod` decides what they (and everyone else) can do with it.
  "
)
```

Same rule as everywhere else with SSH: a `.ssh` directory or key file that's group- or world-readable gets silently ignored.

```{mermaid}
%%| fig-cap: 'One Trusted Key, Two Local Accounts'
%%| fig-width: 8.0
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}}%%

graph LR
    subgraph Laptop["Your Computer"]
      subgraph priv["Private Key"]
        pkey["<strong>~/.ssh/your-key-name</strong>"]
      end
    end
    subgraph EC2["EC2 Instance"]
        ecuser["ec2-user<br/>authorized_keys"]
        testuser["test-user<br/>authorized_keys<br/>(<em>copied</em>)"]
    end
    pkey -->|"<strong>ssh ec2-user@...</strong>"| ecuser
    pkey -->|"<strong>ssh test-user@...</strong>"| testuser

    style pkey font-family:monospace
    style Laptop fill:#E8E8E8,stroke:#333,stroke-width:2px,color:#000
    style EC2 fill:#1B2A41,stroke:#fff,stroke-width:2px,color:#fff
    style priv fill:#D2562B,stroke:#fff,stroke-width:1px,color:#fff
    style ecuser fill:#2A6F77,stroke:#fff,stroke-width:1px,color:#fff,font-family:monospace
    style testuser fill:#2A6F77,stroke:#fff,stroke-width:1px,color:#fff,font-family:monospace
```

The same private key on my computer now opens two separate accounts, because SSH only checks whether the key I present matches a line in the `authorized_keys` of the account I'm connecting to.

From a new terminal (leave the `ec2-user` session open in case something's wrong), test the new account:

```{bash}
#| eval: false
#| code-fold: false
ssh -i ~/.ssh/your-key-name test-user@<ec2-public-ip>
```

If successful, you'll see the following:

```{verbatim}
   ,     #_
   ~\_  ####_        Amazon Linux 2
  ~~  \_#####\
  ~~     \###|       AL2 End of Life is 2026-06-30.
  ~~       \#/ ___
   ~~       V~' '->
    ~~~         /    A newer version of Amazon Linux is available!
      ~~._.   _/
         _/ _/       Amazon Linux 2023, GA and supported until 2029-06-30.
       _/m/'           https://aws.amazon.com/linux/amazon-linux-2023/
       
[test-user@ip-your-ip-address ~]$
```

Confirm `sudo` actually works for this account:

```{bash}
#| eval: false
#| code-fold: false
whoami
```

```{verbatim}
test-user
```

```{bash}
#| eval: false
#| code-fold: false
sudo whoami
```

```{verbatim}
We trust you have received the usual lecture from the local System
Administrator. It usually boils down to these three things:

    #1) Respect the privacy of others.
    #2) Think before you type.
    #3) With great power comes great responsibility.

[sudo] password for test-user:
root
```

## Recap

1. `wheel` (not `sudo`) is the group that grants admin rights on Amazon Linux 2 - check with `id` if you're unsure which convention an AMI uses.

2. SSH doesn't care which account is logging in, only whether the presented private key matches a line in that account's `authorized_keys`. Copying a trusted key to a new account is enough; a new key pair isn't required.
