# <em>R processing layer</em> {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

This document translates the contents of lab 2 in R. The source Quarto document can be found in [`_labs/lab02/R/model-vetiver-r.qmd`](https://github.com/mjfrigaard/do4ds-labs/blob/main/_labs/lab02/R/model-vetiver-r.qmd).

```{r}
#| collapse: true
#| eval: false
fs::dir_tree("_labs/lab02/R/")
## _labs/lab02/R/
## └── model-vetiver-r.qmd
```

## Load packages 

```{r}
#| message: false 
#| warning: false
#| eval: true
library(palmerpenguins)
library(plumber)
library(dplyr)
library(duckdb)
library(fastDummies)
library(vetiver)
library(pins)
```


## Get Data

Load the `palmerpenguins` data, create a connection with `dbConnect()`, register the R data frame with `duckdb`, create the `penguins` table: 

```{r}
#| eval: false
penguins_data <- palmerpenguins::penguins

con <- dbConnect(duckdb::duckdb(), "my-db.duckdb")

duckdb_register(con, "penguins_data", penguins_data)

dbExecute(con, "CREATE OR REPLACE TABLE penguins AS SELECT * FROM penguins_data")
```

Omit missing, disconnect from DB: 

```{r}
#| eval: false
df <- dbGetQuery(con, "SELECT * FROM penguins") |>
  na.omit()

dbDisconnect(con)
```

View top rows of `df`:

```{r}
#| eval: false
head(df, 3)
```

## Define Model and Fit

Create dummy variables (one-hot encoding) and fit linear regression model.

```{r}
#| eval: false
X <- df |>
  select(bill_length_mm, species, sex) |>
  dummy_cols(
    select_columns = c("species", "sex"), 
    remove_first_dummy = TRUE) |>
  select(-species, -sex)

y <- df$body_mass_g

model <- lm(y ~ ., data = X)
```

## Get some information

```{r}
#| eval: false
r_squared <- summary(model)$r.squared
intercept <- coef(model)[1]
coefficients <- coef(model)[-1]
column_names <- names(X)
```

```{r}
#| eval: false
cat("R^2:", r_squared, "\n")
cat("Intercept:", intercept, "\n")
cat("Columns:", paste(column_names, collapse = ", "), "\n")
cat("Coefficients:", paste(coefficients, collapse = ", "), "\n")
```

## Turn into Vetiver Model

```{r}
#| eval: false
#| message: false 
#| warning: false
v <- vetiver_model(model, model_name = "penguin_model")
```

Below we can see the model information stored in the `models/` folder:

```{r}
#| collapse: true
fs::dir_tree("_labs/lab02/R/models/")
```

## Save to Board

```{r}
#| eval: false
model_board <- board_folder("./models")
vetiver_pin_write(model_board, v)
```

## Turn model into API

Unlike Python, we'll need to create a plumber router from `plumber::pr()` to pass to the `vetiver_api()` (along with the deployable `vetiver_model()` object). 

```{r}
#| eval: false
app <- pr() |> 
  vetiver_api(v)
```
