# <em>Python processing layer</em> {#sec-model-vetiver}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

Below is the code I included in `_labs/lab02/model-vetiver/model-vetiver.qmd`: 

```{r}
#| eval: false
#| results: hide
#| message: false 
#| warning: false
library(reticulate)
use_virtualenv("myenv", required = TRUE)
virtualenv_install(
  envname = "myenv", 
  packages = c("palmerpenguins", "pandas", "numpy", 
               "scikit-learn", "duckdb", "vetiver", 
                "pins"))
```

```{python}
#| eval: false
from palmerpenguins import load_penguins #<1>
from pandas import get_dummies
import numpy as np
from sklearn.linear_model import LinearRegression
from sklearn import preprocessing
import duckdb
```
1. Added this for the `penguins` data 

## Get Data

```{python}
#| eval: false
penguins_data = load_penguins() #<1>
con = duckdb.connect('my-db.duckdb')
con.execute("CREATE OR REPLACE TABLE penguins AS SELECT * FROM penguins_data") #<2>
df = con.execute("SELECT * FROM penguins").fetchdf().dropna()
con.close()

df.head(3)
```
1. Load the data for the `duckdb`   
2. Create the `penguins` table in db    

## Define Model and Fit

```{python}
#| eval: false
X = get_dummies(df[['bill_length_mm', 'species', 'sex']], drop_first = True)
y = df['body_mass_g']

model = LinearRegression().fit(X, y)
```

## Get some information

```{python}
#| eval: false
print(f"R^2 {model.score(X,y)}")
print(f"Intercept {model.intercept_}")
print(f"Columns {X.columns}")
print(f"Coefficients {model.coef_}")
```

## Turn into Vetiver Model

```{python}
#| eval: false
from vetiver import VetiverModel
v = VetiverModel(model, model_name='penguin_model', prototype_data=X)
```

## Save to Board

```{python}
#| eval: false
import os #<1>
from pins import board_folder
from vetiver import vetiver_pin_write

model_board = board_folder("./models", allow_pickle_read=True) #<2>

vetiver_pin_write(model_board, v)
```
1. For creating folder    
2. Use the current working directory    

## Turn model into API

```{python}
#| eval: false
from vetiver import VetiverAPI
app = VetiverAPI(v, check_prototype = True)
```

The original code from lab 2 is below:

<details>
  <summary><strong>View model-vetiver.qmd</strong></summary>
  
<br>

```{r}
#| label: print-model-vetiver-qmd
#| echo: false
library(readr)
readr::read_lines("_labs/lab02/model-vetiver.qmd") |> 
  base::noquote() |> 
  writeLines()
```

</details>
