# Plumber vs. Plumber2 Syntax

```{r}
#| eval: true 
#| echo: false 
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_dev
#| echo: false
#| results: asis
#| eval: false
co_box(
  color = "r", 
  look = "minimal",
  header = "Alert",
  contents = "The contents for section are being revised. Thank you for your patience."
)
```

## Overview

`plumber` is the classic way to turn R functions into web APIs using decorators (`#*`). `plumber2` represents a more modern, type-safe approach that leverages R function signatures and schema-driven design to improve robustness and documentation.

## Syntax Comparison

### 1. Basic Routing

**Plumber (Decorator style)**

In `plumber`, you use special comment decorators to define the HTTP method, path, and parameters.

```r
#* @get /hello
#* @param name The name to greet
function(name = "World") {
  list(message = paste("Hello", name))
}
```

**Plumber2 (Signature style)**

`plumber2` focuses on the function definition itself, moving away from heavy reliance on comments and toward more explicit type-driven signatures.

```r
# Using plumber2, the function signature and 
# type hints (if available) guide the API structure.
hello <- function(name: character = "World") {
  list(message = paste("Hello", name))
}

# The routing is often handled by passing these 
# functions to a plumber2 router object.
```


### 2. Data Validation

**Plumber**

Validation in standard `plumber` is typically handled imperatively inside the function body using conditional logic.

```r
#* @post /predict
function(data) {
  if (!is.data.frame(data)) {
    stop("Input must be a data frame")
  }
  # ... prediction logic
}
```

**Plumber2**

`plumber2` encourages declarative validation, where the expected structure and types are defined upfront, often allowing the framework to reject invalid requests before the function is even called.

```r
# plumber2 leverages type annotations or 
# schema definitions to validate input automatically.
predict <- function(data: data.frame) {
  # The framework ensures 'data' is a data.frame
  # before this code even executes.
  model_predict(data)
}
```


### 3. Summary Table

| Feature | `plumber` | `plumber2` |
| :--- | :--- | :--- |
| **Primary Mechanism** | Decorators (`#*` comments) | Function Signatures & Schemas |
| **Type Safety** | Loose; relies on manual checks | Strong; leverages R type hints |
| **Input Validation** | Imperative (inside the function) | Declarative (via schema/signature) |
| **API Documentation**| Swagger/OpenAPI via comments | OpenAPI via type/schema inspection |
