# <em>`plumber2`</em> {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "r",
  look = "default",
  hsize = "1.25",
  size = "1.00",
  header = "Caution",
  fold = FALSE,
  contents = "This section is being developed. Thank you for your patience."
)
```

This chapter covers building and consuming an R API using [`plumber2`](https://plumber2.posit.co/) (when do4ds was written, the original [`plumber`](https://www.rplumber.io/) package was used). We'll build the same penguin body mass prediction API from [lab 3](lab03.qmd) and connect it to a Shiny application, but update the API to use `plumber2`'s syntax.

```{r}
#| eval: false
#| code-fold: false
install.packages("plumber2")
```

## APIs (refresher)

APIs (Application Programming Interfaces) are covered in [lab 3](lab03.qmd). They work like placing an order in a restaurant:

```{=html}
<style>
.codeStyle span:not(.nodeLabel) {
  font-family: monospace;
  font-size: 1.5em;
  font-weight: bold;
  color: #d80808 !important;
  background-color: #ffffff;
  padding: 0.2em;
}
</style>
```

```{mermaid}
%%| fig-cap: 'Refresher on APIs'
%%| fig-width: 7.0
%%| fig-align: center
%%{init: {'theme': 'base', 'look': 'handDrawn', 'themeVariables': {'fontFamily': 'monospace'}}}%%

sequenceDiagram
    participant Customer as Shiny App
    participant Waiter as API
    participant Kitchen as ML Model

    Customer->>Waiter: "I'd like a<br>prediction for<br>bill_length=45,<br>species=Adelie"
    Waiter->>Kitchen: Prepare data &<br>make prediction
    Kitchen->>Waiter: Prediction: 4180.8 grams
    Waiter->>Customer: Returns prediction
    style Customer fill:#F2E6C2,stroke:#2A6F77,color:#1B2A41
    style Waiter fill:#F2E6C2,stroke:#2A6F77,color:#1B2A41
    style Kitchen fill:#F2E6C2,stroke:#2A6F77,color:#1B2A41
```

The process involves three steps:

1. The Shiny app asks for predictions.
2. The `plumber2` API takes the values selected in the UI (as requests), and returns predictions (as responses).
3. ML Model does the *actual* work of calculating the predictions.

## Overview

The updated `plumber2` files are in the following folders:


**`api/`**

```{bash}
#| eval: false
#| code-fold: false
R/api/
├── api.Rproj
├── model.R     #<1>
├── models/
│   └── penguin_model/
├── my-db.duckdb
├── plumber.R   #<2>
├── README.md
├── renv/
└── renv.lock
```
1. `model.R` builds and stores the model used for predictions.
2. `plumber.R` builds and launches the `plumber2` API.

**`app/`**

```{bash}
#| eval: false
#| code-fold: false
R/app/
├── app-api.R   #<1>
├── app.Rproj
├── renv/
└── renv.lock
```
1. `app-api.R` is the Shiny application that sends requests to the API.

## The model (`model.R`)

The `model.R` file contains the packages and functions to build the linear model used to generate predictions from our Shiny app.

```{r}
#| eval: false
#| code-fold: show
#| code-summary: 'show/hide model.R'

library(palmerpenguins) #<1>
library(duckdb)
library(DBI)
library(dplyr)
library(vetiver)
library(pins) #<1>

con <- DBI::dbConnect(duckdb::duckdb(), "my-db.duckdb") #<2>

duckdb::duckdb_register(con, "penguins_raw", palmerpenguins::penguins) #<3>

DBI::dbExecute( #<4>
  con,
  "CREATE OR REPLACE TABLE penguins AS SELECT * FROM penguins_raw"
) #<4>

df <- DBI::dbGetQuery( #<5>
  con,
  "SELECT bill_length_mm, species, sex, body_mass_g
   FROM penguins
   WHERE body_mass_g IS NOT NULL
   AND bill_length_mm BETWEEN 30 AND 60
   AND sex IS NOT NULL
   AND species IS NOT NULL"
) #<5>

DBI::dbDisconnect(con) #<6>

model <- lm(body_mass_g ~ bill_length_mm + species + sex, data = df) #<7>

v <- vetiver::vetiver_model( #<8>
  model,
  model_name = "penguin_model",
  description = "Linear model predicting penguin body mass from bill length, species, and sex",
  save_prototype = TRUE #<9>
) #<8>

model_board <- pins::board_folder("./models") #<10>
vetiver::vetiver_pin_write(model_board, v) #<10>
```
1. Load packages.
2. Connect to `duckdb`.
3. Register the penguins `data.frame` with `duckdb`.
4. Create a persistent table.
5. Query and filter penguins data.
6. Disconnect from database.
7. Train the model using categorical variables directly (R handles dummy variable creation).
8. Create `vetiver` model.
9. Save the expected input format.
10. Write model to board.

```{mermaid}
%%| fig-width: 7.0
%%| fig-align: center
%%{init: {'theme': 'base', 'look': 'handDrawn', 'themeVariables': {'fontFamily': 'monospace'}}}%%

graph LR
    subgraph Training["<strong>Model Training</strong>"]
        ModelR("<strong>model.R</strong>")
        DuckDB("DuckDB <br> (as <strong>my-db.duckdb</strong>)")
        PinBoard("Pin Board<br><strong>models/penguin_model/</strong>")

        ModelR -->|"<em>loads data from</em>"| DuckDB
        ModelR -->|"<em>trains lm() model</em>"| LM("Linear Model")
        LM -->|"<em>wrapped by</em>"| Vetiver("Vetiver Model")
        Vetiver -->|"<em>saved to</em>"| PinBoard
    end
    classDef default fill:#F2E6C2,stroke:#2A6F77,color:#1B2A41
    style Training fill:#d4e7e4,stroke:#2A6F77,color:#1B2A41
```

Running `model.R` creates the `models/` directory and `duckdb` database file:

```{verbatim}
├── models
│   └── penguin_model
│       └── 20251005T062750Z-74445
│           ├── data.txt
│           └── penguin_model.rds
└── my-db.duckdb
```

The `prototype` stored in the `vetiver` model contains the expected input format, including column names, data types, and factor levels:

```{r}
#| eval: false
#| code-fold: false
v <- pins::board_folder("models/") |> vetiver::vetiver_pin_read("penguin_model")
str(v$prototype)
# Classes 'tbl_df', 'tbl' and 'data.frame':	0 obs. of  3 variables:
#  $ bill_length_mm: num
#  $ species       : Factor w/ 3 levels "Adelie","Chinstrap",..:
#  $ sex           : Factor w/ 2 levels "female","male":
```

## The API (`plumber.R`) {#sec-api}

The `plumber.R` file defines handler functions and assembles a `plumber2` router.

### plumber vs plumber2

`plumber2` is a complete rewrite and uses different function names and handler signatures:

| Task | `plumber` | `plumber2` |
|------|-----------|------------|
| Create router | `pr()` | `api()` |
| Add GET route | `pr_get()` | `api_get()` |
| Add POST route | `pr_post()` | `api_post()` |
| Start server | `pr_run()` | `api_run()` |
| JSON serializer | `serializer_json()` | `format_json()` |
| Request body | `jsonlite::fromJSON(req$postBody)` | `body` argument |
| Set response status | `res$status <- 500` | `response$status <- 500L` |

The most important change is in handler function signatures. In `plumber`, query parameters, body parameters, and path parameters are all mixed together as named function arguments, and the raw request body must be parsed manually. 

In `plumber2`, they are separated:

- Path parameters: come as named arguments (e.g., `/users/<id>` → `function(id)`)
- Query parameters: accessed via the `query` argument
- Request body: accessed via the `body` argument (already parsed from JSON)

### Load model

```{r}
#| eval: false
#| code-fold: false
library(vetiver) #<1>
library(pins)
library(plumber2) #<1>

model_board <- pins::board_folder("models/") #<2>

v <- vetiver::vetiver_pin_read(model_board, "penguin_model") #<3>
```
1. Packages.
2. Connect to model board.
3. Read pinned `vetiver` model.

### Prepare prediction data

JSON doesn't have factors—only strings. The `prep_pred_data()` helper bridges the gap by converting `species` and `sex` strings to factors using the levels stored in the `vetiver` model prototype.

```{r}
#| eval: false
#| code-fold: false
prep_pred_data <- function(input_data) {
  species_levels <- levels(v$prototype$species)
  sex_levels <- levels(v$prototype$sex)

  data.frame(
    bill_length_mm = as.numeric(input_data$bill_length_mm),
    species = factor(input_data$species, levels = species_levels),
    sex = factor(input_data$sex, levels = sex_levels),
    stringsAsFactors = FALSE
  )
}
```

```{mermaid}
%%| fig-width: 7.0
%%| fig-align: center
%%{init: {'theme': 'base', 'look': 'handDrawn', 'themeVariables': {'fontFamily': 'monospace'}}}%%

graph TB
    subgraph JSON["JSON from Shiny"]
        J1("species: 'Adelie'<br>(string)")
        J2("sex: 'male'<br>(string)")
    end

    subgraph Model["Model Expects"]
        M1("species: factor<br>(with levels)")
        M2("sex: factor<br>(with levels)")
    end

    subgraph Helper["Helper Function"]
        H("prep_pred_data()")
    end

    J1 & J2 --> Helper
    Helper --> M1 & M2
    classDef default fill:#F2E6C2,stroke:#2A6F77,color:#1B2A41
    style JSON fill:#d4e7e4,stroke:#2A6F77,color:#1B2A41
    style Model fill:#d4e7e4,stroke:#2A6F77,color:#1B2A41
    style Helper fill:#d4e7e4,stroke:#2A6F77,color:#1B2A41
```

### Handler functions

Each handler specializes in a different type of request. We'll organize these into four categories: health, model info, documentation, and predictions.

The `#*` annotation comments define the HTTP method, path, and serializer for each handler when the file is loaded by `plumber2`.

#### Health

##### **`/ping`**

A minimal health check that returns a status and timestamp:

```{r}
#| eval: false
#| code-fold: show
#| code-summary: 'show/hide handle_ping()'
#* Basic health check
#*
#* @get /ping
#* @serializer json
handle_ping <- function() {
  list(
    status = "alive",
    timestamp = Sys.time()
  )
}
```

Test with `curl`:

```{verbatim}
curl http://127.0.0.1:8080/ping
```

```json
{
  "status": ["alive"],
  "timestamp": ["2025-10-08 08:31:51"]
}
```

##### **`/health`**

An extended health check that includes model metadata and R version:

```{r}
#| eval: false
#| code-fold: show
#| code-summary: 'show/hide handle_health()'
#* Detailed health check
#*
#* @get /health
#* @serializer json
handle_health <- function() {
  list(
    status = "healthy",
    timestamp = Sys.time(),
    model_name = v$model_name,
    model_version = v$metadata$version,
    r_version = R.version.string
  )
}
```

```{verbatim}
curl http://127.0.0.1:8080/health
```

```json
{
  "status": ["healthy"],
  "timestamp": ["2025-10-08 08:38:29"],
  "model_name": ["penguin_model"],
  "model_version": ["20251005T062750Z-74445"],
  "r_version": ["R version 4.5.1 (2025-06-13)"]
}
```

#### Model Info

##### **`/model-prototype`**

Returns the expected input format for all prediction requests:

```{r}
#| eval: false
#| code-fold: show
#| code-summary: 'show/hide handle_model_prototype()'
#* Get model prototype information
#*
#* @get /model-prototype
#* @serializer json
handle_model_prototype <- function() {
  list(
    prototype = list(
      bill_length_mm = "numeric",
      species = list(
        type = "factor",
        levels = levels(v$prototype$species)
      ),
      sex = list(
        type = "factor",
        levels = levels(v$prototype$sex)
      )
    ),
    model_class = class(v$model)
  )
}
```

```{verbatim}
curl http://127.0.0.1:8080/model-prototype
```

<details>
<summary>show/hide JSON response from model-prototype</summary>

```json
{
  "prototype": {
    "bill_length_mm": ["numeric"],
    "species": {
      "type": ["factor"],
      "levels": ["Adelie", "Chinstrap", "Gentoo"]
    },
    "sex": {
      "type": ["factor"],
      "levels": ["female", "male"]
    }
  },
  "model_class": ["butchered_lm", "lm"]
}
```

</details>

##### **`/model-info`**

Returns comprehensive metadata about the deployed model:

```{r}
#| eval: false
#| code-fold: true
#| code-summary: 'show/hide handle_model_info()'
#* Get model information and metadata
#*
#* @get /model-info
#* @serializer json
handle_model_info <- function() {
  list(
    model_name = v$model_name,
    model_class = class(v$model)[1],
    version = v$metadata$version,
    created = v$metadata$created,
    required_pkgs = v$metadata$required_pkgs,
    description = v$description %||% "No description available"
  )
}
```

<details>
<summary>show/hide JSON response from model-info</summary>

```json
{
  "model_name": ["penguin_model"],
  "model_class": ["butchered_lm"],
  "version": ["20251005T062750Z-74445"],
  "created": {},
  "required_pkgs": {},
  "description": ["Linear model predicting penguin body mass from bill length, species, and sex"]
}
```

</details>

#### Documentation

##### **`/input-schema`**

Returns the field specifications and an example request:

```{r}
#| eval: false
#| code-fold: show
#| code-summary: 'show/hide handle_input_schema()'
#* Get input schema and example
#*
#* @get /input-schema
#* @serializer json
handle_input_schema <- function() {
  list(
    required_fields = list(
      bill_length_mm = list(
        type = "numeric",
        description = "Bill length in millimeters",
        range = c(30, 60)
      ),
      species = list(
        type = "string (converted to factor)",
        description = "Penguin species",
        valid_values = levels(v$prototype$species)
      ),
      sex = list(
        type = "string (converted to factor)",
        description = "Penguin sex",
        valid_values = levels(v$prototype$sex)
      )
    ),
    example = list(
      bill_length_mm = 45.5,
      species = "Gentoo",
      sex = "male"
    )
  )
}
```

#### Predictions

##### **`/predict`**

The main prediction endpoint. In `plumber2`, the `body` argument replaces the `req$postBody` + `jsonlite::fromJSON()` pattern from `plumber`—`plumber2` parses the JSON request body automatically.

```{r}
#| eval: false
#| code-fold: show
#| code-summary: 'show/hide handle_predict()'
#* Predict penguin body mass
#*
#* @post /predict
#* @serializer json
handle_predict <- function(body, response) { #<1>

  result <- tryCatch({

    if (is.list(body) && !is.data.frame(body)) { #<2>
      body <- as.data.frame(body)
    } #<2>

    pred_data <- prep_pred_data(body) #<3>

    prediction <- predict(v, pred_data) #<4>

    if (is.data.frame(prediction) && ".pred" %in% names(prediction)) { #<5>
      list(.pred = prediction$.pred)
    } else {
      list(.pred = as.numeric(prediction))
    } #<5>

  }, error = function(e) { #<6>
    response$status <- 500L
    list(
      error = conditionMessage(e),
      timestamp = as.character(Sys.time())
    )
  }) #<6>

  return(result)
}
```
1. `body` is parsed from JSON automatically by `plumber2`; `response` is used to set status codes.
2. Handle both single prediction and batch.
3. Convert strings to factors.
4. Make prediction.
5. Handle different return types from `vetiver`.
6. Error handling — `response$status <- 500L` replaces `res$status <- 500` from `plumber`.

```{mermaid}
%%| fig-width: 7.0
%%| fig-align: center
%%{init: {'theme': 'base', 'look': 'handDrawn', 'themeVariables': {'fontFamily': 'monospace'}}}%%

graph TB
    JSON("JSON Body: <br>'{'bill_length_mm':45,...}'")

    JSON -->|"<em>parsed by plumber2</em>"| Body("R List:<br>list(bill_length_mm=45,...)")

    Body -->|"<em>as.data.frame()</em>"| DF("Data Frame:<br>1 row × 3 columns")

    DF -->|"<em>prep_pred_data()</em>"| Factors("Proper Types:<br>numeric + factors")

    Factors -->|"<em>predict()</em>"| Result("Number:<br>4180.797")

    Result -->|"<em>list()</em>"| Response("{'.pred':[4180.797]}")
    classDef default fill:#F2E6C2,stroke:#2A6F77,color:#1B2A41
```

```{verbatim}
curl -X POST http://127.0.0.1:8080/predict \
  -H "Content-Type: application/json" \
  -d '{"bill_length_mm": 45, "species": "Adelie", "sex": "male"}'
```

```json
{
  ".pred": [4180.7965]
}
```

##### **`/predict-validated`**

Validates all inputs before making a prediction and returns descriptive error messages on failure:

```{r}
#| eval: false
#| code-fold: true
#| code-summary: 'show/hide handle_predict_validated()'
#* Predict with input validation
#*
#* @post /predict-validated
#* @serializer json
handle_predict_validated <- function(body, response) {

  result <- tryCatch({

    required_fields <- c("bill_length_mm", "species", "sex") #<1>
    missing_fields <- setdiff(required_fields, names(body)) #<1>

    if (length(missing_fields) > 0) { #<2>
      response$status <- 400L
      return(list(
        error = "Missing required fields",
        missing = missing_fields,
        hint = "Required fields: bill_length_mm, species, sex"
      ))
    } #<2>

    valid_species <- levels(v$prototype$species) #<3>
    if (!body$species %in% valid_species) {
      response$status <- 400L
      return(list(
        error = "Invalid species",
        provided = body$species,
        valid_options = valid_species
      ))
    } #<3>

    valid_sex <- levels(v$prototype$sex) #<4>
    if (!body$sex %in% valid_sex) {
      response$status <- 400L
      return(list(
        error = "Invalid sex",
        provided = body$sex,
        valid_options = valid_sex
      ))
    } #<4>

    if (!is.numeric(body$bill_length_mm) || #<5>
        body$bill_length_mm < 30 ||
        body$bill_length_mm > 60) {
      response$status <- 400L
      return(list(
        error = "bill_length_mm must be numeric between 30 and 60",
        provided = body$bill_length_mm,
        valid_range = c(30, 60)
      ))
    } #<5>

    pred_data <- prep_pred_data(body) #<6>
    prediction <- predict(v, pred_data) #<6>

    list( #<7>
      prediction = as.numeric(prediction),
      input = body,
      model_version = v$metadata$version,
      timestamp = Sys.time()
    ) #<7>

  }, error = function(e) {
    response$status <- 500L
    list(
      error = conditionMessage(e),
      timestamp = as.character(Sys.time())
    )
  })

  return(result)
}
```
1. Check for required fields.
2. Return 400 if fields are missing.
3. Validate species value.
4. Validate sex value.
5. Validate bill length range.
6. Prepare data and predict.
7. Return prediction with metadata.

```{mermaid}
%%| fig-width: 7.0
%%| fig-align: center
%%{init: {'theme': 'base', 'look': 'handDrawn', 'themeVariables': {'fontFamily': 'monospace'}}}%%

graph TD
    Start("Request<br>Received")

    Start --> Check1{All fields<br>present?}
    Check1 -->|No| Error1("400: Missing fields")
    Check1 -->|Yes| Check2{Species<br>valid?}

    Check2 -->|No| Error2("400: Invalid species")
    Check2 -->|Yes| Check3{Sex<br>valid?}

    Check3 -->|No| Error3("400: Invalid sex")
    Check3 -->|Yes| Check4{Bill length<br>30-60mm?}

    Check4 -->|No| Error4("400: Out of range")
    Check4 -->|Yes| Predict("Make Prediction")

    Predict --> Success("Return prediction<br>with metadata")
    classDef default fill:#F2E6C2,stroke:#2A6F77,color:#1B2A41
```

An invalid `species` value returns a 400 error with the valid options:

```{verbatim}
curl -X POST http://127.0.0.1:8080/predict-validated \
   -H "Content-Type: application/json" \
   -d '{"bill_length_mm": 45, "species": "Emperor", "sex": "male"}'
```

```json
{
  "error": ["Invalid species"],
  "provided": ["Emperor"],
  "valid_options": ["Adelie", "Chinstrap", "Gentoo"]
}
```

##### **`/predict-batch`**

Accepts arrays of values to return multiple predictions in a single request:

```{r}
#| eval: false
#| code-fold: true
#| code-summary: 'show/hide handle_predict_batch()'
#* Batch predictions
#*
#* @post /predict-batch
#* @serializer json
handle_predict_batch <- function(body, response) {

  result <- tryCatch({

    if (!is.data.frame(body)) { #<1>
      body <- as.data.frame(body)
    } #<1>

    pred_data <- prep_pred_data(body) #<2>

    predictions <- predict(v, pred_data) #<3>

    list( #<4>
      predictions = as.numeric(predictions),
      count = nrow(pred_data),
      timestamp = Sys.time()
    ) #<4>

  }, error = function(e) {
    response$status <- 500L
    list(
      error = conditionMessage(e),
      timestamp = as.character(Sys.time())
    )
  })

  return(result)
}
```
1. Ensure we have a `data.frame`.
2. Prepare data.
3. Make predictions.
4. Return predictions, count, and timestamp.

```{verbatim}
curl -X POST http://127.0.0.1:8080/predict-batch \
   -H "Content-Type: application/json" \
   -d '{
     "bill_length_mm": [45, 39, 50],
     "species": ["Adelie", "Adelie", "Gentoo"],
     "sex": ["male", "female", "male"]
   }'
```

```json
{
  "predictions": [4180.7965, 3438.2083, 5438.3484],
  "count": [3],
  "timestamp": ["2025-10-08 13:45:41"]
}
```

## The Router

The last section of `plumber.R` assembles a router by chaining handler registrations with `api()`. In `plumber2`, `api()` replaces `pr()` and the `api_get()`/`api_post()` functions replace `pr_get()`/`pr_post()`.

Each `api_get()` or `api_post()` call maps a URL path and HTTP method to a handler function. The `serializer` argument controls how the return value is converted for the response.

### `GET` Endpoints

```{r}
#| eval: false
#| code-fold: show
#| code-summary: 'show/hide GET endpoints'
app <- plumber2::api() |>
  api_get(                              #<1>
    path = "/ping",
    handler = handle_ping,
    serializer = format_json()
  ) |>
  api_get(
    path = "/health",
    handler = handle_health,
    serializer = format_json()
  ) |>
  api_get(
    path = "/model-prototype",
    handler = handle_model_prototype,
    serializer = format_json()
  ) |>
  api_get(
    path = "/model-info",
    handler = handle_model_info,
    serializer = format_json()
  ) |>
  api_get(                              #<2>
    path = "/input-schema",
    handler = handle_input_schema,
    serializer = format_json()
  )
```
1. `api_get()` replaces `pr_get()`; `format_json()` replaces `serializer_json()`.
2. Each `api_get()` call adds one `GET` route to the router's dispatch table.

**Route table after `GET` registrations:**

| Path | Method | Handler | Purpose |
|------|--------|---------|---------|
| `/ping` | GET | `handle_ping()` | Quick health check |
| `/health` | GET | `handle_health()` | Detailed status |
| `/model-prototype` | GET | `handle_model_prototype()` | Input format |
| `/model-info` | GET | `handle_model_info()` | Model metadata |
| `/input-schema` | GET | `handle_input_schema()` | Field documentation |

### `POST` Endpoints

```{r}
#| eval: false
#| code-fold: show
#| code-summary: 'show/hide POST endpoints'
  api_post(
    path = "/predict",
    handler = handle_predict,
    serializer = format_json()
  ) |>
  api_post(
    path = "/predict-validated",
    handler = handle_predict_validated,
    serializer = format_json()
  ) |>
  api_post(
    path = "/predict-batch",
    handler = handle_predict_batch,
    serializer = format_json()
  )
```

**Updated route table:**

| Path | Method | Handler | Purpose |
|------|--------|---------|---------|
| `/predict` | POST | `handle_predict()` | Basic prediction |
| `/predict-validated` | POST | `handle_predict_validated()` | Prediction with validation |
| `/predict-batch` | POST | `handle_predict_batch()` | Multiple predictions |

## The Server

`api_run()` replaces `pr_run()` to start the server:

```{r}
#| eval: false
#| code-fold: false
app |> plumber2::api_run(port = 8080, host = "127.0.0.1")
```

With the API running, we can connect to it from a Shiny application.

## The Shiny App (`app-api.R`)

The `app-api.R` file is a Shiny application that sends user input to the API and displays the prediction. It lives in the separate `app/` directory.

The API URL is defined at the top of the file:

```{r}
#| eval: false
#| code-fold: false
api_url <- "http://127.0.0.1:8080/predict"
```

### UI

```{r}
#| eval: false
#| code-fold: show
#| code-summary: 'show/hide UI'
ui <- page_sidebar(
  title = "Penguin Mass Predictor",
  theme = bs_theme(bootswatch = "sketchy"),
  sidebar = sidebar(
    sliderInput(
      inputId = "bill_length",
      label = "Bill Length (mm)",
      min = 30, max = 60, value = 45, step = 1
    ),
    selectInput(
      inputId = "sex",
      label = "Sex",
      choices = c("male", "female"),
      selected = "male"
    ),
    selectInput(
      inputId = "species",
      label = "Species",
      choices = c("Adelie", "Chinstrap", "Gentoo"),
      selected = "Adelie"
    ),
    actionButton(
      inputId = "predict",
      label = "Predict",
      class = "btn-primary"
    )
  ),
  layout_columns(
    card(
      card_header("Penguin Parameters"),
      card_body(verbatimTextOutput("vals"))
    ),
    card(
      card_header("Predicted Mass"),
      card_body(
        value_box(
          showcase_layout = "left center",
          title = "Grams",
          value = textOutput("pred"),
          showcase = bs_icon("graph-up"),
          max_height = "200px",
          min_height = "200px"
        )
      )
    ),
    col_widths = c(7, 5)
  )
)
```

### Reactive values

Because the `plumber2` API's `/predict` endpoint accepts `species` and `sex` as strings (not dummy variables), `vals()` can pass them directly—no numeric encoding required:

```{r}
#| eval: false
#| code-fold: false
vals <- reactive({
  data.frame(
    bill_length_mm = input$bill_length,
    species = input$species,          #<1>
    sex = input$sex                   #<1>
  )
})
```
1. Species and sex are sent as strings (`"Adelie"`, `"male"`); `prep_pred_data()` in the API converts them to factors.

This is simpler than the original approach, which required manual dummy variable creation (`species_Chinstrap = as.numeric(input$species == "Chinstrap")`, etc.).

### Predictions

The `pred` reactive sends `vals()` to the API and extracts the prediction from the response:

```{r}
#| eval: false
#| code-fold: show
#| code-summary: 'show/hide pred reactive'
pred <- reactive({
  tryCatch({
    showNotification("Predicting penguin mass...", type = "default", duration = 10)

    response <- httr2::request(api_url) |>       #<1>
      httr2::req_method("POST") |>               #<2>
      httr2::req_body_json(vals()) |>            #<3>
      httr2::req_perform() |>                    #<4>
      httr2::resp_body_json()                    #<5>

    showNotification("Prediction successful!", type = "default", duration = 10)

    response$.pred[[1]]

  }, error = function(e) {
    error_msg <- conditionMessage(e)

    if (grepl("Connection refused|couldn't connect", error_msg, ignore.case = TRUE)) {
      user_msg <- "API not available - is the server running on port 8080?"
    } else if (grepl("timeout|timed out", error_msg, ignore.case = TRUE)) {
      user_msg <- "Request timed out - API may be overloaded"
    } else {
      user_msg <- paste("API Error:", substr(error_msg, 1, 50))
    }

    showNotification(paste("Error:", user_msg), type = "warn", duration = 10)
    paste("Error:", user_msg)
  })
}) |>
  bindEvent(input$predict, ignoreInit = TRUE)
```
1. `request()` creates the base request object with the API URL.
2. `req_method("POST")` sets the HTTP verb.
3. `req_body_json()` serializes `vals()` as JSON and sets `Content-Type: application/json`.
4. `req_perform()` sends the request and returns a response object.
5. `resp_body_json()` parses the JSON response body into an R list.

### httr2 request/response pipeline

The five `httr2` functions work as a pipeline, each building on the previous step:

```{mermaid}
%%| fig-width: 7.0
%%| fig-align: center
%%{init: {'theme': 'base', 'look': 'handDrawn', 'themeVariables': {'fontFamily': 'monospace'}}}%%

graph TD
    ReqObj("request(url)") -->|"sets URL"| MethObj
    MethObj("req_method('POST')") -->|"sets verb"| BodyObj
    BodyObj("req_body_json(data)") -->|"attaches body"| PerfObj
    PerfObj("req_perform()") -->|"sends request"| RespObj
    RespObj("resp_body_json()") -->|"parses response"| Result("R list")
    classDef default fill:#F2E6C2,stroke:#2A6F77,color:#1B2A41
```

| Function | Role |
|----------|------|
| `request(url)` | Creates a base request with the target URL |
| `req_method("POST")` | Sets the HTTP verb (`GET`, `POST`, `PUT`, `DELETE`) |
| `req_body_json(data)` | Converts `data` to JSON and sets it as the request body |
| `req_perform()` | Sends the assembled request; returns a response object |
| `resp_body_json()` | Parses the JSON response body into an R list |

### Outputs

The server renders `vals()` and `pred()` to the UI:

```{r}
#| eval: false
#| code-fold: false
output$vals <- renderPrint(vals())
output$pred <- renderText({
  p <- pred()
  if (is.numeric(p)) round(p, 1) else p
})
```

### Full server

```{r}
#| eval: false
#| code-fold: true
#| code-summary: 'show/hide full server'
server <- function(input, output, session) {

  vals <- reactive({
    data.frame(
      bill_length_mm = input$bill_length,
      species = input$species,
      sex = input$sex
    )
  })

  pred <- reactive({
    tryCatch({
      showNotification("Predicting penguin mass...", type = "default", duration = 10)

      response <- httr2::request(api_url) |>
        httr2::req_method("POST") |>
        httr2::req_body_json(vals()) |>
        httr2::req_perform() |>
        httr2::resp_body_json()

      showNotification("Prediction successful!", type = "default", duration = 10)

      response$.pred[[1]]

    }, error = function(e) {
      error_msg <- conditionMessage(e)

      if (grepl("Connection refused|couldn't connect", error_msg, ignore.case = TRUE)) {
        user_msg <- "API not available - is the server running on port 8080?"
      } else if (grepl("timeout|timed out", error_msg, ignore.case = TRUE)) {
        user_msg <- "Request timed out - API may be overloaded"
      } else {
        user_msg <- paste("API Error:", substr(error_msg, 1, 50))
      }

      showNotification(paste("Error:", user_msg), type = "warn", duration = 10)
      paste("Error:", user_msg)
    })
  }) |>
    bindEvent(input$predict, ignoreInit = TRUE)

  output$vals <- renderPrint(vals())
  output$pred <- renderText({
    p <- pred()
    if (is.numeric(p)) round(p, 1) else p
  })
}

shinyApp(ui = ui, server = server)
```

## Verify

The full request/response flow from button click to displayed prediction:

```{mermaid}
%%| fig-width: 7.0
%%| fig-align: center
%%{init: {'theme': 'base', 'look': 'handDrawn', 'themeVariables': {'fontFamily': 'monospace'}}}%%

sequenceDiagram
    participant Shiny as Shiny App
    participant Handler as handle_predict()
    participant Helper as prep_pred_data()
    participant Model as Linear Model

    Shiny->>Handler: POST /predict<br>{"bill_length_mm": 45,<br>"species": "Adelie",<br>"sex": "male"}

    Note over Handler: 1) body parsed<br>by plumber2
    Note over Handler: 2) Convert to<br>data.frame
    Handler->>Helper: prep_pred_data()
    Helper-->>Handler: data.frame with factors

    Note over Handler: 3) Predict
    Handler->>Model: predict(v, data)
    Note over Model: y = β₀ + β₁×bill_length +<br>β₂×species + β₃×sex
    Model-->>Handler: 4180.797

    Note over Handler: 4) Format<br>response
    Handler-->>Shiny: {".pred": [4180.797]}
    style Shiny fill:#F2E6C2,stroke:#2A6F77,color:#1B2A41
    style Handler fill:#F2E6C2,stroke:#2A6F77,color:#1B2A41
    style Helper fill:#F2E6C2,stroke:#2A6F77,color:#1B2A41
    style Model fill:#F2E6C2,stroke:#2A6F77,color:#1B2A41
```

To keep environments self-contained and reproducible, initialize `renv` in each directory:

```{r}
#| eval: false
renv::init()
```
