# <em>GitHub Actions as CI/CD</em> {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

In this lab, we're going to cover using GitHub Actions (or just Actions) for deploying R or Python code (i.e., models, APIs, apps, etc) to a published website. View the original lab file [here](https://github.com/alexkgold/do4ds/tree/main/_labs/lab5).


## Example Actions

To help explain/demonstrate GitHub Actions, I'll cover a few examples I use often. 

### Quarto/GitHub Pages

This book is published using GitHub Pages, which is a free hosting service provided by GitHub. Quarto provides an easy Terminal command for publishing using GitHub Pages:

```bash
quarto publish gh-pages 
```

If I wanted to accomplish the same thing using GitHub Actions, I could create a workflow file in the `.github/workflows` directory of my repository. The workflow file would look something like this:

```yaml
# .github/workflows/publish.yml

name: Publish Quarto Site to GitHub Pages

on:
  push:
    branches: [main]

jobs:
  build-deploy:
    runs-on: ubuntu-latest
    permissions:
      contents: write

    steps:
      - name: Check out repository
        uses: actions/checkout@v4

      - name: Set up Quarto
        uses: quarto-dev/quarto-actions/setup@v2

      - name: Render and Publish to GitHub Pages
        uses: quarto-dev/quarto-actions/publish@v2
        with:
          target: gh-pages
        env:
          GITHUB_TOKEN: ${{ secrets.GITHUB_TOKEN }}
```

#### What it does

- This triggers on every push to `main`, mirroring the behavior of running `quarto publish gh-pages` locally.
- The `contents: write` permission is required so the action can push the rendered output to the `gh-pages` branch.    
- The `GITHUB_TOKEN` is automatically provided by GitHub Actions — no manual secret setup required.

We would also have to make sure our repository's **Pages** settings (Settings → Pages) is configured to deploy from the `gh-pages` branch.

```{=html}
<style>

.codeStyle span:not(.nodeLabel) {
  font-family: monospace;
  font-size: 1.5em;
  font-weight: bold;
  color: #d80808 !important;
  background-color: #ffffff;
  padding: 0.2em;
}
</style>
```

```{mermaid}
%%| fig-cap: 'Publishing to GitHub pages'
%%| fig-width: 7.0
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"14px"}}}%%
flowchart TD
    Push(["Developer<br>pushes to <code>main</code>"]) -->|"triggers"| GHA["GitHub Actions<br>Runner (ubuntu-latest)"]

    GHA --> Step1("<strong>Step 1</strong><br>Checkout Repository<br><code>actions/checkout@v4</code>")
    Step1 --> Step2("<strong>Step 2</strong><br>Set up Quarto<br><code>quarto-actions/setup@v2</code>")
    Step2 --> Step3("<strong>Step 3</strong><br>Render & Publish<br><code>quarto-actions/publish@v2</code>")

    Step3 -->|"authenticates with"| Token["<cdde>GITHUB_TOKEN</code><br>(auto-provided<br>by GitHub)"]
    Token -->|"writes rendered output"| GHPages("<code>gh-pages</code><br>branch")
    GHPages -->|"served as"| Site(["Live GitHub<br>Pages Site"])

    style Push fill:#F9E79F,stroke:#D4AC0D,color:#000
    style GHA fill:#D4E6F1,stroke:#2E86C1,color:#000
    style Step1 fill:#EBF5FB,stroke:#2E86C1,color:#000
    style Step2 fill:#EBF5FB,stroke:#2E86C1,color:#000
    style Step3 fill:#EBF5FB,stroke:#2E86C1,color:#000
    style Token fill:#FDEDEC,stroke:#C0392B,color:#000
    style GHPages fill:#D5F5E3,stroke:#1E8449,color:#000
    style Site fill:#D5F5E3,stroke:#1E8449,color:#000
```


### Running Tests 

You can also use GitHub Actions to run tests on your code. For example, if you have an R package, you can set up a workflow to run `testthat` tests when code is pushed to the `test` branch, and only promote to the `prod` branch if the tests pass. 

```yaml
# stored in .github/workflows/r-package-tests.yml

name: R Package Tests

on:
  push:
    branches: [test]
  pull_request:
    branches: [prod]

jobs:
  test:
    runs-on: ubuntu-latest

    steps:
      - name: Check out repository
        uses: actions/checkout@v4

      - name: Set up R
        uses: r-lib/actions/setup-r@v2
        with:
          use-public-rspm: true

      - name: Install system dependencies
        uses: r-lib/actions/setup-r-dependencies@v2
        with:
          extra-packages: |
            any::rcmdcheck
            any::testthat
            any::devtools

      - name: Run tests
        uses: r-lib/actions/check-r-package@v2
        with:
          error-on: '"error"'
```

#### What it does

- The workflow above triggers on **pushes** to `test` (from `dev`) and on **pull requests** targeting `prod`    
- `error-on: '"error"'` means the workflow only fails on hard errors — warnings won't block promotion   
- To enforce that tests must pass before merging to `prod`, enable branch protection rules in GitHub (Settings → Branches → Add rule for `prod`) and require this workflow as a status check   

```{mermaid}
%%| fig-cap: 'Running tests'
%%| fig-width: 7.0
%%| fig-align: center
%%{init: {'theme': 'neutral', 'look': 'handDrawn', 'themeVariables': { 'fontFamily': 'monospace', "fontSize":"14px"}}}%%
flowchart TD
    DevBr(["<code>dev</code> branch"]) -->|"push to test"| TestBr("<code>test</code> branch")
    TestBr -->|"triggers"| GHA["GitHub Actions<br>Run <code>testthat</code> suite"]
    GHA -->|"✅ tests pass"| PR["Pull Request<br><code>test</code> → <code>prod</code>"]
    GHA -->|"❌ tests fail"| Block["Promotion<br>Blocked"]
    PR -->|"approved & merged"| ProdBr("<code>prod</code> branch")

    style DevBr fill:#D4E6F1,stroke:#2E86C1,color:#000
    style TestBr fill:#D5F5E3,stroke:#1E8449,color:#000
    style ProdBr fill:#FADBD8,stroke:#C0392B,color:#000
    style Block fill:#F9E79F,stroke:#D4AC0D,color:#000
```


