# <em>GitHub Actions as CI/CD</em> {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

In this lab, we're going to cover using GitHub Actions (or just Actions) for deploying R or Python code (i.e., models, APIs, apps, etc) to a published website. View the original lab file [here](https://github.com/alexkgold/do4ds/tree/main/_labs/lab05).

To help explain/demonstrate GitHub Actions, I'll cover a few examples I use often. 

## Publishing Quarto/GitHub Pages

This book is published using GitHub Pages, which is a free hosting service provided by GitHub. Quarto provides an easy Terminal command for publishing using GitHub Pages:

```bash
quarto publish gh-pages 
```

The lab for this section will accomplish the same thing, but does so using a GitHub Action workflow file. However, if you run the commands above (like I did) the workflow files won't be generated because the site isn't published via GitHub Actions. Instead, running `quarto publish gh-pages` creates a _local setup_. This means Quarto renders the files locally, then pushes the rendered HTML directly to the `gh-pages` branch. The commit log for this branch confirms it:

```bash
git log gh-pages --oneline -5
# 76175e88 Built site for gh-pages
# 559df417 Built site for gh-pages
# ...
```

The `Built site for gh-pages` messages are Quarto's signature commit message when it pushes to the branch itself. **No workflow file is involved.**

To convert this to CI publishing, we need to create and commit two files so GitHub Actions can render and deploy on every push to `main`:

1. `.github/workflows/publish.yml`: the Actions workflow that triggers the build
2. `_publish.yml`: the Quarto publish configuration that tells Quarto where to deploy

### Workflow file

Create `.github/workflows/publish.yml` with the following:

```yaml
# .github/workflows/publish.yml

name: Publish to GitHub Pages

on:
  push:
    branches: [main]
  workflow_dispatch:

jobs:
  publish:
    runs-on: ubuntu-latest
    permissions:
      contents: write

    steps:
      - name: Check out repository
        uses: actions/checkout@v4

      - name: Set up Quarto
        uses: quarto-dev/quarto-actions/setup@v2

      - name: Render and publish
        uses: quarto-dev/quarto-actions/publish@v2
        with:
          target: gh-pages
        env:
          GITHUB_TOKEN: ${{ secrets.GITHUB_TOKEN }}
```

The `workflow_dispatch` trigger lets you kick off a manual build from the GitHub Actions UI without needing to push a commit.

### `_publish.yml`

Create `_publish.yml` at the project root and include the following:

```yaml
- source: project
  gh-pages:
    branch: gh-pages
```

This tells Quarto to publish the rendered output to the `gh-pages` branch. Quarto reads this file during the publish step so it knows where to send the output without requiring any flags on the command line.

### What they do

The two files divide the work cleanly between Quarto and GitHub Actions:

* `_publish.yml` records the publish target so the project is reproducibly configured regardless of who (or what) runs the `publish` command. 

* `.github/workflows/publish.yml` listens for a push to `main`, spins up a runner, installs Quarto, and then calls `quarto publish` using the settings in `_publish.yml`.

The `GITHUB_TOKEN` is provided automatically by GitHub for every workflow run, so no manual secret setup is needed. Once both files are committed and pushed, every subsequent push to `main` will trigger a fresh render and deploy.

```{=html}
<style>

.codeStyle span:not(.nodeLabel) {
  font-family: monospace;
  font-size: 1.5em;
  font-weight: bold;
  color: #d80808 !important;
  background-color: #ffffff;
  padding: 0.2em;
}
</style>
```

```{mermaid}
%%| fig-cap: 'Publishing to GitHub pages'
%%| fig-width: 7.0
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%
flowchart TD
    Push("Developer<br>pushes to <code>main</code>") -->|"triggers"| GHA("GitHub Actions<br>Runner<br>(ubuntu-latest)")

    GHA --> Step1("<strong>Step 1</strong><br>Checkout Repository<br><code>actions/checkout@v4</code>")
    Step1 --> Step2("<strong>Step 2</strong><br>Set up Quarto<br><code>quarto-actions/setup@v2</code>")
    Step2 --> Step3("<strong>Step 3</strong><br>Render & Publish<br><code>quarto-actions/publish@v2</code>")

    Step3 -->|"authenticates with"| Token("<cdde>GITHUB_TOKEN</code><br>(auto-provided<br>by GitHub)")
    Token -->|"writes rendered output"| GHPages("<code>gh-pages</code><br>branch")
    GHPages -->|"served as"| Site("Live GitHub<br>Pages Site")
    style Push fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style GHA fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style Step1 fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style Step2 fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style Step3 fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style Token fill:#1B2A41,stroke:#000000,stroke-width:1px,color:#ffffff
    style GHPages fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    style Site fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff

```


## Running Tests 

You can also use GitHub Actions to run tests on your code. For example, if we had Shiny for Python package, we could set up a workflow to run unit tests when the code is pushed to the `test` branch, and only promote to the `prod` branch if the tests pass. We'll use the application from [*Python App Logging*](app-log-py.qmd).[^py-app-log]

[^py-app-log]: The original app is in [`_labs/lab04/Python/app/`](https://github.com/mjfrigaard/do4ds-labs/tree/main/_labs/lab04/Python/app). The refactored package used here is in [`_labs/lab05/cicd/Python/penguin_predictor/`](https://github.com/mjfrigaard/do4ds-labs/tree/main/_labs/lab05/cicd/Python/penguin_predictor).

### Workflow file

```yaml
# stored in _labs/lab05/cicd/Python/penguin_predictor/.github/workflows/python-tests.yml

name: Python Tests

on:
  push:
    branches: [test]
  pull_request:
    branches: [prod]

jobs:
  test:
    runs-on: ubuntu-latest

    steps:
      - name: Check out repository
        uses: actions/checkout@v4

      - name: Set up Python
        uses: actions/setup-python@v5
        with:
          python-version: "3.12"

      - name: Install package and test dependencies
        run: pip install -e ".[test]"

      - name: Run tests
        run: pytest tests/ -v
```

### What it does

- The workflow above triggers on **pushes** to `test` (from `dev`) and on **pull requests** targeting `prod`
- `pip install -e ".[test]"` installs the `penguin_predictor` package and its test dependencies (`pytest`) from `pyproject.toml`
- To enforce that tests must pass before merging to `prod`, enable branch protection rules in GitHub (Settings > Branches > Add rule for `prod`) and require this workflow as a status check

```{mermaid}
%%| fig-cap: 'Running tests'
%%| fig-width: 7.0
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%
flowchart TD
    DevBr("<code>dev</code> branch") -->|"push to test"| TestBr("<code>test</code> branch")
    TestBr -->|"triggers"| GHA("GitHub Actions<br>Run <code>pytest</code> suite")
    GHA -->|"tests pass"| PR("Pull Request<br><code>test</code> to <code>prod</code>")
    GHA -->|"tests fail"| Block("Promotion<br>Blocked")
    PR -->|"approved & merged"| ProdBr("<code>prod</code> branch")
    style DevBr fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style TestBr fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style GHA fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style PR fill:#1B2A41,stroke:#000000,stroke-width:1px,color:#ffffff
    style Block fill:#1B2A41,stroke:#000000,stroke-width:1px,color:#ffffff
    style ProdBr fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff

```

## A note on workflow file placement

The workflow file in [`.github/workflows/python-tests.yml`](https://github.com/mjfrigaard/do4ds-labs/tree/main/_labs/lab05/cicd/Python/penguin_predictor) is included as a reference — it will not actually execute when this book is pushed to GitHub. GitHub Actions only picks up workflow files located in the **repository root's** `.github/workflows/` directory, so anything nested deeper is treated as an ordinary text file.

In your own project, the same workflow file would go at the root of the repository:

```{bash}
#| eval: false 
#| code-fold: false 
penguin_predictor/
├── app.py
├── .github/
│   └── workflows/
│       └── python-tests.yml #<1>
├── .gitignore
├── penguin_predictor/
│   ├── data.py
│   └── logging_setup.py
├── pyproject.toml
├── requirements.txt
└── tests/
    └── test_data.py

```
1. GitHub Actions reads this  

With that structure in place, every push to `test` and every pull request targeting `prod` will trigger the workflow automatically.
