# <em>Docker and Publishing</em> {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

## Create a separate IAM user for GitHub Actions

Do not reuse your personal credentials. Create a dedicated user with only S3 access.

IAM \> User groups \> Create group: `github-actions-s3`

Attach a scoped custom policy instead of full access. Example policy JSON, limited to one bucket:

``` json
{
  "Version": "2012-10-17",
  "Statement": [
    {
      "Effect": "Allow",
      "Action": [
        "s3:GetObject",
        "s3:PutObject",
        "s3:ListBucket"
      ],
      "Resource": [
        "arn:aws:s3:::your-vetiver-bucket-name",
        "arn:aws:s3:::your-vetiver-bucket-name/*"
      ]
    }
  ]
}
```

IAM \> Policies \> Create policy \> JSON tab \> paste above \> name it `github-actions-s3-policy`.

Attach this policy to the `github-actions-s3` group.

Create user: `github-actions-bot` \> add to `github-actions-s3` group \> no console access needed.

Create access key for this user (CLI use case), same as before.

## Add credentials to GitHub

Repo \> Settings \> Secrets and variables \> Actions \> New repository secret.

Add:

- `AWS_ACCESS_KEY_ID`
- `AWS_SECRET_ACCESS_KEY`
- `AWS_REGION`

## Example GitHub Actions step using these secrets

``` yaml
- name: Configure AWS credentials
  uses: aws-actions/configure-aws-credentials@v4
  with:
    aws-access-key-id: ${{ secrets.AWS_ACCESS_KEY_ID }}
    aws-secret-access-key: ${{ secrets.AWS_SECRET_ACCESS_KEY }}
    aws-region: ${{ secrets.AWS_REGION }}

- name: Push to S3
  run: aws s3 cp ./model.rds s3://your-vetiver-bucket-name/model.rds
```

