# <em>GitHub Actions as CI/CD</em> {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

In this lab we're going to cover using GitHub Actions (or just Actions) for deploying your R or Python code (i.e., models, APIs, apps, etc). 

To help explain/demonstrate GitHub Actions, I'll cover a few examples I use often. 

## Quarto/GitHub Pages

This book is published using GitHub Pages, which is a free hosting service provided by GitHub. Quarto provides an easy Terminal command for publishing using GitHub Pages:

```bash
quarto publish gh-pages 
```

If I wanted to accomplish the same thing using GitHub Actions, I could create a workflow file in the `.github/workflows` directory of my repository. The workflow file would look something like this:

```yaml
# .github/workflows/publish.yml

name: Publish Quarto Site to GitHub Pages

on:
  push:
    branches: [main]

jobs:
  build-deploy:
    runs-on: ubuntu-latest
    permissions:
      contents: write

    steps:
      - name: Check out repository
        uses: actions/checkout@v4

      - name: Set up Quarto
        uses: quarto-dev/quarto-actions/setup@v2

      - name: Render and Publish to GitHub Pages
        uses: quarto-dev/quarto-actions/publish@v2
        with:
          target: gh-pages
        env:
          GITHUB_TOKEN: ${{ secrets.GITHUB_TOKEN }}
```

### What it does:

- This triggers on every push to `main`, mirroring the behavior of running `quarto publish gh-pages` locally.
- The `contents: write` permission is required so the action can push the rendered output to the `gh-pages` branch.    
- The `GITHUB_TOKEN` is automatically provided by GitHub Actions — no manual secret setup required.

We would also have to make sure our repository's **Pages** settings (Settings → Pages) is configured to deploy from the `gh-pages` branch.

## Running Tests 

You can also use GitHub Actions to run tests on your code. For example, if you have an R package, you can set up a workflow to run 

