# <em>GitHub Actions as CI/CD</em> {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

In this lab, we're going to cover using GitHub Actions (or just Actions) for deploying R or Python code (i.e., models, APIs, apps, etc) to a published website. View the original lab file [here](https://github.com/alexkgold/do4ds/tree/main/_labs/lab05).

To help explain/demonstrate GitHub Actions, I'll cover a few examples I use often. 

## Publishing Quarto/GitHub Pages

This book is published using GitHub Pages, which is a free hosting service provided by GitHub. Quarto provides an easy Terminal command for publishing using GitHub Pages:

```bash
quarto publish gh-pages 
```

The lab for this section will accomplish the same thing, but does so using a GitHub Action workflow file. However, if you run the commands above -- as I did -- the workflow files don't exist because the site isn't published via GitHub Actions. Instead, this repo uses local publishing.

There are two valid ways to publish a Quarto book to GitHub Pages, and running `quarto publish gh-pages` creates a _local setup_. This means Quarto renders the files locally, then pushes the rendered HTML directly to the `gh-pages` branch. The commit log for this branch confirms it:

```bash
git log gh-pages --oneline -5
# 76175e88 Built site for gh-pages
# 559df417 Built site for gh-pages
# ...
```

The `Built site for gh-pages` messages are Quarto's signature commit message when it pushes to the branch itself. **No workflow file is involved.**

To convert this to CI publishing, we'll have to create and commit `.github/workflows/publish.yml` and  `_publish.yml` files, so GitHub Actions can render and deploy on every push to `main`.

### Workflow file

In the `.github/workflows/publish.yml` directory, create ...


### `_publish.yml`

Create `_publish.yml` and include the following ...

  
### What they do



```{=html}
<style>

.codeStyle span:not(.nodeLabel) {
  font-family: monospace;
  font-size: 1.5em;
  font-weight: bold;
  color: #d80808 !important;
  background-color: #ffffff;
  padding: 0.2em;
}
</style>
```

```{mermaid}
%%| fig-cap: 'Publishing to GitHub pages'
%%| fig-width: 7.0
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%
flowchart TD
    Push("Developer<br>pushes to <code>main</code>") -->|"triggers"| GHA("GitHub Actions<br>Runner (ubuntu-latest)")

    GHA --> Step1("<strong>Step 1</strong><br>Checkout Repository<br><code>actions/checkout@v4</code>")
    Step1 --> Step2("<strong>Step 2</strong><br>Set up Quarto<br><code>quarto-actions/setup@v2</code>")
    Step2 --> Step3("<strong>Step 3</strong><br>Render & Publish<br><code>quarto-actions/publish@v2</code>")

    Step3 -->|"authenticates with"| Token("<cdde>GITHUB_TOKEN</code><br>(auto-provided<br>by GitHub)")
    Token -->|"writes rendered output"| GHPages("<code>gh-pages</code><br>branch")
    GHPages -->|"served as"| Site("Live GitHub<br>Pages Site")
    style Push fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style GHA fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style Step1 fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style Step2 fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style Step3 fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style Token fill:#1B2A41,stroke:#000000,stroke-width:1px,color:#ffffff
    style GHPages fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff
    style Site fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff

```


## Running Tests 

You can also use GitHub Actions to run tests on your code. For example, if you have an R package, you can set up a workflow to run `testthat` tests when code is pushed to the `test` branch, and only promote to the `prod` branch if the tests pass. 

### Workflow file

```yaml
# stored in .github/workflows/r-package-tests.yml

name: R Package Tests

on:
  push:
    branches: [test]
  pull_request:
    branches: [prod]

jobs:
  test:
    runs-on: ubuntu-latest

    steps:
      - name: Check out repository
        uses: actions/checkout@v4

      - name: Set up R
        uses: r-lib/actions/setup-r@v2
        with:
          use-public-rspm: true

      - name: Install system dependencies
        uses: r-lib/actions/setup-r-dependencies@v2
        with:
          extra-packages: |
            any::rcmdcheck
            any::testthat
            any::devtools

      - name: Run tests
        uses: r-lib/actions/check-r-package@v2
        with:
          error-on: '"error"'
```

### What it does

- The workflow above triggers on **pushes** to `test` (from `dev`) and on **pull requests** targeting `prod`    
- `error-on: '"error"'` means the workflow only fails on hard errors — warnings won't block promotion   
- To enforce that tests must pass before merging to `prod`, enable branch protection rules in GitHub (Settings → Branches → Add rule for `prod`) and require this workflow as a status check   

```{mermaid}
%%| fig-cap: 'Running tests'
%%| fig-width: 7.0
%%| fig-align: 'center'
%%{init: {'theme': 'base', 'themeVariables': {'fontFamily': 'monospace'}}}%%
flowchart TD
    DevBr("<code>dev</code> branch") -->|"push to test"| TestBr("<code>test</code> branch")
    TestBr -->|"triggers"| GHA("GitHub Actions<br>Run <code>testthat</code> suite")
    GHA -->|"✅ tests pass"| PR("Pull Request<br><code>test</code> → <code>prod</code>")
    GHA -->|"❌ tests fail"| Block("Promotion<br>Blocked")
    PR -->|"approved & merged"| ProdBr("<code>prod</code> branch")
    style DevBr fill:#5B8C5A,stroke:#000000,stroke-width:1px,color:#ffffff
    style TestBr fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style GHA fill:#D2562B,stroke:#000000,stroke-width:1px,color:#ffffff
    style PR fill:#1B2A41,stroke:#000000,stroke-width:1px,color:#ffffff
    style Block fill:#1B2A41,stroke:#000000,stroke-width:1px,color:#ffffff
    style ProdBr fill:#2A6F77,stroke:#000000,stroke-width:1px,color:#ffffff

```


