# <em>GitHub Actions as CI/CD</em> {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

In this lab we're going to cover using GitHub Actions (or just Actions) for deploying your R or Python code (i.e., models, APIs, apps, etc). 

To help explain/demonstrate GitHub Actions, I'll cover how this book is published.

```bash
quarto publish gh-pages 
```

If I wanted to publish this book using GitHub Actions, I could perform the following:

```bash

```


The original [`_labs/lab5/publish-r-py.yml` file](https://github.com/alexkgold/do4ds/tree/main/_labs/lab5) is below for reference: 

```yaml
on:
  workflow_dispatch:
  push:
    branches: main

name: Quarto Publish

jobs:
  build-deploy:
    runs-on: ubuntu-latest
    permissions:
      contents: write
    steps:
      - name: Check out repository
        uses: actions/checkout@v2

      - name: Set up Quarto
        uses: quarto-dev/quarto-actions/setup@v2

      - name: Install R
        uses: r-lib/actions/setup-r@v2
        with:
          r-version: '4.2.0'
          use-public-rspm: true

      - name: Setup renv and install packages
        uses: r-lib/actions/setup-renv@v2
        with:
          cache-version: 1
        env:
          RENV_CONFIG_REPOS_OVERRIDE: https://packagemanager.rstudio.com/all/latest

      - name: Install Python and Dependencies
        uses: actions/setup-python@v4
        with:
          python-version: '3.10'
          cache: 'pip'
      - run: pip install jupyter
      - run: pip install -r requirements.txt

      - name: Render and Publish
        uses: quarto-dev/quarto-actions/publish@v2
        with:
          target: gh-pages
        env:
          GITHUB_TOKEN: ${{ secrets.GITHUB_TOKEN }}
```

