# <em>Pull API model from S3</em> {.unnumbered}

```{r}
#| label: common
#| include: false
source("_common.R")
```

```{r}
#| label: co_box_rev
#| echo: false
#| results: asis
#| eval: true
co_box(
  color = "o",
  look = "default", 
  hsize = "1.25", 
  size = "1.00", 
  header = "Caution", 
  fold = FALSE,
  contents = "This section is being revised. Thank you for your patience."
)
```

## Step 3: Pull the API model from S3

You’ll also have to configure the API to load the model from the S3 bucket. If you’re using {`vetiver`} that’s as easy as changing the board creation line to:

```python
from pins import board_s3
board = board_s3("do4ds-lab", allow_pickle_read=True)
```

## Step 4: Give GitHub Actions the S3 credentials

We want our model building to correctly push to S3 even when it’s running in GitHub Actions, but since GitHub doesn’t have our S3 credentials by default, so we’ll need to provide them.

We will declare the variables we need in the Render and Publish step of the Action.

Once you’re done, that section of the publish.yml should look something like this:

```yaml
.github/workflows/publish.yml
  env:
    GITHUB_TOKEN: ${{ secrets.GITHUB_TOKEN }}
    AWS_ACCESS_KEY_ID: ${{ secrets.AWS_ACCESS_KEY_ID }}
    AWS_SECRET_ACCESS_KEY: ${{ secrets.AWS_SECRET_ACCESS_KEY }}
    AWS_REGION: us-east-1
```

Now, unlike the `GITHUB_TOKEN` secret, which GitHub Actions automatically provides to itself, we’ll have to give these secrets to the GitHub interface.