---
title: "R Vetiver Model"
format: 
  gfm:
    preview-mode: raw
    code-fold: false 
    embed-resources: true
knitr:
  opts_chunk: 
    collapse: false
    comment: "#>"
---

This document demonstrates how to create a simple linear regression model using the `palmerpenguins` dataset, and then turn it into a `Vetiver` model for deployment using R. 

## Load Libraries

```{r}
#| message: false
#| warning: false
library(palmerpenguins)
library(duckdb)
library(DBI)
library(dplyr)
library(vetiver)
library(pins)
```

## Connect to DuckDB

```{r}
con <- DBI::dbConnect(duckdb::duckdb(), "R/my-db.duckdb")

duckdb::duckdb_register(con, "penguins_raw", palmerpenguins::penguins) #<1>


DBI::dbExecute( #<2>
  con,
  "CREATE OR REPLACE TABLE penguins AS SELECT * FROM penguins_raw"
)#<2>
```
1. Register the `penguins` `data.frame` directly with `duckdb`
2. Create persistent table in `duckdb`

## Get Data

```{r}
df <- DBI::dbGetQuery(
  con,
  "SELECT bill_length_mm, species, sex, body_mass_g 
   FROM penguins 
   WHERE body_mass_g IS NOT NULL 
   AND bill_length_mm BETWEEN 30 AND 60
   AND sex IS NOT NULL
   AND species IS NOT NULL"
)

DBI::dbDisconnect(con)
```

Below we can see how many rows were loaded from the SQL query, and a preview of the first three rows of the data frame.
```{r}
nrow(df)
head(df, 3)
```

## Define Model and Fit

```{r}
model <- lm(body_mass_g ~ bill_length_mm + species + sex, data = df)
```

```{r}
print(formula(model))
print(coef(model))
```

## Turn into Vetiver Model

```{r}
v <- vetiver::vetiver_model(
  model,
  model_name = "penguin_model",
  description = "Linear model predicting penguin body mass from bill length, species, and sex",
  save_prototype = TRUE  
)
```

Check the prototype data that was saved with the model. This is the data that will be used to validate new data when making predictions with the model.

```{r}
print(v$prototype)
```

## Write to Local Board

Write the model to a local board so that it can be used for deployment. The model will be saved in the `R/models` directory.

```{r}
#| collapse: false
model_board <- pins::board_folder("R/models")
vetiver::vetiver_pin_write(model_board, v)
```

## Write to S3 Bucket 

We'll start by storing the necessary environment variables:

1. Create a `.Renviron` file in your project root:

```bash
touch .Renviron
```

2. Place the environment variables in the `.Renviron` file.

```bash
AWS_ACCESS_KEY_ID=your_access_key
AWS_SECRET_ACCESS_KEY=your_secret_key
AWS_REGION=us-east-1
```

3. Restart R session to load `.Renviron`.

4. We can access variables using `Sys.getenv()` to confirm they are loaded correctly.

```{r}
#| eval: false
Sys.getenv("AWS_ACCESS_KEY_ID")
Sys.getenv("AWS_SECRET_ACCESS_KEY")
Sys.getenv("AWS_REGION")
```

5. Add `.Renviron` to `.gitignore`

Now we're ready to write the model data to the S3 bucket. You might need the `paws.storage` package installed to write to the S3 board. Install it with the following command:

```{r}
# install.packages("paws.storage")
```

`pins::board_s3()` will create a connection to the S3 bucket using the environment variables we set earlier.

```{r}
#| collapse: false
s3_board <- pins::board_s3(
  bucket = "penguin-vetiver-model-data",
  prefix = "R/models/",
  region = Sys.getenv("AWS_REGION"),
  access_key = Sys.getenv("AWS_ACCESS_KEY_ID"),
  secret_access_key = Sys.getenv("AWS_SECRET_ACCESS_KEY")
)

vetiver::vetiver_pin_write(s3_board, v)
```

We can confirm this with the following commands: 

```{r}
pins::pin_list(s3_board)
```
